---
title: "BASiCS - 40,000 iterations"
date: 2015-08-10
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

We analyzed our single cell data with [BASiCS][] developed by [Vallejos et al., 2015][vallejos2015].
The results shown here are from a model fit with 40,000 iterations.

[basics]: https://github.com/catavallejos/BASiCS
[vallejos2015]: http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004333

BASiCS and its dependency, RcppArmadillo, were able to be installed on the cluster using a new version of gcc.
Since this took a long time to run, it was submitted via the following:

```bash
echo "Rscript -e 'library(rmarkdown); render(\"basics.Rmd\")'" | \
  qsub -l h_vmem=32g -cwd -V -j y -o ~/log/ -N basics
```

```{r packages}
library("BASiCS")
library("data.table")
source("functions.R")
library("ggplot2")
theme_set(theme_bw(base_size = 16))
library("tidyr")
library("edgeR")
```

##  Input

Below is the description of the data from the BASiCS vignette, interspersed with my code to load the data.

> The input dataset for BASiCS must contain the following 3 elements:

> * `Counts`: a matrix of raw expression counts with dimensions $q$ times $n$. First $q_0$ rows must correspond to biological genes. Last $q-q_0$ rows must correspond to technical spike-in genes.

Input annotation.

```{r input-annotation}
anno_filter <- read.table("../data/annotation-filter.txt", header = TRUE,
                          stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules_filter <- read.table("../data/molecules-filter.txt", header = TRUE,
                               stringsAsFactors = FALSE)
stopifnot(nrow(anno_filter) == ncol(molecules_filter),
          colnames(molecules_filter) == anno_filter$sample_id,
          anno_filter$well != "bulk")
```

Remove outlier batch NA19098.r2.

```{r remove-NA19098.r2}
molecules_filter <- molecules_filter[, anno_filter$batch != "NA19098.r2"]
anno_filter <- anno_filter[anno_filter$batch != "NA19098.r2", ]
stopifnot(nrow(anno_filter) == ncol(molecules_filter),
          colnames(molecules_filter) == anno_filter$sample_id,
          anno_filter$well != "bulk")
```

> * `Tech`: a vector of `TRUE`/`FALSE` elements with length $q$. If `Tech[i] = FALSE` the gene `i` is biological; otherwise the gene is spike-in.

```{r tech}
tech <- grepl("ERCC", rownames(molecules_filter))
```

> * `SpikeInput`: a vector of length $q-q_0$ whose elements contain the input number of molecules for the spike-in genes (amount per cell).

```{r load-spike-input}
spike <- read.table("../data/expected-ercc-molecules.txt", header = TRUE,
                    sep = "\t", stringsAsFactors = FALSE)
```

Only keep the spike-ins that were kept after applying the expression cutoff.

```{r spike-input}
spike_input <- spike$ercc_molecules_well[spike$id %in% rownames(molecules_filter)]
names(spike_input) <- spike$id[spike$id %in% rownames(molecules_filter)]
spike_input <- spike_input[order(names(spike_input))]
stopifnot(sum(tech) == length(spike_input),
          rownames(molecules_filter)[tech] == names(spike_input))
```

`r length(spike_input)` of the ERCC spike-ins were observed in the single cell data.

> These elements must be stored into an object of class `BASiCS_Data`.

```{r create-object}
basics_data <- newBASiCS_Data(as.matrix(molecules_filter), tech, spike_input)
```

## Fit the model

```{r fit-model}
store_dir <- "../data"
run_name <- "new"
if (file.exists(paste0(store_dir, "/chain_phi_", run_name, ".txt"))) {
  chain_mu = as.matrix(fread(paste0(store_dir, "/chain_mu_", run_name, ".txt")))
  chain_delta = as.matrix(fread(paste0(store_dir, "/chain_delta_", run_name, ".txt")))
  chain_phi = as.matrix(fread(paste0(store_dir, "/chain_phi_", run_name, ".txt")))
  chain_s = as.matrix(fread(paste0(store_dir, "/chain_s_", run_name, ".txt")))
  chain_nu = as.matrix(fread(paste0(store_dir, "/chain_nu_", run_name, ".txt")))
  chain_theta = as.matrix(fread(paste0(store_dir, "/chain_theta_", run_name, ".txt")))

  mcmc_output <- newBASiCS_Chain(mu = chain_mu, delta = chain_delta,
                                 phi = chain_phi, s = chain_s,
                                 nu = chain_nu, theta = chain_theta)

  time_total <- readRDS(paste0(store_dir, "/time_total_", run_name, ".rds"))
} else {
  time_start <- Sys.time()
  mcmc_output <- BASiCS_MCMC(basics_data, N = 40000, Thin = 10, Burn = 20000,
                             PrintProgress = TRUE, StoreChains = TRUE,
                             StoreDir = store_dir, RunName = run_name)
  time_end <- Sys.time()
  time_total <- difftime(time_end, time_start, units = "hours")
  saveRDS(time_total, paste0(store_dir, "/time_total_", run_name, ".rds"))
}
```

Fitting the model took `r round(as.numeric(time_total), 2)` hours.

Summarize the results.

```{r summary}
mcmc_summary <- Summary(mcmc_output)
```

## Cell-specific normalizing constants

BASiCS models two cell-specific parameters.
Phi models the differences in gene molecules.
S models the differences in ERCC molecules.

```{r phi-rna-content}
plot(mcmc_summary, Param = "phi")
```

```{r s-capture-efficiency}
plot(mcmc_summary, Param = "s")
```

```{r extract-model-parameters}
phi <- displaySummaryBASiCS(mcmc_summary, Param = "phi")
s <- displaySummaryBASiCS(mcmc_summary, Param = "s")
parameters <- cbind(phi, s, anno_filter)
parameters$gene_count <- colSums(counts(basics_data, type = "biological"))
parameters$ercc_count <- colSums(counts(basics_data, type = "technical"))
```

### Phi versus gene molecule count

```{r phi-versus-gene-molecule-count}
phi_gene_count <- ggplot(parameters, aes(x = gene_count, y = Phi)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Gene molecule count",
       title = paste0("Phi measures gene molecule count differences\nr = ",
                      round(cor(parameters$gene_count, parameters$Phi), 2)))
phi_gene_count
```

### Phi versus ERCC molecule count

```{r phi-versus-ercc-molecule-count}
phi_ercc_count <- phi_gene_count %+% aes(x = ercc_count) +
  labs(x = "ERCC molecule count",
       title = paste0("Phi does not measure ERCC molecule count differences\nr = ",
                      round(cor(parameters$ercc_count, parameters$Phi), 2)))
phi_ercc_count
```

### S versus ERCC molecule count

```{r s-versus-ercc-molecule-count}
s_ercc_count <- phi_ercc_count %+% aes(y = S) +
  labs(y = "S",
       title = paste0("S measures ERCC molecule count differences\nr = ",
                      round(cor(parameters$ercc_count, parameters$S), 2)))
s_ercc_count
```

### S versus gene molecule count

```{r s-versus-gene-molecule-count}
s_gene_count <- phi_gene_count %+% aes(y = S) +
  labs(y = "S",
       title = paste0("S does not measure gene molecule count differences\nr = ",
                      round(cor(parameters$gene_count, parameters$S), 2)))
s_gene_count
```

### Phi versus S

```{r phi-versus-s}
phi_s <- ggplot(parameters, aes(x = S, y = Phi)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(       title = paste0("Phi versus S\nr = ",
                      round(cor(parameters$S, parameters$Phi), 2)))
phi_s
```

## Denoised data

Remove technical noise (i.e. normalize using the ERCC spike-ins).

```{r denoised-counts}
denoised = BASiCS_DenoisedCounts(Data = basics_data, Chain = mcmc_output)
```

### PCA - BASiCS Denoised

Both the raw and the cpm versions of the BASiCS denoised data appear similar to the result with the [non-normalized cpm data](#pca-non-normalized-cpm).
This does not change substantially when increasing the iterations from a few thousands to a few tens of thousands.

```{r pca-basics}
pca_basics <- run_pca(denoised)
plot_pca(pca_basics$PCs, explained = pca_basics$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate")
```

### PCA - BASiCS Denoised cpm

```{r pca-basics-cpm}
denoised_cpm <- cpm(denoised, log = TRUE,
                    lib.size = colSums(denoised) *
                               calcNormFactors(denoised, method = "TMM"))
pca_basics_cpm <- run_pca(denoised_cpm)
plot_pca(pca_basics_cpm$PCs, explained = pca_basics_cpm$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate")
```

### PCA - non-normalized

```{r pca-non-normalized}
pca_non <- run_pca(counts(basics_data))
plot_pca(pca_non$PCs, explained = pca_non$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate")
```

### PCA - non-normalized cpm

```{r pca-non-normalized-cpm}
non_cpm <- cpm(counts(basics_data), log = TRUE,
               lib.size = colSums(counts(basics_data)) *
                          calcNormFactors(counts(basics_data), method = "TMM"))
pca_non_cpm <- run_pca(non_cpm)
plot_pca(pca_non_cpm$PCs, explained = pca_non_cpm$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate")
```

## Session information

```{r info}
sessionInfo()
```
