---
title: "Mixed effect model for batch correction - limma"
author: "Joyce Hsiao"
date: 2015-10-30
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objective 

Run Mixed model for batch-effect correction on up-to-date filtered data. See [link1] for details.

[link1](https://jdblischak.github.io/singleCellSeq/analysis/data-transformation.html)


## Setup

```{r}
source("functions.R")
library("limma")
library("edgeR")
library(ggplot2)
theme_set(theme_bw(base_size = 16))
```


## Data preparation

Input quality single cells

```{r}
quality_single_cells <- read.table("../data/quality-single-cells.txt", 
                                  header = TRUE,
                                  stringsAsFactors = FALSE)
str(quality_single_cells)
```


Input annotation

```{r}
anno <- read.table("../data/annotation.txt", 
                    header = TRUE,
                    stringsAsFactors = FALSE)
anno_filter <- anno[ which(anno$sample_id %in% quality_single_cells[[1]]), ]
dim(anno_filter)
```


*Molecule counts

Input molecule counts.

```{r}
molecules <- read.table("../data/molecules.txt", 
                          header = TRUE,
                          stringsAsFactors = FALSE)
dim(molecules)
```

Filter cells.

```{r}
molecules_filter <- molecules[ , which(colnames(molecules) %in% quality_single_cells[[1]])]
dim(molecules_filter)
```

Filter genes.

```{r}
molecules_cpm_mean <- rowMeans(cpm(molecules_filter, log = TRUE))
lower_exp_cutoff <- 2
genes_pass_filter <- rownames(molecules_filter)[molecules_cpm_mean > lower_exp_cutoff]
```


```{r}
overexpressed_rows <- apply(molecules_filter, 1, function(x) any(x >= 1024))
overexpressed_genes <- rownames(molecules_filter)[overexpressed_rows]
genes_pass_filter <- setdiff(genes_pass_filter, overexpressed_genes)
```


```{r}
molecules_filter <- molecules_filter[rownames(molecules_filter) %in% genes_pass_filter, ]
```

*Correct for collision probability

```{r}
molecules_collision <- -1024 * log(1 - molecules_filter / 1024)
```

*Molecules single cell endogeneous

```{r}
ercc_rows <- grepl("ERCC", rownames(molecules_filter))
molecules_cpm <- cpm(molecules_collision[!ercc_rows, ], log = TRUE)
```

*Linear transformation

```{r}
molecules_cpm_ercc <- cpm(molecules_collision[ercc_rows, ], log = TRUE)
ercc <- read.table("../data/expected-ercc-molecules.txt", header = TRUE,
                   stringsAsFactors = FALSE)
ercc <- ercc[ercc$id %in% rownames(molecules_cpm_ercc), ]
ercc <- ercc[order(ercc$id), ]
ercc$log2_cpm <- cpm(ercc$ercc_molecules_well, log = TRUE)
```

```{r}
molecules_cpm_trans <- molecules_cpm
molecules_cpm_trans[, ] <- NA
intercept <- numeric(length = ncol(molecules_cpm_trans))
slope <- numeric(length = ncol(molecules_cpm_trans))
for (i in 1:ncol(molecules_cpm_trans)) {
  fit <- lm(molecules_cpm_ercc[, i] ~ ercc$log2_cpm)
  intercept[i] <- fit$coefficients[1]
  slope[i] <- fit$coefficients[2]
  # Y = mX + b -> X = (Y - b) / m
  molecules_cpm_trans[, i] <- (molecules_cpm[, i] - intercept[i]) / slope[i]
}
dim(molecules_cpm_trans)
```


## Remove unwanted variation 

Load the Humanzee package

```{r}
if (!require(Humanzee, quietly = TRUE)) {
  library(devtools)
  install_github("jhsiao999/Humanzee")
  library(Humanzee)
}
```

Create design matrix and compute a consensus correlation coefficient using
limma's duplicateCorrelation function.

```{r}
block <- anno_filter$batch
design <- model.matrix(~ 1 + individual, data = anno_filter)
```

Compute correlation between replicates.

```{r}
dup_corrs_file <- "../data/dup-corrs.rda"
if (file.exists(dup_corrs_file)) {
  load(dup_corrs_file)
} else{
  dup_corrs <- duplicateCorrelation(molecules_cpm_trans,
                                    design = design, block = block)
  save(dup_corrs, file = dup_corrs_file)
}
str(dup_corrs)
```

Fit a mixed model with the 9 batches being the random effect.

```{r}
if (file.exists("../data/limma-crossed.rda")) {
  load("../data/limma-crossed.rda")
} else {
  gls_fit <- Humanzee::ruv_mixed_model(molecules_cpm_trans,
                      ndups = 1,
                      design = design, block = block,
                      correlation = dup_corrs$cons)
  save(gls_fit, file = "../data/limma-crossed.rda")
}
```

Compute expression levels after removing variation due to random effects.

```{r remove-unwanted-variation}
molecules_final <- t( design %*% t(gls_fit$coef) ) + gls_fit$resid
```

```{r pca-molecules-final}
pca_final <- run_pca(molecules_final)
pca_final_plot <- plot_pca(pca_final$PCs, explained = pca_final$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate") +
  labs(title = "Molecules (batch corrected) for single cells")
pca_final_plot
```



## Session information

```{r info}
sessionInfo()
```
