---
title: "Mixed effect model for batch correction"
date: 2015-09-09
---
  
**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(fig.width = 8, fig.height = 8)
```

```{r packages, message=FALSE}
library("limma")
library("edgeR")
library("ggplot2")
theme_set(theme_bw(base_size = 12))
source("functions.R")
library("Humanzee")
```

This file uses a mixed effects model to remove technical batch effects.
The modeling function is adapted from the package [limma][] and is implemented in the package [Humanzee][].

Creates the following file:

**[molecules-final.txt][]** - Molecules in high quality single cells after removing unwanted variation with mixed model

[molecules-final.txt]: https://github.com/jdblischak/singleCellSeq/blob/master/data/molecules-final.txt
[limma]: http://www.bioconductor.org/packages/release/bioc/html/limma.html
[Humanzee]: https://github.com/jhsiao999/Humanzee

## Input

Input filtered annotation.

```{r input-annotation-filter}
anno_filter <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_filter)
```

Input Poisson GLM transformed molecule counts per million.

```{r input-molecule-counts-trans}
molecules_cpm_trans <- read.table("../data/molecules-cpm-trans.txt", header = TRUE,
                               stringsAsFactors = FALSE)
stopifnot(ncol(molecules_cpm_trans) == nrow(anno_filter),
          colnames(molecules_cpm_trans) == anno_filter$sample_id)
```

## Mixed model for batch-effect correction

Because the [Poisson transformation with the ERCC controls](poison-glm-normalization.html) was not sufficient to remove all the unwanted technical variation, we used a mixed model to correct for batch effects.

We adapted limma's algorithm for estimating variance components due to random effects. 
This analysis operates under the assumption that biological replicates (or batches within an individual
in this case) share similar correlation across genes. Morever, the analysis permits negative
correlation between replicates. 

### Crossed Model

For every single gene, we will fit a mixed model assuming
differences between batches are not individual-specific as follows

$$ y_{ijk} = \mu + \alpha_i + b_j + \epsilon_{ijk} $$,

where $y_{ijk}$ is the log2 counts-per-million (cpm) for any gene 
in individual $i$, batch $j$, and cell $k$, $\mu$ is the gene-specific
expression level across all cells, $\alpha_i$ is the expression level
specific to individual $i$, $b_j$ is batch $j$'s deviation of expression
level from gene-specific expression levels, and $\epsilon_{ijk}$ is 
the models' residual error. 

We assume that $b_j$ follows a normal distribution with $b_j \sim N(0, \sigma^2_b)$ for $j = 1, \dots, 9$, and $\epsilon_{ijk} \sim N(0, \sigma^2_\epsilon)$ for $i = 1, 2, 3; j = 1, \dots, 9; and k = 1, \dots, n_{ij}$, where $n_ij$ denotes the number of cells in individual $i$, batch $j$.

### Remove unwanted variation 

Create design matrix and compute a consensus correlation coefficient using
limma's duplicateCorrelation function.

```{r design-matrix}
block <- anno_filter$batch
design <- model.matrix(~ 1 + individual, data = anno_filter)
```

```{r duplicate-correlation, cache=TRUE}
dup_corrs <- duplicateCorrelation(molecules_cpm_trans, design = design,
                                  block = block)
```

Fit a mixed model with the 9 batches being the random effect.

```{r fit-mixed-model, cache=TRUE}
gls_fit <- Humanzee::ruv_mixed_model(molecules_cpm_trans,
                                     ndups = 1,
                                     design = design,
                                     block = block,
                                     correlation = dup_corrs$cons)
```

Compute expression levels after removing variation due to random effects.

```{r remove-unwanted-variation}
molecules_final <- t( design %*% t(gls_fit$coef) ) + gls_fit$resid
```

Output the cleaned data.

```{r ouptut-mixed-model}
colnames(molecules_final) <- colnames(molecules_cpm_trans)
write.table(round(molecules_final, digits = 6), "../data/molecules-final.txt", quote = FALSE,
            sep = "\t", col.names = NA)
```

## PCA

```{r pca-molecules-final}
pca_final <- run_pca(molecules_final)
pca_final_plot <- plot_pca(pca_final$PCs, explained = pca_final$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate") +
  labs(title = "Batch corrected")
pca_final_plot
```

## Session information

```{r info}
sessionInfo()
```
