---
title: "Mixed effect model for batch correction"
author: "Joyce Hsiao"
date: 2016-01-12
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objective 

Update the batch-corrected counts using data that were processed well-specific technical bias in molecule counts using ERCC molecule counts under a Poisson-based framework. See [link1] for details.

[link1](http://jdblischak.github.io/singleCellSeq/analysis/poisson-glm-normalization.html)


## Setup

```{r}
source("functions.R")
require("limma")
require("edgeR")
require(ggplot2)
require(dplyr)
require(data.table)
theme_set(theme_bw(base_size = 12))
```


## Poisson normalization

Import data: ERCC counts after filtering out low quality single cells are used to compute well-specific bias in ENSG molecule counts. 

```{r}
# Annotation for single cells included in downstream analysis
anno <- read.table("../data/annotation.txt",header=T,stringsAsFactors=F)
quality_single_cells <- scan("../data/quality-single-cells.txt",
                             what = "character")
anno_filter <- anno %>% filter(sample_id %in% quality_single_cells)

# Import expected ERCC counts
spike <- read.table("../data/expected-ercc-molecules.txt", header = TRUE,
                    sep = "\t", stringsAsFactors = FALSE)

# Import molecule counts of all genes (ERCC + ENSG) after filtering out low 
# quality single cells
molecules_filter <- read.table("../data/molecules-filter.txt", header = TRUE,
                               stringsAsFactors = FALSE)

# ERCC observed count
spike_input <- spike$ercc_molecules_well[spike$id %in% rownames(molecules_filter)]
names(spike_input) <- spike$id[spike$id %in% rownames(molecules_filter)]
spike_input <- spike_input[order(names(spike_input))]
tech <- grepl("ERCC", rownames(molecules_filter))
molecules_filter <- as.matrix(molecules_filter)
batches <- unique(anno_filter$batch)

# CPM corrected ENSG counts
molecules_cpm <- fread( "../data/molecules-cpm.txt", header = TRUE,
                         stringsAsFactors = FALSE)
setDF(molecules_cpm)
rownames(molecules_cpm) <- molecules_cpm$V1
molecules_cpm$V1 <- NULL 
molecules_cpm <- as.matrix(molecules_cpm)
```

Apply poisson regression to compute expected-to-observed count transformation paramters 
using ERCC count infomration.

```{r}
pois_glm_params <- lapply(1:3,function(g) numeric(ncol(molecules_cpm)))
names(pois_glm_params) = c("intercept","nu","theta")
for (i in 1:ncol(molecules_cpm)) {
  fit <- glm(molecules_filter[names(spike_input), i] ~ log(spike_input), family="poisson")
  pois_glm_params$intercept[i] <- fit$coefficients[1]
  pois_glm_params$nu[i] <- fit$coefficients[2]
}
pois_glm_expression <- sweep( sweep( molecules_cpm, 2, pois_glm_params$intercept, "-"), 2, pois_glm_params$nu, "/" )

```


PCA

```{r}
molecules_pois_transform <- run_pca(pois_glm_expression)
molecules_pois_transform_plot <- plot_pca(molecules_pois_transform$PCs, 
                                          explained = molecules_pois_transform$explained,
                                          metadata = anno_filter, 
                                          color = "individual",
                                          shape = "replicate") +
                                    labs(title = "After poisson normalization")
```


## Remove unwanted variation

Load the Humanzee package

```{r}
if (!require(Humanzee, quietly = TRUE)) {
  library(devtools)
  install_github("jhsiao999/Humanzee")
  library(Humanzee)
}
```

Create design matrix and compute a consensus correlation coefficient using
limma's duplicateCorrelation function.

```{r}
block <- anno_filter$batch
design <- model.matrix(~ 1 + individual, data = anno_filter)
```

Compute correlation between replicates.

```{r}
dup_corrs_file <- "../data/dup-corrs-poisson.rda"
if (file.exists(dup_corrs_file)) {
  load(dup_corrs_file)
} else{
  dup_corrs <- duplicateCorrelation(pois_glm_expression,
                                    design = design, block = block)
  save(dup_corrs, file = dup_corrs_file)
}
str(dup_corrs)
```

Fit a mixed model with the 8 batches being the random effect.

```{r}
if (file.exists("../data/limma-crossed-poisson.rda")) {
  load("../data/limma-crossed-poisson.rda")
} else {
  gls_fit <- Humanzee::ruv_mixed_model(pois_glm_expression,
                      ndups = 1,
                      design = design, block = block,
                      correlation = dup_corrs$cons)
  save(gls_fit, file = "../data/limma-crossed-poisson.rda")
}
```

Compute expression levels after removing variation due to random effects.

```{r remove-unwanted-variation}
molecules_final <- t( design %*% t(gls_fit$coef) ) + gls_fit$resid
colnames(molecules_final) <- anno_filter$sample_id
rownames(molecules_final) <- rownames(molecules_filter)
```

Export final data

```{r export-final-data}
data_file <- "../data/molecules-final-pois.txt"
if (!file.exists(data_file)) {
  write.table(molecules_final, 
              data_file, quote = FALSE, sep = "\t",
              row.names = TRUE)
}
```


```{r pca-molecules-final}
pca_final <- run_pca(molecules_final)
pca_final_plot <- plot_pca(pca_final$PCs, explained = pca_final$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate") +
  labs(title = "Batch-corrected")
```


## PCA plots

```{r, fig.height = 3}
theme_set(theme_bw(base_size = 8))
cowplot::plot_grid(
    molecules_pois_transform_plot,
    pca_final_plot,
  ncol = 2,
  labels = LETTERS[1:2])
```




## Session information

```{r info}
sessionInfo()
```
