---
title: "Compare read and molecule counts per batch"
date: 2015-08-18
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(cache = TRUE)
```

Comparing the conversion of reads to molecules for each of the 9 batches.
Used three different metrics:

*  Raw counts
*  Log2 counts (pseudocount of 1)
*  Log2 TMM-normalized counts per million (pseudocount of 0.25)

## Input

```{r packages, message=FALSE, cache=FALSE}
library("dplyr")
library("ggplot2")
theme_set(theme_bw(base_size = 16))
library("edgeR")
source("functions.R")
library("tidyr")
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input read counts.

```{r input-read-counts}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input list of quality single cells.

```{r input-quality-single-cells}
quality_single_cells <- scan("../data/quality-single-cells.txt",
                             what = "character")
```

## Filter

Keep only the single cells that passed the [QC filters](qc-cell-ipsc.html).

```{r filter-cells}
reads <- reads[, colnames(reads) %in% quality_single_cells]
molecules <- molecules[, colnames(molecules) %in% quality_single_cells]
anno <- anno[anno$sample_id %in% quality_single_cells, ]
stopifnot(dim(reads) == dim(molecules),
          nrow(anno) == ncol(reads))
```

Only keep the following genes:

*  ERCC genes with at least one molecule observed in at least one single cell

```{r keep-ercc}
ercc_keep <- rownames(molecules)[grepl("ERCC", rownames(molecules)) &
                                 rowSums(molecules) > 1]
```

*  The top expressed endogenous genes in terms of mean molecule counts per million

```{r top-expressed-genes}
num_genes <- 12000
mean_cpm <- molecules %>%
  filter(!grepl("ERCC", rownames(molecules))) %>%
  cpm %>%
  rowMeans
gene_keep <- rownames(molecules)[!grepl("ERCC", rownames(molecules))][order(mean_cpm, decreasing = TRUE)][1:num_genes]
```

Filter the genes:

```{r filter-genes}
reads <- reads[rownames(reads) %in% c(gene_keep, ercc_keep), ]
molecules <- molecules[rownames(molecules) %in% c(gene_keep, ercc_keep), ]
```

## Transformation

In addition to comparing the raw counts of reads and molecules, we compare the log2 counts and the log counts per million.

For the log counts, I add a pseudocount of 1.

```{r log-counts}
reads_log <- log2(reads + 1)
molecules_log <- log2(molecules + 1)
```

Calculate cpm for the reads data using TMM-normalization.

```{r calc-cpm-reads}
norm_factors_reads <- calcNormFactors(reads, method = "TMM")
reads_cpm <- cpm(reads, lib.size = colSums(reads) * norm_factors_reads,
                 log = TRUE)
```

And for the molecules.

```{r calc-cpm-molecules}
norm_factors_mol <- calcNormFactors(molecules, method = "TMM")
molecules_cpm <- cpm(molecules, lib.size = colSums(molecules) * norm_factors_mol,
                     log = TRUE)
```

## Differences in conversion of reads to molecules

As seen above with the total counts, the conversion of reads to molecules varies between each of the 9 batches.
Below I fit a loess transformation to each individually.
It would be ideal if we could somehow we could correct for these differences and have them all follow a similar transformation from reads to molecules.

```{r gather-cols-to-rows}
convert_to_long <- function(r, m) {
  # Combines reads and molecules into long format for comparison of conversion
  #
  # r - reads in wide format
  # m - molecules in wide format
  r <- data.frame(gene = rownames(r), r)
  m <- data.frame(gene = rownames(m), m)
  r_long <- gather_(r, key = "id", value = "reads",
                    grep("NA", colnames(r), value = TRUE))
  m_long <- gather_(m, key = "id", value = "molecules",
                    grep("NA", colnames(m), value = TRUE))
  r_long <- separate_(r_long, col = "id",  sep = "\\.", remove = FALSE,
                     into = c("individual", "batch", "well"))
  stopifnot(r_long$id == m_long$id,
            r_long$gene == m_long$gene)
  conversion <- cbind(r_long, m_long$molecules)
  colnames(conversion)[ncol(conversion)] <- "molecules"
  stopifnot(nrow(conversion) == nrow(m_long))
  return(conversion)
}
```

In order to be able to make this plot, I have to subsample to fewer genes.
Otherwise it runs out of memory.

```{r subsample, cache=FALSE}
set.seed(12345)
num_subsampled <- 5000
sub_indices <- sample(1:nrow(reads), num_subsampled)
# counts
reads_sub <- reads[sub_indices, ]
molecules_sub <- molecules[sub_indices, ]
# log counts
reads_log_sub <- reads_log[sub_indices, ]
molecules_log_sub <- molecules_log[sub_indices, ]
# log counts per million
reads_cpm_sub <- reads_cpm[sub_indices, ]
molecules_cpm_sub <- molecules_cpm[sub_indices, ]
# Convert to long format
conversion <- convert_to_long(reads_sub, molecules_sub)
conversion_log <- convert_to_long(reads_log_sub, molecules_log_sub)
conversion_cpm <- convert_to_long(reads_cpm_sub, molecules_cpm_sub)
head(conversion)
```

Summarize across the single cells for each of the 9 batches.

```{r summarize-conversion, cache=FALSE}
# counts
conversion_mean <- conversion %>%
  filter(well != "bulk") %>%
  group_by(individual, batch, gene) %>%
  summarize(reads_mean = mean(reads),
            reads_sem = sd(reads) / sqrt(length(reads)),
            molecules_mean = mean(molecules),
            molecules_sem = sd(molecules) / sqrt(length(molecules)))
# log counts
conversion_log_mean <- conversion_log %>%
  filter(well != "bulk") %>%
  group_by(individual, batch, gene) %>%
  summarize(reads_mean = mean(reads),
            reads_sem = sd(reads) / sqrt(length(reads)),
            molecules_mean = mean(molecules),
            molecules_sem = sd(molecules) / sqrt(length(molecules)))

# counts per million
conversion_cpm_mean <- conversion_cpm %>%
  filter(well != "bulk") %>%
  group_by(individual, batch, gene) %>%
  summarize(reads_mean = mean(reads),
            reads_sem = sd(reads) / sqrt(length(reads)),
            molecules_mean = mean(molecules),
            molecules_sem = sd(molecules) / sqrt(length(molecules)))
head(conversion_cpm_mean)
```

### Counts

Compare the counts.

```{r convert-reads-molecules-per-batch, fig.width=9, fig.height=9, cache=FALSE}
conver_plot_counts <- ggplot(conversion_mean,
                             aes(x = reads_mean, y = molecules_mean, col = individual,
                                 shape = as.factor(batch))) +
  geom_point() +
  geom_errorbar(aes(ymin = molecules_mean - molecules_sem,
                    ymax = molecules_mean + molecules_sem)) +
  geom_errorbarh(aes(xmin = reads_mean - reads_sem,
                     xmax = reads_mean + reads_sem)) +
  geom_smooth(method = "loess", color = "black") +
  facet_grid(batch ~ individual) +
  theme(legend.position = "none") +
  labs(x = "Reads per gene",
       y = "Molecules per gene",
       title = sprintf("Conversion per batch (subsample of %d genes)", num_subsampled))
conver_plot_counts
```

### Log2 counts

Compare the log counts.

```{r convert-reads-molecules-per-batch-log, fig.width=9, fig.height=9, cache=FALSE}
conver_plot_log <- conver_plot_counts %+%
  conversion_log_mean +
  geom_line(aes(x = reads_mean, y = reads_mean)) +
  labs(x = "Reads per gene (log2)",
       y = "Molecules per gene (log2)",
       title = sprintf("Conversion per batch (subsample of %d genes) - log", num_subsampled))
conver_plot_log
```

### Log2 counts per million

Compare the counts per million.

```{r convert-reads-molecules-per-batch-cpm, fig.width=9, fig.height=9, cache=FALSE}
conver_plot_cpm <- conver_plot_log %+%
  conversion_cpm_mean +
  labs(x = "Reads per million per gene (log2)",
       y = "Molecules per million per gene (log2)",
       title = sprintf("Conversion per batch (subsample of %d genes) - cpm", num_subsampled))
conver_plot_cpm + geom_line(aes(x = reads_mean, y = reads_mean))
```

Now visualizing only the ERCC.

```{r prepare-ercc, cache=FALSE}
conversion_ercc <- convert_to_long(reads[grep("ERCC", rownames(reads)), ],
                                   molecules[grep("ERCC", rownames(molecules)), ])
conversion_ercc_log <- convert_to_long(reads_log[grep("ERCC", rownames(reads_log)), ],
                                       molecules_log[grep("ERCC", rownames(molecules_log)), ])
conversion_ercc_cpm <- convert_to_long(reads_cpm[grep("ERCC", rownames(reads_cpm)), ],
                                       molecules_cpm[grep("ERCC", rownames(molecules_cpm)), ])
# Remove 19098 batch 2 because the outlier throws off the axes
conversion_ercc <- conversion_ercc[!(conversion_ercc$individual == "NA19098" &
                                     conversion_ercc$batch == 2), ]
conversion_ercc_log <- conversion_ercc_log[!(conversion_ercc_log$individual == "NA19098" &
                                             conversion_ercc_log$batch == 2), ]
conversion_ercc_cpm <- conversion_ercc_cpm[!(conversion_ercc_cpm$individual == "NA19098" &
                                             conversion_ercc_cpm$batch == 2), ]
```

```{r summarize-conversion-ercc, cache=FALSE}
# counts
conversion_ercc_mean <- conversion_ercc %>%
  filter(well != "bulk") %>%
  group_by(individual, batch, gene) %>%
  summarize(reads_mean = mean(reads),
         reads_sem = sd(reads) / sqrt(length(reads)),
         molecules_mean = mean(molecules),
         molecules_sem = sd(molecules) / sqrt(length(molecules)))
# log counts
conversion_ercc_log_mean <- conversion_ercc_log %>%
  filter(well != "bulk") %>%
  group_by(individual, batch, gene) %>%
  summarize(reads_mean = mean(reads),
         reads_sem = sd(reads) / sqrt(length(reads)),
         molecules_mean = mean(molecules),
         molecules_sem = sd(molecules) / sqrt(length(molecules)))
# counts per million
conversion_ercc_cpm_mean <- conversion_ercc_cpm %>%
  filter(well != "bulk") %>%
  group_by(individual, batch, gene) %>%
  summarize(reads_mean = mean(reads),
         reads_sem = sd(reads) / sqrt(length(reads)),
         molecules_mean = mean(molecules),
         molecules_sem = sd(molecules) / sqrt(length(molecules)))
head(conversion_ercc_cpm_mean)
```

### ERCC counts

```{r convert-reads-molecules-per-batch-ercc, fig.width=9, fig.height=9, cache=FALSE}
conver_plot_ercc <- conver_plot_counts %+%
  conversion_ercc_mean +
  labs(title = "Conversion per batch (ERCC genes)")
conver_plot_ercc
```

### ERCC log2 counts

```{r convert-reads-molecules-per-batch-ercc-log, fig.width=9, fig.height=9, cache=FALSE}
conver_plot_ercc_log <- conver_plot_log %+%
  conversion_ercc_log_mean +
  labs(x = "Reads per million per gene (log2)",
       y = "Molecules per million per gene (log2)",
       title = "Conversion per batch (ERCC genes) - log")
conver_plot_ercc_log
```

### ERCC log2 counts per million

```{r convert-reads-molecules-per-batch-ercc-cpm, fig.width=9, fig.height=9, cache=FALSE}
conver_plot_ercc_cpm <- conver_plot_log %+%
  conversion_ercc_cpm_mean +
  labs(x = "Reads per million per gene",
       y = "Molecules per million per gene",
       title = "Conversion per batch (ERCC genes) - cpm")
conver_plot_ercc_cpm
```

## Comparing loess curves

```{r}
conversion_chunk <- filter(conversion_cpm_mean, individual == "NA19101", batch == 1)
dim(conversion_chunk)
head(conversion_chunk)
loess_model <- loess(molecules_mean ~ reads_mean, data = conversion_chunk)
loess_predict <- predict(loess_model)
str(loess_predict)
plot(conversion_chunk$reads_mean, loess_predict)
abline(a = 0, b = 1, col = "red")
plot(conversion_chunk$molecules_mean, loess_predict)
plot(conversion_chunk$reads_mean, conversion_chunk$reads_mean - loess_predict)
```

## Session information

```{r info}
sessionInfo()
```
