---
title: "Random effect model to compare variations"
author: "Joyce Hsiao"
date: 2015-10-07
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objectives


## Set up

```{r}
library("data.table")
library("dplyr")
library("limma")
library("edgeR")
library("ggplot2")
library("grid")
theme_set(theme_bw(base_size = 12))
source("functions.R")
```



## Prepare data


Input annotation of only QC-filtered single cells

```{r}
anno_qc <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_qc)
```


Input molecule counts that are filtered, transformed, and also processed to remove unwanted variation from batch effet. ERCC genes are also removed.

```{r}
molecules_qc <- read.table("../data/molecules-final.txt", header = TRUE, stringsAsFactors = FALSE)
```


## Compute coefficient of variation

Compute per batch coefficient of variation.

Include only genes with positive coefficient of variation. Some genes in this data may have zero coefficient of variation, because we include gene with more than 0 count 
across all cells.


```{r}
molecules_cv_batch <- 
  lapply(1:length(unique(anno_qc$batch)), function(per_batch) {
        molecules_per_batch <- molecules_qc[ , unique(anno_qc$batch) == unique(anno_qc$batch)[per_batch] ]
        mean_per_gene <- apply(molecules_per_batch, 1, mean, na.rm = TRUE)
        sd_per_gene <- apply(molecules_per_batch, 1, sd, na.rm = TRUE)
        cv_per_gene <- data.frame(mean = mean_per_gene,
                                  cv = sd_per_gene/mean_per_gene)
        rownames(cv_per_gene) <- rownames(molecules_qc)
    
        cv_per_gene <- cv_per_gene[rowSums(is.na(cv_per_gene)) == 0, ]
        cv_per_gene$batch <- unique(anno_qc$batch)[per_batch]
        return(cv_per_gene)
      }) 
names(molecules_cv_batch) <- unique(anno_qc$batch)

dim(molecules_cv_batch[[1]])
```



## F-test


```{r}
individual <- anno_qc$individual

var_Ftest <- 
  lapply(1:dim(molecules_qc)[1], function(per_gene) {
        values <- unlist(molecules_qc[per_gene, ])
        data.frame(NA19098_19101 = var.test(values[individual == "NA19098"],
                                            values[individual == "NA19101"])$p.value,
                   NA19098_19239 = var.test(values[individual == "NA19098"],
                                            values[individual == "NA19239"])$p.value, 
                   NA19239_19101 = var.test(values[individual == "NA19239"],
                                            values[individual == "NA19101"])$p.value )
  })
var_Ftest <- do.call(rbind, var_Ftest)

head(var_Ftest)
```

p-value distribution

```{r}
par(mfrow = c(2,2))
hist(var_Ftest[ ,1], main = colnames(var_Ftest)[1], breaks = 100)
hist(var_Ftest[ ,2], main = colnames(var_Ftest)[2], breaks = 100)
hist(var_Ftest[ ,3], main = colnames(var_Ftest)[3], breaks = 100)
```


Does p-value ranking track with CV difference?

```{r}

```


## Brown-Forsythe test


## Leven's test



## Session information

```{r info}
sessionInfo()
```
