---
title: "Levene's test and CV"
author: "Joyce Hsiao"
date: 2015-10-07
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objectives


## Set up

```{r, message=FALSE, warning=FALSE}
library("data.table")
library("dplyr")
library("limma")
library("edgeR")
library("ggplot2")
library("grid")
theme_set(theme_bw(base_size = 12))
source("functions.R")
```



## Prepare data


Input annotation of only QC-filtered single cells

```{r}
anno_qc <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_qc)
```


Input molecule counts that are filtered, transformed, and also processed to remove unwanted variation from batch effet. ERCC genes are also removed.

```{r}
molecules_qc <- read.table("../data/molecules-final.txt", header = TRUE, stringsAsFactors = FALSE)
```


## Compute coefficient of variation

Compute per batch coefficient of variation.

Include only genes with positive coefficient of variation. Some genes in this data may have zero coefficient of variation, because we include gene with more than 0 count 
across all cells.


```{r}
molecules_cv_batch <- 
  lapply(1:length(unique(anno_qc$batch)), function(per_batch) {
        molecules_per_batch <- molecules_qc[ , unique(anno_qc$batch) == unique(anno_qc$batch)[per_batch] ]
        mean_per_gene <- apply(molecules_per_batch, 1, mean, na.rm = TRUE)
        sd_per_gene <- apply(molecules_per_batch, 1, sd, na.rm = TRUE)
        cv_per_gene <- data.frame(mean = mean_per_gene,
                                  sd = sd_per_gene,
                                  cv = sd_per_gene/mean_per_gene)
        rownames(cv_per_gene) <- rownames(molecules_qc)
    
        cv_per_gene <- cv_per_gene[rowSums(is.na(cv_per_gene)) == 0, ]
        cv_per_gene$batch <- unique(anno_qc$batch)[per_batch]
        return(cv_per_gene)
      }) 
names(molecules_cv_batch) <- unique(anno_qc$batch)

dim(molecules_cv_batch[[1]])
```



## F-test


```{r}
individual <- anno_qc$individual

var_Ftest <- 
  lapply(1:dim(molecules_qc)[1], function(per_gene) {
        values <- unlist(molecules_qc[per_gene, ])
        data.frame(NA19098_19101 = var.test(values[individual == "NA19098"],
                                            values[individual == "NA19101"])$p.value,
                   NA19098_19239 = var.test(values[individual == "NA19098"],
                                            values[individual == "NA19239"])$p.value, 
                   NA19239_19101 = var.test(values[individual == "NA19239"],
                                            values[individual == "NA19101"])$p.value )
  })
var_Ftest <- do.call(rbind, var_Ftest)

head(var_Ftest)
```

p-value distribution

```{r}
par(mfrow = c(2,2))
hist(var_Ftest[ ,1], main = colnames(var_Ftest)[1], 
     breaks = 100, xlab = "p-value of the F statistic")
hist(var_Ftest[ ,2], main = colnames(var_Ftest)[2], 
     breaks = 100, xlab = "p-value of the F statistic")
hist(var_Ftest[ ,3], main = colnames(var_Ftest)[3], 
     breaks = 100, xlab = "p-value of the F statistic")
title(main = "Levene's test (based on normality)", outer = TRUE)
```


## F-test versus indiviudal CVs

Compute individual CVs by taking the average of batch CVs.

```{r}
molecules_cv_indiviudal <-
  data.frame(NA19098 = rowMeans(cbind(molecules_cv_batch[[1]]$cv,
                                      molecules_cv_batch[[2]]$cv,
                                      molecules_cv_batch[[3]]$cv) ),
             NA19101 = rowMeans(cbind(molecules_cv_batch[[4]]$cv,
                                      molecules_cv_batch[[5]]$cv,
                                      molecules_cv_batch[[6]]$cv) ),
             NA19239 = rowMeans(cbind(molecules_cv_batch[[7]]$cv,
                                      molecules_cv_batch[[8]]$cv,
                                      molecules_cv_batch[[9]]$cv) ) )
```

Plot CVs.


```{r}
require(scales)
require(broman)
crayons <- brocolors("crayons")

par(mfrow = c(2,2))
df_xy <- data.frame(x = molecules_cv_indiviudal$NA19098,
                    y = molecules_cv_indiviudal$NA19101)
plot(df_xy,
     xlab = "NA19098 CV", ylab = "NA19101 CV", pch = 1)
select_pts <- var_Ftest$NA19098_19101 < quantile(var_Ftest$NA19098_19101, prob = .01)
points(df_xy[ select_pts, ], cex = .6, 
     pch = 16, col = alpha(crayons["Orange Red"], 1) )

df_xy <- data.frame(x = molecules_cv_indiviudal$NA19239,
                    y = molecules_cv_indiviudal$NA19101)
plot(df_xy,
     xlab = "NA19239 CV", ylab = "NA19101 CV", pch = 1)
select_pts <- var_Ftest$NA19239_19101 < quantile(var_Ftest$NA19239_19101, prob = .01)
points(df_xy[ select_pts, ], cex = .6,
     pch = 16, col = alpha(crayons["Orange Red"], 1) )

plot(0, axes = F, ann = F, pch = "")

df_xy <- data.frame(x = molecules_cv_indiviudal$NA19098,
                    y = molecules_cv_indiviudal$NA19239)
plot(df_xy,
     xlab = "NA19098 CV", ylab = "NA19239 CV", pch = 1)
select_pts <- var_Ftest$NA19098_19239 < quantile(var_Ftest$NA19098_19239, prob = .01)
points(df_xy[ select_pts, ], cex = .6,
     pch = 16, col = alpha(crayons["Orange Red"], 1) )

title(main = "Top 1000 sig. genes", outer = TRUE, line = -1)

```


## F-test versus indiviudal SDs

Compute individual SDs by taking the average of batch SDs.

```{r}
molecules_sd_indiviudal <-
  data.frame(NA19098 = rowMeans(cbind(molecules_cv_batch[[1]]$sd,
                                      molecules_cv_batch[[2]]$sd,
                                      molecules_cv_batch[[3]]$sd) ),
             NA19101 = rowMeans(cbind(molecules_cv_batch[[4]]$sd,
                                      molecules_cv_batch[[5]]$sd,
                                      molecules_cv_batch[[6]]$sd) ),
             NA19239 = rowMeans(cbind(molecules_cv_batch[[7]]$sd,
                                      molecules_cv_batch[[8]]$sd,
                                      molecules_cv_batch[[9]]$sd) ) )
```

Plot standard deviations.

```{r}
require(scales)
require(broman)
crayons <- brocolors("crayons")

par(mfrow = c(2,2))
df_xy <- data.frame(x = molecules_sd_indiviudal$NA19098,
                    y = molecules_sd_indiviudal$NA19101)
plot(df_xy,
     xlab = "NA19098 SD", ylab = "NA19101 SD", pch = 1)
select_pts <- var_Ftest$NA19098_19101 < quantile(var_Ftest$NA19098_19101, prob = .01)
points(df_xy[ select_pts, ], cex = .6, 
     pch = 16, col = alpha(crayons["Orange Red"], 1) )


df_xy <- data.frame(x = molecules_sd_indiviudal$NA19239,
                    y = molecules_sd_indiviudal$NA19101)
plot(df_xy,
     xlab = "NA19239 SD", ylab = "NA19101 SD", pch = 1)
select_pts <- var_Ftest$NA19239_19101 < quantile(var_Ftest$NA19239_19101, prob = .01)
points(df_xy[ select_pts, ], cex = .6,
     pch = 16, col = alpha(crayons["Orange Red"], 1) )

plot(0, axes = F, ann = F, pch = "")

df_xy <- data.frame(x = molecules_sd_indiviudal$NA19098,
                    y = molecules_sd_indiviudal$NA19239)
plot(df_xy,
     xlab = "NA19098 SD", ylab = "NA19239 SD", pch = 1)
select_pts <- var_Ftest$NA19098_19239 < quantile(var_Ftest$NA19098_19239, prob = .01)
points(df_xy[ select_pts, ], cex = .6,
     pch = 16, col = alpha(crayons["Orange Red"], 1) )

title(main = "Top 1000 sig. genes", outer = TRUE, line = -1)

```



## Session information

```{r info}
sessionInfo()
```
