---
title: "Batch effects of conversion efficiency"
date: 2015-10-23
author: Joyce Hsiao & PoYuan Tung
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
# opts_chunk$set(cache = TRUE)
```

## Objective

Compute conversion efficiency of each cell using the filerted set of cells. Compare conversion efficiency between batches: is there an individual effect or batch effect of coversion efficiency?

The definition of conversion efficiency: total molecule counts/ total read counts

Note that we will remove the NA19098 batch 2 in the following analyses. 

We will use analysis of variance to test whether conversion rates are significantly different between individuals. Results below are described at an alpha level of .05.


## Input

```{r packages, message=FALSE}
library("dplyr")
library("ggplot2")
library("cowplot")
theme_set(theme_bw(base_size = 16))
library("edgeR")
source("functions.R")
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input read counts.

```{r input-read-counts}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input list of quality single cells.

```{r input-quality-single-cells}
quality_single_cells <- scan("../data/quality-single-cells.txt",
                             what = "character")
```

Keep only the single cells that passed the [QC filters](qc-filter-ipsc.html).

```{r filter-cells}
reads <- reads[, colnames(reads) %in% quality_single_cells]
molecules <- molecules[, colnames(molecules) %in% quality_single_cells]
anno <- anno[anno$sample_id %in% quality_single_cells, ]
stopifnot(dim(reads) == dim(molecules),
          nrow(anno) == ncol(reads))
```

Seperate ERCC and endogenous genes

```{r ercc-genes}
reads_ENSG <- reads[grep("ERCC", rownames(reads), invert = TRUE), ]
reads_ERCC <- reads[grep("ERCC", rownames(reads), invert = FALSE), ]

molecules_ENSG <- molecules[grep("ERCC", rownames(molecules), invert = TRUE), ]
molecules_ERCC <- molecules[grep("ERCC", rownames(molecules), invert = FALSE), ]
```

## Reads to molecule conversion efficiency

```{r conversion-rate}
total_counts_ERCC <- data.frame(total_reads = colSums(reads_ERCC),
                                total_molecules = colSums(molecules_ERCC))
total_counts_ERCC$conversion <- with(total_counts_ERCC,
                                     total_molecules/total_reads)
summary(total_counts_ERCC)


total_counts_ENSG <- data.frame(total_reads = colSums(reads_ENSG),
                                total_molecules = colSums(molecules_ENSG))
total_counts_ENSG$conversion <- with(total_counts_ENSG,
                                     total_molecules/total_reads)
summary(total_counts_ENSG)
```


```{r plot-convertion}
## create a color palette with one color per individual and different shades for repplicates
great_color <- c("#CC3300", "#FF9966", "#006633", "#009900", "#99FF99", "#3366FF", "#6699FF", "#66CCFF")

depth_plot_ensg <- ggplot(data.frame(total_counts_ENSG,
                  batch = anno$batch), 
       aes(x = total_reads/10^6, y = total_molecules/10^3, col = factor(batch))) +
  geom_point() +
  scale_color_manual(values = great_color) +
  labs(x = "Total number of reads (x10^6)",
       y = "Total number of molecules (x10^3)",
       title = "Effect of read depth on single cells (genes)")

depth_plot_ensg

depth_plot_ercc <- ggplot(data.frame(total_counts_ERCC,
                  batch = anno$batch), 
       aes(x = total_reads/10^6, y = total_molecules/10^3, col = factor(batch))) +
  geom_point() +
  scale_color_manual(values = great_color) +
  labs(x = "Total number of reads (x10^6)",
       y = "Total number of molecules (x10^3)",
       title = "Effect of read depth on single cells (ERCC spike-in)")

depth_plot_ercc
```

## Compare variances of conversion rates between batches

### ENSG

```{r}
convertion_ensg <- ggplot(data.frame(total_counts_ENSG,
                  batch = anno$batch),
       aes(x = factor(batch), y = conversion,
           fill = factor(batch)), height = 600, width = 2000) +
geom_violin(alpha = .5) + 
geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
  scale_fill_manual(values = great_color) +
labs(x = "batch", y = "conversion efficiency", title = "ENSG batch effect of conversion efficiency") +
theme(axis.text.x = element_text(hjust=1, angle = 45))

convertion_ensg
```


*Between individual

For individual $i$, sample $j$, we have $n_{ij}$ cells. Here we compare the per cell conversion rate from molecule count to read count between the three individuals:

$$ Y_{ijk} = \mu_i + b_{ij} + \epsilon_{ijk} $$

where $i = 1, 2, 3$, $j = 1, 2, 3$, and $k = 1, 2, \dots, n_{ij}$.


```{r}
df_ENSG <- data.frame(total_counts_ENSG,
                      individual = factor(anno$individual),
                      replicate = factor(anno$replicate),
                      batch = factor(anno$batch))

library(lme4)
lmer_ENSG_full <- lmer(conversion ~ individual + (1| individual:replicate), 
                        data = df_ENSG)
lmer_ENSG_null <- lmer(conversion ~ 1 + (1| individual:replicate), 
                        data = df_ENSG)
anova(lmer_ENSG_null, lmer_ENSG_full)
```


*For each individual, between batches.

```{r}
for (i in 1:length(unique(anno$individual))) {
  print(unique(anno$individual)[i])
  select_individual <- with(anno, individual == unique(individual)[i])
  print( summary(aov(conversion ~ batch, 
                     data = df_ENSG[select_individual, ]) ) )
}
```

### ERCC

```{r}
convertion_ercc <- ggplot(data.frame(total_counts_ERCC,
                  batch = anno$batch),
       aes(x = factor(batch), y = conversion,
           fill = factor(batch)), height = 600, width = 2000) +
geom_violin(alpha = .5) + 
geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
  scale_fill_manual(values = great_color) +
labs(x = "batch", y = "conversion efficiency", title = "ERCC batch effect of conversion efficiency") +
theme(axis.text.x = element_text(hjust=1, angle = 45))

convertion_ercc
```

* Between individuals

```{r}
df_ERCC <- data.frame(total_counts_ERCC,
                      individual = factor(anno$individual),
                      replicate = factor(anno$replicate),
                      batch = factor(anno$batch))

library(lme4)
lmer_ERCC_full <- lmer(conversion ~ individual + (1| individual:replicate), 
                        data = df_ERCC)
lmer_ERCC_null <- lmer(conversion ~ 1 + (1| individual:replicate), 
                        data = df_ERCC)
anova(lmer_ERCC_null, lmer_ERCC_full)
```

*For each individual, between batches.

```{r}
for (i in 1:length(unique(anno$individual))) {
  print(unique(anno$individual)[i])
  select_individual <- with(anno, individual == unique(individual)[i])
  print( summary(aov(conversion ~ batch, 
                     data = df_ERCC[select_individual, ]) ) )
}
```

## Summary plots

```{r, fig.width = 12, fig.height=6}
library(gridExtra)
grid.arrange(
  ggplot(data.frame(total_counts_ERCC,
                    batch = anno$batch),
         aes(x = factor(batch), y = conversion,
             fill = factor(batch)), height = 600, width = 2000) +
  geom_violin(alpha = .5) + 
  geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
    scale_fill_manual(values = great_color) +
  labs(x = "batch", y = "conversion rate", title = "Batch effect of conversion rate (ERCC spike-in)") +
  theme(axis.text.x = element_text(hjust=1, angle = 45))
  ,
  ggplot(data.frame(total_counts_ENSG,
                    batch = anno$batch),
         aes(x = factor(batch), y = conversion,
             fill = factor(batch)), height = 600, width = 2000) +
  geom_violin(alpha = .5) + 
  geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
    scale_fill_manual(values = great_color) +
  labs(x = "batch", y = "conversion rate", title = "Batch effect of conversion rate (genes)") +
  theme(axis.text.x = element_text(hjust=1, angle = 45)),
  ncol = 2
)

```

```{r plot-reads-molecule, fig.width=12, fig.height=8}
theme_set(theme_bw(base_size = 12))
plot_grid(depth_plot_ensg + theme(legend.position = "none"), 
          convertion_ensg + theme(legend.position = "none"),
          depth_plot_ercc + theme(legend.position = "none"),
          convertion_ercc + theme(legend.position = "none"),
          labels = LETTERS[1:4])
```

## Session information

```{r info}
sessionInfo()
```
