---
title: "Coverage of endogenous genes"
date: 2015-02-22
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(cache = TRUE)
```

The sequencing coverage from the UMI protocol should show a very strong 5' bias.
Do we observe this in our data?
Here we explore this in a few samples using the [genomation][] package.
Specifically, we calculate the mean coverage across all the genes that passed our expression filters for two regions:

*  The transcription start site (TSS) +/- 1 kb
*  The transcription end site (TES) +/- 1 kb

[genomation]: http://www.bioconductor.org/packages/release/bioc/html/genomation.html

```{r packages, message=FALSE, cache=FALSE}
library("genomation")
library("Rsamtools")
library("plyr")
library("tidyr")
library("ggplot2")
theme_set(theme_bw(base_size = 14))
theme_update(panel.grid.minor.x = element_blank(),
             panel.grid.minor.y = element_blank(),
             panel.grid.major.x = element_blank(),
             panel.grid.major.y = element_blank())
```

## Input

Input filtered molecule counts.

```{r input-molecule-counts-filter}
molecules_filter <- read.table("../data/molecules-filter.txt", header = TRUE,
                               stringsAsFactors = FALSE)
```

## Prepare bam files from high quality cells

To investigate the coverage, we select one high quality cell from each individual.
Note that for the actual data files, the names are more succinct to keep them shorter.

```{r select-quality-cells}
quality_cells <- c("19098.1.A01", "19101.1.A02", "19239.1.A01")
names(quality_cells) <- c("NA19098.r1.A01", "NA19101.r1.A02", "NA19239.r1.A01")
stopifnot(names(quality_cells) %in% colnames(molecules_filter))
```

From the [sequencing pipeline][pipeline], the bam files for the molecules are in `bam-rmdup-umi` and have the filename structure `individual.replicate.well.trim.sickle.sorted.combined.rmdup.bam`.
Unfortunately, after processing with UMI-tools, the bam files are no longer sorted.
It is necessary to sort them and then index them to use with [genomation][].

[pipeline]: process-samples.html

```{r prepare-bam}
bam_molecules_unsorted <- paste0(quality_cells, ".trim.sickle.sorted.combined.rmdup.bam")
data_dir <- "/mnt/gluster/home/jdblischak/ssd"
from_file <- file.path(data_dir, "bam-rmdup-umi", bam_molecules_unsorted)
to_file <- file.path("../data", bam_molecules_unsorted)
sorted_file_prefix <- paste0(sub(".bam$", "", to_file), ".sorted")
sorted_file <- paste0(sorted_file_prefix, ".bam")
indexed_file <- paste0(sorted_file, ".bai")
for (f in 1:length(bam_molecules_unsorted)) {
  if (!file.exists(indexed_file[f])) {
    stopifnot(file.exists(from_file[f]))
    file.copy(from_file[f], to_file[f])
    sortBam(to_file[f], sorted_file_prefix[f])
    indexBam(sorted_file[f])
  }
}
stopifnot(file.exists(indexed_file))
bam <- sorted_file
bam
```

## Prepare genomic features

The genomic features are created with the script [create-transcripts.R](https://github.com/jdblischak/singleCellSeq/blob/master/code/create-transcripts.R).

Input transcription start sites (TSS).

```{r input-tss}
tss <- readBed("../data/tss.bed")
tss <- tss[tss$name %in% rownames(molecules_filter)]
```

Input transcription end sites (TES).

```{r input-tes}
tes <- readBed("../data/tes.bed")
tes <- tes[tes$name %in% rownames(molecules_filter)]
```

## Calculate coverage over genomic features

TSS

```{r coverage-tss}
tss_sm = ScoreMatrixList(target = bam, windows = tss, type = "bam",
                         rpm = TRUE, strand.aware = TRUE)
tss_sm
```

TES

```{r coverage-tes}
tes_sm = ScoreMatrixList(target = bam, windows = tes, type = "bam",
                         rpm = TRUE, strand.aware = TRUE)
tes_sm
```

## Summarize coverage

Calculate the mean coverage per base pair for the TSS and and TES.

```{r gather-tss}
names(tss_sm) <- names(quality_cells)
tss_sm_df <- ldply(tss_sm, colMeans, .id = "sample_id")
colnames(tss_sm_df)[-1] <- paste0("p", 1:(ncol(tss_sm_df) - 1))
tss_sm_df$feature = "TSS"
tss_sm_df_long <- gather(tss_sm_df, key = "pos", value = "rpm", p1:p2001)
```

```{r gather-tes}
names(tes_sm) <- names(quality_cells)
tes_sm_df <- ldply(tes_sm, colMeans, .id = "sample_id")
colnames(tes_sm_df)[-1] <- paste0("p", 1:(ncol(tes_sm_df) - 1))
tes_sm_df$feature = "TES"
tes_sm_df_long <- gather(tes_sm_df, key = "pos", value = "rpm", p1:p2001)
```

Combine the two features.

```{r combine-features}
features <- rbind(tss_sm_df_long, tes_sm_df_long)
# Convert base position back to integer value
features$pos <- sub("p", "", features$pos)
features$pos <- as.numeric(features$pos)
# Subtract 1001 to recalibrate as +/- 1 kb
features$pos <- features$pos - 1001
# Order factor so that TSS is displayed left of TES
features$feature <- factor(features$feature, levels = c("TSS", "TES"),
                           labels = c("Transcription start site",
                                      "Transription end site"))
```

## Metagene plot

```{r metaplot, fig.width=8, cache=FALSE}
ggplot(features, aes(x = pos, y = rpm, color = sample_id)) +
  geom_line() +
  facet_wrap(~feature) +
  scale_color_discrete(name = "Sample") +
  labs(x = "Relative position (bp)",
       y = "Counts per million (mean)",
       title = "5' bias of UMI protocol") +
  theme(legend.position = "bottom")
```

## Caveat

The main caveat of this analysis is the very low coverage per gene of each single cell.

```{r mean-expression}
molecules_filter_sub <- molecules_filter[, names(quality_cells)]
colSums(molecules_filter_sub) / 10^3
mean_expr <- rowMeans(molecules_filter_sub)
summary(mean_expr)
```

The median across the genes for the mean number of molecules across these three samples is only `r median(mean_expr)` molecules.

## Session information

```{r info}
sessionInfo()
```
