---
title: "Per gene statistical comparison of CVs"
author: "Joyce Hsiao"
date: 2015-11-12
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")

library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE)
```


## Objective

Quantify statistical significance of individual differences between adjusted CVs. Here we focus on bootstrapped CIs of individual CVs.

For cells within each individual cell line, we constructed 1000 samples with replacment and then computed 95% confidence interval. This method resulted in narrow confidence interval. 

Since sample size is a major factor in the width of confidence interval, instead of the entire set of cells for each individual, we sample a subset of the cells with replacement; the number of the sample is set to be 90, close to the number of wells on each C1 plate.


## Set up

```{r, message=FALSE, warning=FALSE}
library("data.table")
library("dplyr")
library("limma")
library("edgeR")
library("ggplot2")
library("grid")
library("Humanzee")
theme_set(theme_bw(base_size = 12))
source("functions.R")
```


## Prepare data

Input quality single cells

```{r import-quality-single-cell-list}
quality_single_cells <- read.table("../data/quality-single-cells.txt", 
                                  header = FALSE,
                                  stringsAsFactors = FALSE)
str(quality_single_cells)
```

Input annotation of only QC-filtered single cells. Remove NA19098.r2

```{r import-annotation-for-filtered-data}
anno_filter <- read.table("../data/annotation-filter.txt", 
                    header = TRUE,
                    stringsAsFactors = FALSE)
dim(anno_filter)
```

Import endogeneous gene molecule counts that are QC-filtered, CPM-normalized, ERCC-normalized, and also processed to remove unwanted variation from batch effet. ERCC genes are removed from this file.

```{r import-filtered-molecule-counts}
molecules_ENSG <- read.table("../data/molecules-final.txt", 
                             header = TRUE, stringsAsFactors = FALSE)
stopifnot(NCOL(molecules_ENSG) == NROW(anno_filter))
```


## Import gene symbols

```{r import-gene-symbols}
gene_info <- read.table("../data/gene-info.txt", sep = "\t",
                        header = TRUE, stringsAsFactors = FALSE)
str(gene_info)
```


## Compute normalized CV 

We compute squared CV across cells for each individual and then for each individual CV profile, account for mean dependency by computing distance with respect to the data-wide coefficient variation on the log10 scale. 

```{r compute-normalized-cv}
ENSG_cv <- Humanzee::compute_cv(log2counts = molecules_ENSG,
                      grouping_vector = anno_filter$individual)

ENSG_cv_adj <- Humanzee::normalize_cv(group_cv = ENSG_cv, 
                            log2counts = molecules_ENSG, 
                            anno = anno_filter)
```

## Compute summary measure of deviation 

* Standardize the each CV vectors

Individual CV vectors are standarized for individual CV mean and coefficients of variation across genes. 

```{r compute-summary-deviation}
df_cv <- data.frame(NA19098 = ENSG_cv_adj[[1]]$log10cv2_adj,
                    NA19101 = ENSG_cv_adj[[2]]$log10cv2_adj,
                    NA19239 = ENSG_cv_adj[[3]]$log10cv2_adj)

library(matrixStats)
df_norm <- sweep(df_cv, MARGIN = 2, STATS = colMeans(as.matrix(df_cv)), FUN = "-")
df_norm <- sweep(df_norm, MARGIN = 2, STATS = sqrt(colVars(as.matrix(df_cv))), FUN = "/")
colnames(df_norm) <- names(ENSG_cv_adj)
```

* SSM and SAM

Compute metrics for quantifying similarity between the three individual coefficients of variation. 

```{r}
library(matrixStats)

df_norm <- as.data.frame(df_norm)
df_norm$squared_dev <- rowSums( ( df_norm - rowMedians(as.matrix(df_norm)) )^2 )
df_norm$abs_dev <- rowSums(abs( df_norm - rowMedians(as.matrix(df_norm)) ))
```

## Bootstrapped CIs

### Sample size is the orignal sample size

```{r}
load("rda/cv-adjusted-statistical-test-bootstrap/bootstrap-ci.rda")
rownames(abs_dev_pnt) <- rownames(molecules_ENSG)
rownames(squared_dev_pnt) <- rownames(molecules_ENSG)

head(cbind(squared_dev_pnt, df_norm$squared_dev))
head(cbind(abs_dev_pnt, df_norm$abs_dev))
```

Genes with SSM outside of the 95% boostrapped CI are consistented to be statistical significant.

```{r}
sig_gene_CI_squared_dev <- df_norm$squared_dev > squared_dev_pnt[, 2] | df_norm$squared_dev < squared_dev_pnt[, 1]

sig_gene_CI_abs_dev <- df_norm$abs_dev > abs_dev_pnt[, 2] | df_norm$abs_dev < abs_dev_pnt[, 1]
```


```{r}
library(broman)
crayon <- brocolors("crayon")
ggplot(data.frame(index = c(mean(sig_gene_CI_squared_dev == 0), 
                            mean(sig_gene_CI_squared_dev == 1),
                            mean(sig_gene_CI_abs_dev == 0 ),
                            mean(sig_gene_CI_abs_dev == 1) ),
                  type = factor( rep(c(0,1), times = 2),
                                 labels = c("Non-sig.", "Sig.") ),
                  metric = rep( c("SSM", "SAM"), each = 2)  ),
      aes(x = metric, y = 100*index, fill = type) ) +
  geom_bar(stat = "identity", colour = "white") +
  scale_fill_manual(values = c("orange", "grey80")) +
  ylab("Percent") + xlab("Similarity metric") 
```


Correspondance of SSM and SAM results.

```{r}
library(gplots)
venn(list(SSM = rownames(molecules_ENSG)[which(sig_gene_CI_squared_dev)],
          SAM = rownames(molecules_ENSG)[which(sig_gene_CI_abs_dev)] ))
```


## Session information

```{r info}
sessionInfo()
```
