---
title: "igraphScatter instructions"
author: "Joyce Hsiao"
date: 2016-02-13
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objectives

We built a shiny app to interatively visualize each gene's count distribution across cells. The app is hosted in Humanzee package. 

This is our first attempt in building an interative plot for our single cell data. More exciting visualizations to come...


## Set up

```{r, eval = FALSE}
# Go singleCellSeq/project/analysis (default working directory)

# Install the most up-to-date version of Humanzee
require(devtools)
install_github("jhsiao999/Humanzee")
require(Humanzee)
```

## Prepare data

Load the molecule count data after filtering and before transformations.

```{r, eval = FALSE}
load("../data/igraphScatter-demo.rda")
demo_data <- demo_data(molecules_cv, anno_single, molecules_single_collision, 
                       per_person = "19098")
molecules_cv <- demo_data$molecules_cv
anno_single <- demo_data$anno_single
molecules_single_collision <- demo_data$molecules_single_collision
```


## Shiny!

Now the app is build for visualizing one individual at a time.

```{r, eval = F}
run_igraphScatter()
```



## Session information

```{r info}
sessionInfo()
```
