---
title: "Coefficient of variation"
author: "Joyce Hsiao"
date: 2015-09-26
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objectives

We used Kruskal-Wallis test to compare coefficient of variation between individual
(3 coefficient of variations from the batches for per individual). The p-value distribution of the per gene Kruskal-Wallis test does not resemeble p-value 
distribution of the commonly know per gene tests of differential gene expression. 

Visulization of the count distributions of the top 9 and bottom 9 genes that 
are different in coefficient of variation between individuals using the Kruskal-Wallis
test did not reveal systematic different between the count distributions.

We then applied Kwame Okrah's HTShape package to compute skewness and kurtoss estimtes
of each gene with each unique batches, and also across the three individuals. The
SO-plot shows that there are a good proportion of the genes that are skewed in their count distribution. 


## Set up

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
library(edgeR)
library(limma)
```


## Prepare data

Input annotation

```{r}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```


Input molecule counts

```{r}
molecules <- read.table("../data/molecules.txt", header = TRUE, stringsAsFactors = FALSE)
```

Remove bulk samples

```{r}
single_samples <- anno$well != "bulk"
anno_single <- anno[ which(single_samples), ]
molecules_single <- molecules[ , which(single_samples)]
stopifnot(ncol(molecules_single) == nrow(anno_single),
          colnames(molecules_single) == anno_single$sample_id)
```

Remove ERCC genes.

```{r}
nonERCC <- grep("ERCC", rownames(molecules_single), invert = TRUE)
molecules_single <- molecules_single[nonERCC, ]
```


Remove genes with no counts and also overly expressed genes.

```{r}
## remove gene with 0 counts
expressed_single <- rowSums(molecules_single) > 0
molecules_single <- molecules_single[expressed_single, ]
dim(molecules_single)

## remove gene with molecule count larger than 1024 (15 if them)
overexpressed_genes <- rownames(molecules_single)[apply(molecules_single, 1,
                                                        function(x) any(x >= 1024))]
molecules_single <- molecules_single[!(rownames(molecules_single) %in% overexpressed_genes), ]

## collision probability and cpm molecule counts
molecules_single_collision <- -1024 * log(1 - molecules_single / 1024)
molecules_single_cpm <- cpm(molecules_single_collision, log = TRUE)
```



## Compute coefficient of variation

Per individual coefficient of variation.

```{r, eval = FALSE}
molecules_cv <- 
  lapply(1:length(unique(anno_single$individual)), function(per_person) {
    molecules_per_person <- molecules_single_collision[ , anno_single$individual == unique(anno_single$individual)]
    cv_foo <- data.frame(mean = apply(molecules_per_person, 1, mean, na.rm = TRUE),
           cv = apply(molecules_per_person, 1, sd, na.rm = TRUE) / 
                apply(molecules_per_person, 1, mean, na.rm = TRUE) )
    rownames(cv_foo) <- rownames(molecules_single_collision)
    
    cv_foo <- cv_foo[rowSums(is.na(cv_foo)) == 0, ]
    cv_foo$individual <- unique(anno_single$individual)[per_person]
      
    return(cv_foo)
    }) 
names(molecules_cv) <- unique(anno_single$individual)
```


Per batch coefficient of variation.

```{r}
unique_batch <- with(anno_single, paste(individual, batch, sep = "_"))

molecules_cv_batch <- 
  lapply(1:length(unique(unique_batch)), function(per_batch) {
    molecules_per_batch <- molecules_single_collision[ , unique_batch == unique(unique_batch)[per_batch]]
    cv_foo <- data.frame(mean = apply(molecules_per_batch, 1, mean, na.rm = TRUE),
           cv = apply(molecules_per_batch, 1, sd, na.rm = TRUE) / 
                apply(molecules_per_batch, 1, mean, na.rm = TRUE) )
    rownames(cv_foo) <- rownames(molecules_single_collision)
    
    cv_foo <- cv_foo[rowSums(is.na(cv_foo)) == 0, ]
    cv_foo$batch <- unique(unique_batch)[per_batch]
      
    return(cv_foo)
    }) 
names(molecules_cv_batch) <- unique(unique_batch)
```


Subset CV data to include only genes that have data across all batches.

```{r}
all_genes <- lapply(1:3, function(ii) {
                intersect(intersect(rownames(molecules_cv_batch[[ 1+ (ii-1)*3]]),
                                    rownames(molecules_cv_batch[[ 2+ (ii-1)*3]]) ),
                          rownames(molecules_cv_batch[[3 + (ii-1)*3]]) ) })
all_genes <- intersect(intersect(all_genes[[1]], all_genes[[2]]), all_genes[[3]])
str(all_genes)

molecules_cv_batch <- 
  lapply(1: length(molecules_cv_batch), function(per_batch) {
          ii_include <- which(rownames(molecules_cv_batch[[per_batch]]) %in% all_genes)
          molecules_cv_batch[[per_batch]][ ii_include, ]
  })

stopifnot(all.equal(rownames(molecules_cv_batch[[9]]), rownames(molecules_cv_batch[[8]]) ) )

nrow(molecules_cv_batch[[1]])
```


## Kruskal-Wallis test

A nonparametric test commonly used for comparison of multiple groups that have 
a very small sample size.

```{r}
df_cv <- lapply(1:nrow(molecules_cv_batch[[1]]), function(per_gene) {
  do.call(rbind, lapply(1: length(molecules_cv_batch), function(per_batch) {
                    molecules_cv_batch[[per_batch]][per_gene, ] })
                  )
})

# Create a factor variable of individual
individual <- rep(unique(anno_single$individual), times = 3)

# Kruskal walls
kw_pval <- lapply(1:nrow(molecules_cv_batch[[1]]), function(per_gene) {
  result <- kruskal.test(cv ~ factor(individual), data = df_cv[[per_gene]])
  data.frame(pval = result$p.value, 
             gene = rownames(molecules_cv_batch[[1]])[per_gene])
})
kw_pval <- do.call(rbind, kw_pval)


summary(kw_pval$pval)
ggplot(data = kw_pval, aes(x = pval)) + geom_histogram() +
  ggtitle("Kruskal-Wallis p-value")

# library(ggplot2)
# ggplot(data = molecules_cv[[1]], aes(x = mean, y = cv)) +
#   geom_point(size = 2, alpha = .5) + scale_x_log10() +
#   geom_point(data = data.frame(mean = molecules_cv[[1]]$mean,
#                                cv_adj = cv_adj), aes(x = mean, y = cv_adj), 
#              size = 2, alpha = .5, col = "blue")
```


## Top 9 in Kruskal-Wallis

```{r}
ii_diff <- order(kw_pval$pval)[1:9]
gene_subset <- df_cv[ii_diff]

df_cv_merge <- lapply(1:9, function(per_gene) {
          data.frame(gene_subset[[per_gene]], 
                     individual = rep(c("19098", "19101", "19239"), 3),
                     gene = kw_pval$gene[ii_diff[per_gene]])
})
df_cv_merge <- do.call(rbind, df_cv_merge)

ggplot(df_cv_merge, aes( x = factor(individual), y = cv)) +
  geom_point(aes(col = factor(individual))) + facet_grid(. ~ gene)
```


```{r}
unique_batch <- with(anno_single, paste(individual, batch, sep = "_"))
count_cell <- lapply(1:9, function(per_batch) {
    ii_batch <- which(unique_batch %in% unique(unique_batch)[per_batch])
    ii_gene <- which( rownames(molecules_single_collision) %in% kw_pval$gene[ii_diff] )
    df_foo <- molecules_single_collision[ii_gene, ii_batch]
    colnames(df_foo) <- ""

    df_meta <- data.frame(df_foo, 
                         individual = anno_single$individual[ii_batch][1],
                         batch = anno_single$batch[ii_batch][1],
                         genes = rownames(df_foo))
    df_meta
})
names(count_cell) <- unique(unique_batch)

count_cell_merge <- do.call(rbind, count_cell)

table(count_cell_merge$individual)

count_cell_merge <- reshape(count_cell_merge, direction = "long",
                            timevar = "gene", varying = list(1:96) )

ggplot(count_cell_merge, 
        aes( x = Var.1, group = factor(paste(individual, batch, "_"))) ) +
        geom_density(aes(x = Var.1, col = factor(individual)) ) + 
        facet_wrap( ~ genes, scales = "free")
```


## Bottom 9 in Kruskal-Wallis

```{r}
ii_diff <- order(kw_pval$pval, decreasing = TRUE)[1:9]
gene_subset <- df_cv[ii_diff]

df_cv_merge <- lapply(1:9, function(per_gene) {
          data.frame(gene_subset[[per_gene]], 
                     individual = rep(c("19098", "19101", "19239"), 3),
                     gene = kw_pval$gene[ii_diff[per_gene]])
})
df_cv_merge <- do.call(rbind, df_cv_merge)

ggplot(df_cv_merge, aes( x = factor(individual), y = cv)) +
  geom_point(aes(col = factor(individual))) + facet_grid(. ~ gene)
```


```{r}
unique_batch <- with(anno_single, paste(individual, batch, sep = "_"))
count_cell <- lapply(1:9, function(per_batch) {
    ii_batch <- which(unique_batch %in% unique(unique_batch)[per_batch])
    ii_gene <- which( rownames(molecules_single_collision) %in% kw_pval$gene[ii_diff] )
    df_foo <- molecules_single_collision[ii_gene, ii_batch]
    colnames(df_foo) <- ""

    df_meta <- data.frame(df_foo, 
                         individual = anno_single$individual[ii_batch][1],
                         batch = anno_single$batch[ii_batch][1],
                         genes = rownames(df_foo))
    df_meta
})
names(count_cell) <- unique(unique_batch)

count_cell_merge <- do.call(rbind, count_cell)

count_cell_merge <- reshape(count_cell_merge, direction = "long",
                            timevar = "gene", varying = list(1:96) )

ggplot(count_cell_merge, 
        aes( x = Var.1, group = factor(paste(individual, batch, "_"))) ) +
        geom_density(aes(x = Var.1, col = factor(individual)) ) + 
        facet_wrap( ~ genes, scales = "free")
```



## Explore quantiles

Use Kwame's package to compare shapes of distributions.

```{r}
require(devtools)
install_github("kokrah/HTshape")
require(HTShape)
vignette("HTShape")
```

Select genes that have valid CV measures in all batches. 

SO-plot for one batch...

```{r}
ii_genes <- which(rownames(molecules_single_collision) %in% all_genes )
unique_batch <- with(anno_single, paste(individual, batch, sep = "_"))

fit_shape <- lapply(1:9, function(per_batch) {
                    ii_batch <- which(unique_batch == unique(unique_batch)[per_batch])
                    count_foo <- as.matrix(molecules_single_collision[ii_genes, ii_batch])
                    shape_results <- fitShape( log(count_foo + 1) )
                    list(L3 = shape_results$lrats[ "LR3", ],
                         L4 = shape_results$lrats[ "LR4", ] )
                })
plotSO(fit_shape[[1]]$L3, fit_shape[[1]]$L4)
```


SO-plot for the entire data set.

```{r}
ii_genes <- which(rownames(molecules_single_collision) %in% all_genes )

fit_shape <- fitShape( log(molecules_single_collision[ii_genes, ] + 1) )

plotSO(fit_shape$lrats[1, ], fit_shape$lrats[2, ])
```




## Session information

```{r info}
sessionInfo()
```
