---
title: "CV and sparsity"
author: "Joyce Hsiao"
date: 2015-10-07
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Objective

Explore the relationship between coefficient of variation and sparsity.

As expected, a large coefficient of variation is associated with a large number of zero-count cells in a gene, an observation that is consistent with an important feature of the count data where the coefficient of variation is large for genes with low expression levels.

## Set up

```{r, message=FALSE, warning=FALSE}
library("data.table")
library("dplyr")
library("limma")
library("edgeR")
library("ggplot2")
library("grid")
theme_set(theme_bw(base_size = 12))
source("functions.R")

## Knitr settings
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE)
```



## Prepare data


Input annotation of only QC-filtered single cells

```{r}
anno_qc <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_qc)
```


Input molecule counts that are filtered, transformed, and also processed to remove unwanted variation from batch effet. ERCC genes are also removed.

```{r}
molecules_qc <- read.table("../data/molecules-final.txt", header = TRUE, stringsAsFactors = FALSE)
```


## Compute coefficient of variation

Compute per batch coefficient of variation.

Include only genes with positive coefficient of variation. Some genes in this data may have zero coefficient of variation, because we include gene with more than 0 count 
across all cells.


```{r}
molecules_cv_batch <- 
  lapply(1:length(unique(anno_qc$batch)), function(per_batch) {
        molecules_per_batch <- molecules_qc[ , unique(anno_qc$batch) == unique(anno_qc$batch)[per_batch] ]
        mean_per_gene <- apply(molecules_per_batch, 1, mean, na.rm = TRUE)
        sd_per_gene <- apply(molecules_per_batch, 1, sd, na.rm = TRUE)
        cv_per_gene <- data.frame(mean = mean_per_gene,
                                  sd = sd_per_gene,
                                  cv = sd_per_gene/mean_per_gene)
        rownames(cv_per_gene) <- rownames(molecules_qc)
    
        cv_per_gene <- cv_per_gene[rowSums(is.na(cv_per_gene)) == 0, ]
        cv_per_gene$batch <- unique(anno_qc$batch)[per_batch]
        return(cv_per_gene)
      }) 
names(molecules_cv_batch) <- unique(anno_qc$batch)

dim(molecules_cv_batch[[1]])
```


## Sparsity

Import molecule count file before normlization.

```{r}
molecules_filter <- read.table("../data/molecules-filter.txt")

select_nonERCC <- grep("ERCC", rownames(molecules_filter), invert = TRUE)
molecules_ENSG <- molecules_filter[select_nonERCC, ]

stopifnot(dim(molecules_ENSG) == dim(molecules_qc))
```

Compute the proportion of zero count cells in each gene. 

per_batch = 2
```{r}
molecules_sparse_batch <- 
  lapply(1:length(unique(anno_qc$batch)), function(per_batch) {
        sparse_per_batch <- molecules_ENSG[ , unique(anno_qc$batch) == unique(anno_qc$batch)[per_batch] ]
        sparse_per_gene <- apply(sparse_per_batch, 1, function(x) {
                                  mean(x == 0)
                                })
        return(sparse_per_gene)
      }) 
names(molecules_sparse_batch) <- unique(anno_qc$batch)
str(molecules_sparse_batch[[1]])
str(molecules_sparse_batch[[2]])
```

Frequency of zero-count cell proportions.

```{r}
par(mfrow = c(3,3), mar = c(4,2,2,3))
for (per_batch in 1:length(unique(anno_qc$batch))) {
  hist( molecules_sparse_batch[[per_batch]],
        main = unique(anno_qc$batch)[per_batch], breaks = 100,
        ylim = c(0, 3000), xlim = c(0, 1),
        xlab = "No. of zero-count cells")
}
```

## CV and sparsity

```{r}
require(broman)
require(scales)
crayons <- brocolors("crayon")
par(mfrow = c(3,3), mar = c(4,2,2,3))
for (per_batch in 1:length(unique(anno_qc$batch))) {
  df_xy <- data.frame(x = molecules_sparse_batch[[per_batch]],
                      y = molecules_cv_batch[[per_batch]]$cv)
  plot(df_xy,
       xlab = "Proportion of zero-count cells",
       ylab = "CV",
       xlim = c(0, 1), ylim = c(0, 1),
       main = unique(anno_qc$batch)[per_batch], 
       cex = .6, pch = 16, col = alpha(crayons["Outer Space"], .8))
  loess_fit <- loess(y ~ x, df_xy)
  x_seq <- with(df_xy, seq(min(x), max(x), by = .01))
  lines(x_seq, 
        predict(loess_fit, x_seq), 
        col = crayons["Outrageous Orange"],
        lwd = 1)
}
```





## Session information

```{r info}
sessionInfo()
```
