---
title: "Data transformation"
date: 2015-09-30
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(fig.width = 8, fig.height = 8)
```

```{r packages, message=FALSE}
library("ggplot2")
library("cowplot")
theme_set(theme_bw(base_size = 16))
theme_update(panel.grid.minor.x = element_blank(),
             panel.grid.minor.y = element_blank(),
             panel.grid.major.x = element_blank(),
             panel.grid.major.y = element_blank())
source("functions.R")
```

This file performs principal components analysis (PCA) and displays the results for the data at each stage of our data transformation pipeline.
Furthermore, it quantifies and tests the PCA results using [pca-utils][] (commit [8540ff2][]).
The methodology is described in [Worley et al., 2013][worley2013].
Briefly, the differences between samples are calculated using the squared Mahalanobis distance.
The distances are then scaled so that the hypothesis test that the samples from two groups come from separate distributions can be computed with an F-test.

[pca-utils]: https://github.com/geekysuavo/pca-utils
[8540ff2]: https://github.com/geekysuavo/pca-utils/tree/8540ff2c6769031b0505af78f2dcb5703ae09294
[worley2013]: http://www.ncbi.nlm.nih.gov/pubmed/23079505

## Input

Input filtered annotation.

```{r input-annotation-filter}
anno_filter <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_filter)
```

Input filtered molecule counts.

```{r input-molecule-counts-filter}
molecules_filter <- read.table("../data/molecules-filter.txt", header = TRUE,
                               stringsAsFactors = FALSE)
stopifnot(ncol(molecules_filter) == nrow(anno_filter),
          colnames(molecules_filter) == anno_filter$sample_id)
```

Input standardized molecule counts.

```{r input-molecule-counts-cpm}
molecules_cpm <- read.table("../data/molecules-cpm.txt", header = TRUE,
                            stringsAsFactors = FALSE)
stopifnot(ncol(molecules_cpm) == nrow(anno_filter),
          colnames(molecules_cpm) == anno_filter$sample_id)
```

Input Poisson GLM transformed molecule counts per million.

```{r input-molecule-counts-trans}
molecules_cpm_trans <- read.table("../data/molecules-cpm-trans.txt", header = TRUE,
                               stringsAsFactors = FALSE)
stopifnot(ncol(molecules_cpm_trans) == nrow(anno_filter),
          colnames(molecules_cpm_trans) == anno_filter$sample_id)
```

Input final batch-corrected molecule counts per million.

```{r input-molecule-counts-final}
molecules_final <- read.table("../data/molecules-final.txt", header = TRUE,
                              stringsAsFactors = FALSE)
stopifnot(ncol(molecules_final) == nrow(anno_filter),
          colnames(molecules_final) == anno_filter$sample_id)
```

## PCA

### PCA of filtered data

```{r pca-molecules-filter}
pca_molecules_filter <- run_pca(molecules_filter)
pca_molecules_filter_plot <- plot_pca(pca_molecules_filter$PCs,
         explained = pca_molecules_filter$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate", alpha = 0.5, size = 2.2) +
  labs(title = "Counts")
```

```{r pca-molecules-filter-out}
pca_molecules_filter_out <- data.frame(Obs.id = 1:nrow(anno_filter),
                                       Obs.batch = anno_filter$batch,
                                       pca_molecules_filter$PCs[, 1:2])
write.table(pca_molecules_filter_out,
            file = "../data/pca-molecules-filter.txt",
            quote = FALSE, sep = "\t", row.names = FALSE)
```

```{r pca-molecules-filter-dist, engine='bash'}
pca-distances -i ../data/pca-molecules-filter.txt
```

```{r pca-molecules-filter-dendro, engine='bash'}
pca-dendrogram -i ../data/pca-molecules-filter.txt
```

### PCA of standardized data

```{r pca-molecules-cpm}
pca_molecules_cpm <- run_pca(molecules_cpm)
pca_molecules_cpm_plot <- plot_pca(pca_molecules_cpm$PCs,
         explained = pca_molecules_cpm$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate", alpha = 0.5, size = 2.2) +
  labs(title = "Counts per million (log)")
```

```{r pca-molecules-cpm-out}
pca_molecules_cpm_out <- data.frame(Obs.id = 1:nrow(anno_filter),
                                       Obs.batch = anno_filter$batch,
                                       pca_molecules_cpm$PCs[, 1:2])
write.table(pca_molecules_cpm_out,
            file = "../data/pca-molecules-cpm.txt",
            quote = FALSE, sep = "\t", row.names = FALSE)
```

```{r pca-molecules-cpm-dist, engine='bash'}
pca-distances -i ../data/pca-molecules-cpm.txt
```

```{r pca-molecules-cpm-dendro, engine='bash'}
pca-dendrogram -i ../data/pca-molecules-cpm.txt
```

### PCA of Poisson GLM transformed molecule counts per million

```{r pca-molecules-cpm-trans}
pca_molecules_cpm_trans <- run_pca(molecules_cpm_trans)
pca_molecules_cpm_trans_plot <- plot_pca(pca_molecules_cpm_trans$PCs,
         explained = pca_molecules_cpm_trans$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate",  alpha = 0.5, size = 2.2) +
  labs(title = "Poisson transformation")
```

```{r pca-molecules-cpm-trans-out}
pca_molecules_cpm_trans_out <- data.frame(Obs.id = 1:nrow(anno_filter),
                                       Obs.batch = anno_filter$batch,
                                       pca_molecules_cpm_trans$PCs[, 1:2])
write.table(pca_molecules_cpm_trans_out,
            file = "../data/pca-molecules-cpm-trans.txt",
            quote = FALSE, sep = "\t", row.names = FALSE)
```

```{r pca-molecules-cpm-trans-dist, engine='bash'}
pca-distances -i ../data/pca-molecules-cpm-trans.txt
```

```{r pca-molecules-cpm-trans-dendro, engine='bash'}
pca-dendrogram -i ../data/pca-molecules-cpm-trans.txt
```

### PCA of final batch-corrected data

```{r pca-molecules-final}
pca_final <- run_pca(molecules_final)
pca_final_plot <- plot_pca(pca_final$PCs, explained = pca_final$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate", alpha = 0.5, size = 2.2) +
  labs(title = "Batch corrected")
```

```{r pca-final-out}
pca_final_out <- data.frame(Obs.id = 1:nrow(anno_filter),
                            Obs.batch = anno_filter$batch,
                            pca_final$PCs[, 1:2])
write.table(pca_final_out,
            file = "../data/pca-final.txt",
            quote = FALSE, sep = "\t", row.names = FALSE)
```

```{r pca-molecules-final-dist, engine='bash'}
pca-distances -i ../data/pca-final.txt
```

```{r pca-molecules-final-dendro, engine='bash'}
pca-dendrogram -i ../data/pca-final.txt
```

## Data transformation figure

```{r data-transformation, fig.width=12, fig.height=12}
plot_grid(pca_molecules_filter_plot +
            guides(shape = FALSE) + theme(legend.position = "bottom"),
          pca_molecules_cpm_plot +
            guides(color = FALSE) + theme(legend.position = "bottom"),
          pca_molecules_cpm_trans_plot +
            guides(shape = FALSE) + theme(legend.position = "bottom"),
          pca_final_plot +
            guides(color = FALSE) + theme(legend.position = "bottom"),
          labels = LETTERS[1:4])
```

## Session information

```{r info}
sessionInfo()
```
