---
title: "Data transformation"
date: 2015-09-30
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(fig.width = 8, fig.height = 8)
```

```{r packages, message=FALSE}
library("ggplot2")
library("cowplot")
theme_set(theme_bw(base_size = 12))
source("functions.R")
```

## Input

Input filtered annotation.

```{r input-annotation-filter}
anno_filter <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_filter)
```

Input filtered molecule counts.

```{r input-molecule-counts-filter}
molecules_filter <- read.table("../data/molecules-filter.txt", header = TRUE,
                               stringsAsFactors = FALSE)
stopifnot(ncol(molecules_filter) == nrow(anno_filter),
          colnames(molecules_filter) == anno_filter$sample_id)
```

Input standardized molecule counts.

```{r input-molecule-counts-cpm}
molecules_cpm <- read.table("../data/molecules-cpm.txt", header = TRUE,
                            stringsAsFactors = FALSE)
stopifnot(ncol(molecules_cpm) == nrow(anno_filter),
          colnames(molecules_cpm) == anno_filter$sample_id)
```

Input Poisson GLM transformed molecule counts per million.

```{r input-molecule-counts-trans}
molecules_cpm_trans <- read.table("../data/molecules-cpm-trans.txt", header = TRUE,
                               stringsAsFactors = FALSE)
stopifnot(ncol(molecules_cpm_trans) == nrow(anno_filter),
          colnames(molecules_cpm_trans) == anno_filter$sample_id)
```

Input final batch-corrected molecule counts per million.

```{r input-molecule-counts-final}
molecules_final <- read.table("../data/molecules-final.txt", header = TRUE,
                              stringsAsFactors = FALSE)
stopifnot(ncol(molecules_final) == nrow(anno_filter),
          colnames(molecules_final) == anno_filter$sample_id)
```

## PCA

### PCA of filtered data

```{r pca-molecules-filter}
pca_molecules_filter <- run_pca(molecules_filter)
pca_molecules_filter_plot <- plot_pca(pca_molecules_filter$PCs,
         explained = pca_molecules_filter$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate", alpha = 0.5, size = 1.5) +
  labs(title = "Counts")
```

### PCA of standardized data

```{r pca-molecules-cpm}
pca_molecules_cpm <- run_pca(molecules_cpm)
pca_molecules_cpm_plot <- plot_pca(pca_molecules_cpm$PCs,
         explained = pca_molecules_cpm$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate", alpha = 0.5, size = 1.5) +
  labs(title = "Counts per million (log))")
```

### PCA of Poisson GLM transformed molecule counts per million

```{r pca-molecules-cpm-trans}
pca_molecules_cpm_trans <- run_pca(molecules_cpm_trans)
pca_molecules_cpm_trans_plot <- plot_pca(pca_molecules_cpm_trans$PCs,
         explained = pca_molecules_cpm_trans$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate",  alpha = 0.5, size = 1.5) +
  labs(title = "Poisson transformation")
```

### PCA of final batch-corrected data

```{r pca-molecules-final}
pca_final <- run_pca(molecules_final)
pca_final_plot <- plot_pca(pca_final$PCs, explained = pca_final$explained,
         metadata = anno_filter, color = "individual",
         shape = "replicate", alpha = 0.5, size = 1.5) +
  labs(title = "Batch corrected")
```

## Data transformation figure

```{r data-transformation, fig.width=8, fig.height=8}
plot_grid(pca_molecules_filter_plot + theme(legend.position = "none"),
          pca_molecules_cpm_plot + theme(legend.position = "none"),
          pca_molecules_cpm_trans_plot + theme(legend.position = "none"),
          pca_final_plot + theme(legend.position = "none"),
          labels = LETTERS[1:4])
```

## Session information

```{r info}
sessionInfo()
```
