---
title: "Draft draft draft"
author: "Po-Yuan Tung"
date: 2015-09-21
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

A draft for a fabulous paper.

## Introduction

* challenges of single cell RNA-seq
* molecule and read:UMI
* normalization and techinical noise:ERCC
* sample size: number of reads, number of cells
* novelty of this study: multiple c1 bathes per individual and multiple individuals

## M and M
1. cell culture (charaterization?)
2. single cell collection, library prep, sequencing
3. mapping, featurecounts, umi tool
4. linear shift and input counts of each ERCC gene 
5. mix effect model

## Figures
1. study design and QC
2. standardization: cmp ERCC genes and endogenous genes separately
3. subsampling: a. gene expression b. number of genes c. variance
4. normalization: pca plots showing before and after linear shift and mix effect model
5. biological findings: differential CV genes

## Sub Figures
1. data collection: sequencing 
2. reads to molecule: individual effect and batch effect
3. noisy genes: based on non standardized counts
4. ERCC not good: RUV and BASiCs
5. cell state: cell cycle and pluripotency
6. OE genes: before and after removal 

## Session information

```{r info}
sessionInfo()
```
