---
title: "ERCC counts"
author: "PoYuan Tung"
date: 2015-07-07
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


## Input

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input ERCC concentration information.

```{r input-ercc}
ercc <- read.table("../data/ercc-info.txt", header = TRUE, sep = "\t",
                   stringsAsFactors = FALSE)
colnames(ercc) <- c("num", "id", "subgroup", "conc_mix1", "conc_mix2",
                    "expected_fc", "log2_mix1_mix2")
head(ercc)
stopifnot(nrow(ercc) == 92)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input list of quality single cells.

```{r input-quality-single-cells}
quality_single_cells <- scan("../data/quality-single-cells.txt",
                             what = "character")
```

## Prepare single cell molecule data

Keep only the single cells that passed the [QC filters](qc-cell-ipsc.html).
This also removes the bulk samples.

```{r qc-filter}
molecules_single <- molecules[, colnames(molecules) %in% quality_single_cells]
anno_single <- anno[anno$sample_id %in% quality_single_cells, ]
stopifnot(ncol(molecules_single) == nrow(anno_single),
          colnames(molecules_single) == anno_single$sample_id)
```

Also remove batch 2 of individual 19098.

```{r remove-19098-batch-2-single}
molecules_single <- molecules_single[, !(anno_single$individual == 19098 & anno_single$batch == 2)]
anno_single <- anno_single[!(anno_single$individual == 19098 & anno_single$batch == 2), ]
stopifnot(ncol(molecules_single) == nrow(anno_single))
```

Remove genes with zero read counts in the single cells.

```{r remove-non-expressed-genes}
expressed_single <- rowSums(molecules_single) > 0
molecules_single <- molecules_single[expressed_single, ]
dim(molecules_single)
```

How many genes have greater than or equal to 1,024 molecules in at least one of the cells?

```{r over-expressed}
overexpressed_genes <- rownames(molecules_single)[apply(molecules_single, 1,
                                                        function(x) any(x >= 1024))]
```

`r length(overexpressed_genes)` have greater than or equal to 1,024 molecules.
Remove them.

```{r remove-over-expressed}
molecules_single <- molecules_single[!(rownames(molecules_single) %in% overexpressed_genes), ]
```

Correct for the collision probability.
See [Grun et al. 2014][Grun2014] for details.

```{r collision-probability}
molecules_single_collision <- -1024 * log(1 - molecules_single / 1024)
```

[Grun2014]: http://www.nature.com/nmeth/journal/v11/n6/full/nmeth.2930.html#methods

## Calculate the molecule counts of each ERCC in single cell
Accorindg to Pollen et al.[Pollen2014], there are total 28,000 ERCC molecule per cell when the ERCC stock is used in 1:20,000 dilution. In our study, I used 1:50,000 dilution. Therefore, we have total 28000/2.5= 11200 ERCC molecule per cell.

[Pollen2014]: http://www.ncbi.nlm.nih.gov/pmc/articles/PMC4191988/

```{r ercc-number}
#### sort ercc file by gene ID
ercc <- ercc[order(ercc$id), ]

#### number of genes detected in all single cells
ercc_rows_molecules_single <- grep("ERCC", rownames(molecules_single))
length(ercc_rows_molecules_single)

#### keep only the ercc detected in single cells
ercc_detected <- ercc[ercc$id %in% rownames(molecules_single), ]

#### calculate the number of each ERCC based on the genes detected
ratio <- 11200/ sum(ercc_detected[,"conc_mix1"])
ratio
ercc_detected$counts <- ercc_detected[,"conc_mix1"] * ratio
head(ercc_detected)

#### try 28000 to see if integers
ratio_t <- 28000/ sum(ercc_detected[,"conc_mix1"])
ercc_detected$counts_t <- ercc_detected[,"conc_mix1"] * ratio_t
head(ercc_detected)
```



## Session information

```{r info}
sessionInfo()
```
