---
title: "ERCC normalization"
date: 2015-06-15
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

## Setup

```{r packages, message=FALSE}
library("ggplot2")
theme_set(theme_bw(base_size = 16))
library("RUVSeq")
source("functions.R")
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input ERCC concentration information.

```{r input-ercc}
ercc <- read.table("../data/ercc-info.txt", header = TRUE, sep = "\t",
                   stringsAsFactors = FALSE)
colnames(ercc) <- c("num", "id", "subgroup", "conc_mix1", "conc_mix2",
                    "expected_fc", "log2_mix1_mix2")
head(ercc)
stopifnot(nrow(ercc) == 92)
```

Input list of quality single cells.

```{r input-quality-single-cells}
quality_single_cells <- scan("../data/quality-single-cells.txt",
                             what = "character")
```

Keep only the single cells that passed the [QC filters](qc-cell-ipsc.html) and the bulk samples.

```{r qc-filter}
molecules <- molecules[, grepl("bulk", colnames(molecules)) |
                         colnames(molecules) %in% quality_single_cells]
anno <- anno[anno$well == "bulk" | anno$sample_id %in% quality_single_cells, ]
stopifnot(ncol(molecules) == nrow(anno),
          colnames(molecules) == anno$sample_id)
```

Remove genes with zero read counts in the single cells or bulk samples.

```{r remove-non-expressed-genes}
expressed <- rowSums(molecules[, anno$well == "bulk"]) > 0 &
             rowSums(molecules[, anno$well != "bulk"]) > 0
molecules <- molecules[expressed, ]
```

## Correlation with ERCC

How many of the 92 ERCC spike-ins had at least one read sequenced in at least one of the samples?

```{r num-ercc-present}
sum(grepl("ERCC", rownames(molecules)))
```

```{r}
ercc_counts <- molecules[grep("ERCC", rownames(molecules)), ]
```

Sort both the concentration and the counts by the ERCC ID.

```{r}
ercc_counts <- ercc_counts[order(rownames(ercc_counts)), ]
ercc <- ercc[order(ercc$id), ]
# Also remove spike-ins with no counts
ercc <- ercc[ercc$id %in% rownames(ercc_counts), ]
stopifnot(rownames(ercc_counts) == ercc$id)
```

What is the correlation of the mean molecules counts to the expected concentrations?

```{r}
cor(rowMeans(ercc_counts), ercc$conc_mix1)
plot(rowMeans(ercc_counts), ercc$conc_mix1)
ercc_fit <- lm(ercc$conc_mix1 ~ rowMeans(ercc_counts))
abline(ercc_fit, col = "red")
title(sprintf("Y ~ %.2fX + %.2f \n R-squared: %.2f", ercc_fit$coefficients[2],
              ercc_fit$coefficients[1], summary(ercc_fit)$r.squared))
```

What is the correlation of the mean molecules counts to the expected concentrations using only the single cells?

```{r}
cor(rowMeans(ercc_counts[, grep("bulk", colnames(ercc_counts), invert = TRUE)]),
    ercc$conc_mix1)
```

What is the correlation of the mean molecules counts to the expected concentrations using only the bulk samples?

```{r}
cor(rowMeans(ercc_counts[, grep("bulk", colnames(ercc_counts))]),
    ercc$conc_mix1)
```

How much variation is there in the correlation between the samples?

```{r}
ercc_per_sample <- matrix(nrow = ncol(ercc_counts), ncol = 6)
colnames(ercc_per_sample) <- c("individual", "batch", "well",
                               "intercept", "slope", "r2")
for (i in 1:ncol(ercc_counts)) {
  fit <- lm(ercc$conc_mix1 ~ ercc_counts[, i])
  
  ercc_per_sample[i, ] <- c(anno$individual[i], anno$batch[i], anno$well[i],
                            fit$coefficients[1], fit$coefficients[2],
                            summary(fit)$r.squared)
}
ercc_per_sample <- as.data.frame(ercc_per_sample, stringsAsFactors = FALSE)
ercc_per_sample$intercept <- as.numeric(ercc_per_sample$intercept)
ercc_per_sample$slope <- as.numeric(ercc_per_sample$slope)
ercc_per_sample$r2 <- as.numeric(ercc_per_sample$r2)
str(ercc_per_sample)
```

```{r}
boxplot(intercept ~ individual + batch, data = ercc_per_sample)
```

```{r}
boxplot(slope ~ individual + batch, data = ercc_per_sample)
```


```{r}
boxplot(r2 ~ individual + batch, data = ercc_per_sample)
```

## Linear shift normalization

The normalization will be performed separately for bulk and single cells.

```{r}
bulk_raw <- molecules[grep("ERCC", rownames(molecules), invert = TRUE),
                      grep("bulk", colnames(molecules))]
bulk_ercc <- ercc_counts[, grep("bulk", colnames(molecules))]
single_raw <- molecules[grep("ERCC", rownames(molecules), invert = TRUE),
                        grep("bulk", colnames(molecules), invert = TRUE)]
single_ercc <- ercc_counts[, grep("bulk", colnames(molecules), invert = TRUE)]
stopifnot(rownames(bulk_raw) == rownames(single_raw),
          rownames(bulk_ercc) == rownames(single_ercc),
          colnames(bulk_raw) == colnames(bulk_ercc),
          colnames(single_raw) == colnames(single_ercc))
```

Fit a linear model using the mean ERCC counts.

```{r}
bulk_fit <- lm(ercc$conc_mix1 ~ rowMeans(bulk_ercc))
plot(rowMeans(bulk_ercc), ercc$conc_mix1, main = "bulk")
abline(bulk_fit, col = "red")
plot(fitted(bulk_fit), residuals(bulk_fit))
```

```{r}
single_fit <- lm(ercc$conc_mix1 ~ rowMeans(single_ercc))
plot(rowMeans(single_ercc), ercc$conc_mix1, main = "single")
abline(single_fit, col = "red")
plot(fitted(single_fit), residuals(single_fit))
```

Adjust each individual sample based on the mean fit.

```{r bulk-adjust}
bulk_norm <- bulk_raw
bulk_norm[, ] <- NA
for (i in 1:ncol(bulk_norm)) {
#   bulk_norm[, i] <- predict(bulk_fit, as.data.frame(bulk_raw[, i]),
#                             interval = "prediction")
  bulk_norm[, i] <- bulk_raw[, i] * bulk_fit$coefficients[2] +
                    bulk_fit$coefficients[1]
}
```

```{r}
heatmap(cor(bulk_raw))
heatmap(cor(bulk_norm))
```


```{r single-adjust}
single_norm <- single_raw
single_norm[, ] <- NA
for (i in 1:ncol(single_norm)) {
#   single_norm[, i] <- predict(single_fit, as.data.frame(single_raw[, i]),
#                             interval = "prediction")
  single_norm[, i] <- single_raw[, i] * single_fit$coefficients[2] +
                    single_fit$coefficients[1]
}
```

```{r}
heatmap(cor(single_raw))
heatmap(cor(single_norm))
```


## RUVseq normalization

```{r bulk-ruv}
bulk_ruv_object <- RUVg(x = as.matrix(rbind(bulk_ercc, bulk_raw)),
                        cIdx = 1:nrow(bulk_ercc), k = 1)
bulk_ruv <- bulk_ruv_object$normalizedCounts[
  grep("ERCC", rownames(bulk_ruv_object$normalizedCounts), invert = TRUE), ]
```

```{r single-ruv}
single_ruv_object <- RUVg(x = as.matrix(rbind(single_ercc, single_raw)),
                        cIdx = 1:nrow(single_ercc), k = 1)
single_ruv <- single_ruv_object$normalizedCounts[
  grep("ERCC", rownames(single_ruv_object$normalizedCounts), invert = TRUE), ]
```

## PCA comparison

With the bulk data:

```{r}
pca_bulk_raw <- run_pca(bulk_raw)
plot_pca(pca_bulk_raw$PCs, color = anno$individual[anno$well == "bulk"])
```


```{r}
pca_bulk_norm <- run_pca(bulk_norm)
plot_pca(pca_bulk_norm$PCs, color = anno$individual[anno$well == "bulk"])
```

```{r}
pca_bulk_ruv <- run_pca(bulk_ruv[apply(bulk_ruv, 1, var) > 0, ])
plot_pca(pca_bulk_ruv$PCs, color = anno$individual[anno$well == "bulk"])
```

With the single cell data:


```{r}
pca_single_raw <- run_pca(single_raw)
plot_pca(pca_single_raw$PCs, color = anno$individual[anno$well != "bulk"])
```


```{r}
pca_single_norm <- run_pca(single_norm)
plot_pca(pca_single_norm$PCs, color = anno$individual[anno$well != "bulk"])
```

```{r}
pca_single_ruv <- run_pca(single_ruv[apply(single_ruv, 1, var) > 0, ])
plot_pca(pca_single_ruv$PCs, color = anno$individual[anno$well != "bulk"])
```

```{r}
plot(bulk_raw[, 1], bulk_ruv[, 1])
cor(bulk_raw[, 1], bulk_ruv[, 1])
```


## Correlations

## Session information

```{r info}
sessionInfo()
```
