---
title: "Download Islam et al. 2014"
date: 2015-09-24
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(cache = FALSE)
```

```{r}
library("dplyr")
library("ggplot2")
theme_set(theme_bw(base_size = 12))
```


## Download molecule counts file

```{r}
islam_file <- "../data/GSE46980_CombinedMoleculeCounts.tab.gz"
if (!file.exists(islam_file)) {
  download.file(url = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE46980&format=file&file=GSE46980%5FCombinedMoleculeCounts%2Etab%2Egz",
  destfile = islam_file, method = "wget")
}
islam_data <- read.delim(islam_file, skip = 6)
islam_meta <- islam_data %>% select(Feature:ExonHits)
islam_counts <- islam_data %>% select(X:X.95)
colnames(islam_counts) <- paste0("ESCell_", 1:96)
```

## Download fastq files

First obtain the mapping between each sample in GEO to the sample ID in the Sequence Read Archive (SRA) (strategy from this [Biostars comment](https://www.biostars.org/p/93494/#131604)).

```{r}
library("GEOquery")
gse <- getGEO("GSE46980")
gse <- as.data.frame(gse[[1]], stringsAsFactors = FALSE)
sra_info <- gse %>% filter(grepl("single", title)) %>% # Only keep the data without extra amplication
  select(title, supplementary_file_1) %>%
  mutate(srx = substr(supplementary_file_1, start = 77, stop = 85))
```

Download the SRA meta-information.

```{r}
library("SRAdb")
sqlfile <- "/mnt/gluster/home/jdblischak/SRAmetadb.sqlite"
if(!file.exists(sqlfile)) {
  sqlfile <- getSRAdbFile(destdir = "/mnt/gluster/home/jdblischak/", method = "wget")
}
sra_con <- dbConnect(SQLite(), sqlfile)
```

For each sample, download the associated fastq files and give them a meaningful name.

```{r}
# directory to save Islam et al. data
islam_dir <- "/mnt/gluster/data/internal_supp/singleCellSeq/islam"
dir.create(islam_dir, showWarnings = FALSE)
for (srx_index in 1:nrow(sra_info)) {
  srr <- sraConvert(sra_info$srx[srx_index], "run", sra_con)$run
  for (srr_index in 1:length(srr)) {
    original_name <- sprintf("%s/%s.fastq.gz", islam_dir, srr[srr_index])
    new_name <- sprintf("%s/%s_%d.fastq.gz", islam_dir, sra_info$title[srx_index], srr_index)
    cat(sprintf("%s\t%s\n", original_name, new_name))
    if (!file.exists(new_name)) {
      getSRAfile(srr[srr_index], sra_con, destDir = islam_dir, fileType = "fastq", method = "wget")
      stopifnot(file.exists(original_name))
      file.rename(original_name, new_name)
    }
  }
}
```

Count the number of raw reads in each file.

```{r}
islam_total_reads_file <- "../data/islam-2014-total-reads.txt"
if (!file.exists(islam_total_reads_file)) {
  total_reads <- numeric(length = ncol(islam_counts))
  names(total_reads) <- colnames(islam_counts)
  total_reads <- total_reads[c(1:32, 36:96)]
  for (cell in names(total_reads)) {
    cmd <- sprintf("zcat %s/%s_* | grep '@SRR' | wc -l",
                   islam_dir, cell)
    print(cmd)
    total_reads[cell] <- as.numeric(system(cmd, intern = TRUE))
  }
  write.table(total_reads, file = islam_total_reads_file, quote = FALSE,
              sep = "\t", col.names = FALSE)
} else {
  total_reads_df <- read.table(islam_total_reads_file, stringsAsFactors = FALSE)
  colnames(total_reads_df) <- c("cell", "total_reads")
  total_reads <- total_reads_df$total_reads
  names(total_reads) <- total_reads_df$cell
}
```

## Relationship between total reads and total molecules

```{r}
total_molecules <- colSums(islam_counts)
names(total_molecules) <- colnames(islam_counts)
summary(total_molecules)
```

Merge the two.

```{r}
total_molecules <- total_molecules[names(total_molecules) %in% names(total_reads)]
stopifnot(names(total_molecules) == names(total_reads),
          length(total_molecules) == length(total_reads))
total_counts <- data.frame(cell = names(total_reads), total_reads, total_molecules,
                           stringsAsFactors = FALSE)
```

```{r}
ggplot(total_counts, aes(x = total_reads, y = total_molecules)) +
  geom_point() +
  labs(title = "Islam et al., 2014")
```

## Session information

```{r info}
sessionInfo()
```
