---
title: "Using ngsplot to calculate coverage over ERCC genes"
date: 2015-03-15
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(cache = FALSE)
```

Here I use [ngsplot][] to calculate coverage.

Conclusions:

*  The coverage across the ERCC genes is uniform with a 3' increase
*  Though minimal, there is unexpected coverage in the antisense direction
*  The majority of the signal is coming from the highest expressed genes
*  When split by length, the shortest and longest genes contribute most to the coverage signal

[ngsplot]: https://github.com/shenlab-sinai/ngsplot

Since the ERCC are quite short, I use a fragment length of 50.
The [endogenous](ngsplot-endogenous.html) analysis used a fragment length of 100.

```{r packages, message=FALSE, cache=FALSE}
library("tidyr")
library("ggplot2")
library("cowplot")
theme_set(theme_bw(base_size = 12))
theme_update(panel.grid.minor.x = element_blank(),
             panel.grid.minor.y = element_blank(),
             panel.grid.major.x = element_blank(),
             panel.grid.major.y = element_blank())
```

## Functions

The following function aggregate results from the various ngsplot runs.

```{r functions}
import_ngsplot <- function(results, id = 1:length(results)) {
  # Imports and combines results from multiple ngsplot analyses 
  #
  # results - name of ngsplot results (specified with -O flag)
  # id - description of analysis
  library("tidyr")
  stopifnot(length(results) > 0, length(results) == length(id))
  avgprof_list <- list()
  sem_list <- list()
  for (i in seq_along(results)) {
    zipfile <- paste0(results[i], ".zip")
    extract_zip(zipfile)
    # Import mean coverage
    avgprof_list[[i]] <- import_data(path = results[i], datatype = "avgprof",
                                     id = id[i])
    # Import standard error of mean coverage
    sem_list[[i]] <- import_data(path = results[i], datatype = "sem",
                                 id = id[i])
  }
  avgprof_df <- do.call(rbind, avgprof_list)
  sem_df <- do.call(rbind, sem_list)
  final <- merge(avgprof_df, sem_df)
  return(final)
}

extract_zip <- function(zipfile) {
  # Unzip the ngsplot results into the same directory
  stopifnot(length(zipfile) == 1, file.exists(zipfile))
  unzip(zipfile, exdir = dirname(zipfile))
  return(invisible())
}

import_data <- function(path, datatype, id) {
  # Import the data from a specific ngsplot file.
  #
  # path - path to the ngsplot results directory
  # datatype - either "avgprof" for the mean coverage or
  #            "sem" for the standard error of the mean coverage
  # id - description of analysis (length == 1)
  stopifnot(datatype == "avgprof" | datatype == "sem",
            length(id) == 1)
  fname <- paste0(path, "/", datatype, ".txt")
  df <- read.delim(fname)
  df$position <- paste0("p", 1:nrow(df))
  df$id <- id
  df_long <- gather_(df, key_col = "metainfo", value = datatype)
  df_long$metainfo <- as.character(df_long$metainfo)
  df_long$position <- sub("^p", "", df_long$position)
  df_long$position <- as.numeric(df_long$position)
  return(df_long)
}
```

## Coverage

First I observe the coverage for all filtered ERCC genes.

Unzip and import the raw coverage data.

```{r import-coverage}
cov <- import_ngsplot(results = c("../data/ngsplot-molecules-ercc-both",
                                  "../data/ngsplot-molecules-ercc-same",
                                  "../data/ngsplot-molecules-ercc-opposite"),
                      id = c("ercc-both",
                             "ercc-same",
                             "ercc-opposite"))
cov <- separate(cov, "id", into = c("feature", "strand"), sep = "-")
```

Plotting results.

```{r plot-base}
p <- ggplot(NULL, aes(x = position, y = avgprof, color = metainfo)) +
  geom_line()+
  geom_ribbon(aes(ymin = avgprof - sem, ymax = avgprof + sem,
                  color = NULL, fill = metainfo), alpha = 0.25) +
  geom_vline(x = c(20, 80), color = "grey") +
  scale_x_continuous(breaks = c(0, 20, 40, 60, 80, 100),
                     labels = c(-100, "TSS", "33%", "66%", "TES", 100)) +
  facet_wrap(~strand, ncol = 1, scales = "free_y") +
  theme(legend.position = "none") +
  labs(x = "Position", y = "Mean molecules per million")
```

```{r plot-ercc}
plot_ercc <- p %+% cov + labs(title = "ERCC")
```

### Plot

Note the smaller y-axis for the opposite strand.
While it is very small compared to the sense strand, I highlight it because we expect no reads mapping in the antisense direction.

```{r ngsplot-coverage-ercc, fig.width=8, fig.height=8}
plot_ercc
```

## Coverage by expression level

Next I compare the coverage for NA19091 for genes split into expression quartiles.

```{r import-coverage-ercc-expr}
cov_expr <- import_ngsplot(results = c("../data/ngsplot-ercc-expr-both",
                                       "../data/ngsplot-ercc-expr-same",
                                       "../data/ngsplot-ercc-expr-opposite"),
                           id = c("both", "same", "opposite"))
colnames(cov_expr)[colnames(cov_expr) == "id"] <- "strand"
```

### Plot

```{r plot-ercc-expr, fig.width = 4, fig.height=8}
plot_expr <- plot_ercc %+% cov_expr +
  scale_color_discrete(name = "Expression quartile") +
  scale_fill_discrete(name = "Expression quartile") +
  theme(legend.position = "bottom")
plot_expr
```

Notice the increased y-axis.

## Coverage by gene length

Next I compare the coverage for NA19091 for genes split by gene length.

```{r import-coverage-ercc-len}
cov_len <- import_ngsplot(results = c("../data/ngsplot-ercc-len-both",
                                  "../data/ngsplot-ercc-len-same",
                                  "../data/ngsplot-ercc-len-opposite"),
                      id = c("both", "same", "opposite"))
colnames(cov_len)[colnames(cov_len) == "id"] <- "strand"
```

### Plot

```{r plot-ercc-len, fig.width = 4, fig.height=8}
plot_len <- plot_ercc %+% cov_len +
  scale_color_discrete(name = "Length quartile") +
  scale_fill_discrete(name = "Length quartile") +
  theme(legend.position = "bottom")
plot_len
```

## Session information

```{r info}
sessionInfo()
```
