---
title: "Setup for using ngsplot"
date: 2015-03-07
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(cache = FALSE)
```

[ngsplot][] calculates coverage of genomic regions.
It provides the option to calcalate coverage over the exons for each gene, which is ideal for RNA-seq data.
Furthermore, it has the option to filter the reads based on which strand they map to.
See [here][args] for an explanation of all the program arguments.

[ngsplot]: https://github.com/shenlab-sinai/ngsplot
[args]:https://github.com/shenlab-sinai/ngsplot/wiki/ProgramArguments101

```{r packages, message=FALSE, cache=FALSE}
library("biomaRt")
```

## Modifying the source code

I initially could not run ngsplot because of the ERCC chromosomes in my BAM files.
ngsplot requires all the chromosomes names to either start with "chr" or not start with "chr".
They do this to accomodate reads mapped to UCSC or Ensembl named chromosomes, but I found this [causes problems with mitochondrial reads][issue59].
Because the ngsplot annotation uses the UCSC chromosome names like in my BAM files, I can overide this check, which I did by having the function `chrTag` always return `TRUE` (see [lib/coverage.r][]).

[issue59]: https://github.com/shenlab-sinai/ngsplot/issues/59
[lib/coverage.r]: https://github.com/shenlab-sinai/ngsplot/blob/develop/lib/coverage.r#L544

## Preparing the BAM files

For the ngsplot analyses, I use the BAM files of single cell molecules [created in the coverage analysis where I used the genomation library][bam].

[bam]: coverage-endogenous.html#prepare-bam-files-from-high-quality-cells

## Preparing gene lists

Instead of using all genes, you can provide ngsplot with a list of Ensembl genes to subset (flag `-E`).

### All filtered genes

First I create a list of all endogenous genes that passed the expression filter.

Input filtered molecule counts.

```{r input-molecule-counts-filter}
molecules_filter <- read.table("../data/molecules-filter.txt", header = TRUE,
                               stringsAsFactors = FALSE)
# Remove ERCC
molecules_filter_endo <- molecules_filter[grep("ENSG", rownames(molecules_filter)), ]
cat(rownames(molecules_filter_endo),
    file = "../data/endogenous-filter.txt", sep = "\n")
```

### Coverage by expression level

Next I create four gene lists corresponding to the quartiles of their mean expression level.

```{r expression-quartiles}
mean_expr <- rowMeans(molecules_filter_endo)
expr_q <- quantile(mean_expr)
genes_by_quartile <- cut(mean_expr, breaks = expr_q)
# Purposely converting the factor to the underlying numeric representation
genes_by_quartile <- as.numeric(genes_by_quartile)
names(genes_by_quartile) <- names(mean_expr)
for (i in 1:4) {
  genes <- names(genes_by_quartile)[genes_by_quartile == i]
  q_fname <- paste0("../data/endogenous-filter-q", i, ".txt")
  cat(genes, file = q_fname, sep = "\n")
}
```

### Coverage by gene length

Next I create four gene lists corresponding to the quartiles of their maximum transcript length.

```{r length-quartiles}
ensembl <- useMart(host = "grch37.ensembl.org",
                   biomart = "ENSEMBL_MART_ENSEMBL",
                   dataset = "hsapiens_gene_ensembl")
len_transcripts <- getBM(attributes = c("ensembl_gene_id", "transcript_length"),
                         filters = "ensembl_gene_id",
                         values = rownames(molecules_filter_endo),
                         mart = ensembl)
len_genes <- tapply(len_transcripts$transcript_length,
                    len_transcripts$ensembl_gene_id, max)
stopifnot(rownames(molecules_filter_endo) %in% names(len_genes),
          nrow(molecules_filter_endo) == length(len_genes))
len_q <- quantile(len_genes)
genes_by_len <- cut(len_genes, breaks = len_q)
# Purposely converting the factor to the underlying numeric representation
genes_by_len <- as.numeric(genes_by_len)
names(genes_by_len) <- names(mean_expr)
for (i in 1:4) {
  genes <- names(genes_by_len)[genes_by_len == i]
  q_fname <- paste0("../data/endogenous-filter-len-q", i, ".txt")
  cat(genes, file = q_fname, sep = "\n")
}
```

## Creating ngsplot configuration files

To run multiple BAM files or gene lists at the same time, you can create a configuration file to pass to ngsplot instead of a single BAM file (flag `-C`).

The first configuration file contains the BAM files for one example single cell per individual.
The gene set is all filtered genes.

```{r ngsplot-molecules, engine='bash'}
cat ../data/ngsplot-molecules.txt
```

The second runs the BAM file for 19101 for different subsets of expression level.

```{r ngsplot-expression, engine='bash'}
cat ../data/ngsplot-expression.txt
```

The third runs the BAM file for 19101 for different subsets of gene length.

```{r ngsplot-length, engine='bash'}
cat ../data/ngsplot-length.txt
```

## Makefile

I wrote a Makefile, [make-ngsplot][], that runs the three configuration files for different features and strand specificity.

[make-ngsplot]: https://github.com/jdblischak/singleCellSeq/tree/master/analysis/make-ngsplot

## Session information

```{r info}
sessionInfo()
```
