---
title: "identification of noisy genes"
author: "Po-Yuan Tung"
date: 2015-06-03
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(fig.width = 8, fig.height = 8)
```

## Input

```{r packages, message=FALSE}
library("dplyr")
library("ggplot2")
theme_set(theme_bw(base_size = 16))
library("edgeR")
```

Summary counts from featureCounts.
Created with [gather-summary-counts.py](https://github.com/jdblischak/singleCellSeq/blob/master/code/gather-summary-counts.py).

```{r input-summary-counts}
summary_counts <- read.table("../data/summary-counts.txt", header = TRUE,
                             stringsAsFactors = FALSE)
```

Using only the sickle-trimmed data, sum the counts across all the sequencing lanes for a given sample.

```{r sum-per-sample}
summary_per_sample <- summary_counts %>%
  filter(sickle == "quality-trimmed") %>%
  select(-index, -lane, -flow_cell, -sickle) %>%
  group_by(individual, batch, well, rmdup) %>%
  summarise_each(funs(sum)) %>%
  ungroup %>%
  as.data.frame
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input read counts.

```{r input-read-counts}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input single cell observational quality control data.

```{r input-qc}
qc <- read.table("../data/qc-ipsc.txt", header = TRUE,
                 stringsAsFactors = FALSE)
head(qc)
```

## Remove bad quality cells
### Remove cells with total reads < 2 millons
```{r remove}
#reads per sample
summary_per_sample_reads <- summary_per_sample %>% filter(rmdup == "reads")

#create unmapped ratios
summary_per_sample_reads$unmapped.ratios <- summary_per_sample_reads[,9]/apply(summary_per_sample_reads[,5:13],1,sum)

#create total mapped reads
summary_per_sample_reads$total.mapped <- apply(summary_per_sample_reads[,5:8],1,sum)

#creat ERCC ratios
summary_per_sample_reads$ERCC.ratios <- apply(reads[grep("ERCC", rownames(reads)), ],2,sum)/apply(summary_per_sample_reads[,5:8],1,sum)

#remove bulk keep single cell
summary_per_sample_reads_single <- summary_per_sample_reads[summary_per_sample_reads$well!="bulk",]

#add cell number per well by merging qc file
summary_per_sample_reads_single_qc <- merge(summary_per_sample_reads_single,qc,by=c("individual","batch","well"))

#qc filter
summary_per_sample_reads_single_qc$qc_filter <- summary_per_sample_reads_single_qc$cell_number == 1 & summary_per_sample_reads_single_qc$total.mapped > 2 * 10^6

sum(summary_per_sample_reads_single_qc$qc_filter)

ggplot(summary_per_sample_reads_single_qc, aes(x = total.mapped , y = unmapped.ratios, col = qc_filter)) + geom_text(aes(label = cell_number))

ggplot(summary_per_sample_reads_single_qc, aes(x = total.mapped , y = ERCC.ratios, col = qc_filter)) + geom_text(aes(label = cell_number))

```

##  Total molecule number of ERCC

```{r ERCC}
# molecules per sample
summary_per_sample_molecules <- summary_per_sample %>% filter(rmdup == "molecules")

# total ERCC molecule
summary_per_sample_molecules$total.ERCC.mol <- apply(molecules[grep("ERCC", rownames(reads)), ],2,sum)

# ERCC molecule ratio
summary_per_sample_molecules$ERCC.ratio.mol <- summary_per_sample_molecules$total.ERCC.mol/summary_per_sample_molecules$Assigned

# remove bulk keep single cell
summary_per_sample_molecules_single <- summary_per_sample_molecules[summary_per_sample_molecules$well!="bulk",]

plot(summary_per_sample_molecules_single$total.ERCC.mol)

# adjust total ERCC molecules of 19098 batch2
summary_per_sample_molecules_single$index_19098_2 <- (summary_per_sample_molecules_single$individual == "19098" & summary_per_sample_molecules_single$batch == "2")

# calculating the ratio of 19098 batch 2 to the rest
adjusted_ratio.mol <- mean(summary_per_sample_molecules_single$total.ERCC.mol[summary_per_sample_molecules_single$index_19098_2])/mean(summary_per_sample_molecules_single$total.ERCC[!summary_per_sample_molecules_single$index_19098_2])

adjusted_ratio.mol 

# adjusted total ERCC reads
summary_per_sample_molecules_single$adj.total.ERCC.mol <- summary_per_sample_molecules_single$total.ERCC.mol  

summary_per_sample_molecules_single$adj.total.ERCC.mol[summary_per_sample_molecules_single$index_19098_2] <- summary_per_sample_molecules_single$adj.total.ERCC.mol[summary_per_sample_molecules_single$index_19098_2]/adjusted_ratio.mol

# adjusted ERCC ratios
summary_per_sample_molecules_single$adj.ERCC.ratios.mol <-
summary_per_sample_molecules_single$adj.total.ERCC.mol/summary_per_sample_molecules_single$Assigned

# add qc filter and cell number
summary_per_sample_molecules_single$qc_filter <- summary_per_sample_reads_single_qc$qc_filter
summary_per_sample_molecules_single$cell_number <- summary_per_sample_reads_single_qc$cell_number

ggplot(summary_per_sample_molecules_single, aes(x = Assigned, y = ERCC.ratio.mol, col = qc_filter)) + geom_text(aes(label = cell_number)) + facet_grid(individual ~ batch) + theme(axis.text.x = element_text(angle = 90, hjust = 0.9, vjust = 0.5))

ggplot(summary_per_sample_molecules_single, aes(x = Assigned, y = adj.ERCC.ratios.mol, col = qc_filter)) + geom_text(aes(label = cell_number)) + facet_grid(individual ~ batch) + theme(axis.text.x = element_text(angle = 90, hjust = 0.9, vjust = 0.5))
```

## CV and mean

### Looking at molecule
```{r CV&mean-molecule}
#correct for collision probability
molecules.crt <- -1024*log(1-molecules/1024)

#remove bulk
molecules_single <- molecules.crt %>% select(-contains("bulk"))

# apply qc_filter
molecules_single_qc <- molecules_single[,summary_per_sample_reads_single_qc$qc_filter]
dim(molecules_single_qc)
sum(summary_per_sample_reads_single_qc$qc_filter)

# create a new dataset
molecules_single_qc_w_mean_cv <- molecules_single_qc

# add mean
molecules_single_qc_w_mean_cv$mean <- apply(molecules_single_qc, 1, mean)
# add CV
molecules_single_qc_w_mean_cv$CV <- apply(molecules_single_qc, 1, sd)/ apply(molecules_single_qc, 1, mean)

# remove non-expressed 
molecules_single_qc_expressed <- molecules_single_qc_w_mean_cv[molecules_single_qc_w_mean_cv$mean >0,]
dim(molecules_single_qc_expressed)

# sellect ERCC
molecules_single_qc_expressed_ERCC <- molecules_single_qc_expressed[grep("ERCC",rownames(molecules_single_qc_expressed)),]

molecules_single_qc_expressed$ERCC <- grepl("ERCC",rownames(molecules_single_qc_expressed))

# plot with color-blind-friendly palettes
cbPalette <- c("#999999", "#0000FF", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(molecules_single_qc_expressed, aes(x = mean, y = CV, col = ERCC))  + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_colour_manual(values=cbPalette)

```


### Looking at reads
```{r CV&mean-reads}
# remove bulk
reads_single <- reads %>% select(-contains("bulk"))

# apply qc_filter
reads_single_qc <- reads_single[,summary_per_sample_reads_single_qc$qc_filter]
dim(reads_single_qc)
sum(summary_per_sample_reads_single_qc$qc_filter)

# create a new dataset
reads_single_qc_w_mean_cv <- reads_single_qc

# add mean
reads_single_qc_w_mean_cv$mean <- apply(reads_single_qc, 1, mean)

# add CV
reads_single_qc_w_mean_cv$CV <- apply(reads_single_qc, 1, sd)/ apply(reads_single_qc, 1, mean)

# remove non-expressed 
reads_single_qc_expressed <- reads_single_qc_w_mean_cv[reads_single_qc_w_mean_cv$mean >0,]
dim(reads_single_qc_expressed)

# sellect ERCC
reads_single_qc_expressed$ERCC <- grepl("ERCC",rownames(reads_single_qc_expressed))

# plot with color-blind-friendly palettes
cbPalette <- c("#999999", "#0000FF", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot(reads_single_qc_expressed, aes(x = mean, y = CV, col = ERCC))  + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_colour_manual(values=cbPalette)

```



## Session information

```{r info}
sessionInfo()
```
