---
title: "Ordering effect of the capture sites"
author: "Joyce Hsiao & Po-Yuang Tung"
date: 2015-09-26
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

## Goal

We investigated potential ordering effect of wells on each place on gene
expression across the 9 plates (batches). [Leng](http://dx.doi.org/10.1038/nmeth.3549) et al. discussed ordering effect 
in single-cell RNA-seq experiments using Fluidigm C1 and implemented an algorithm that detects ordering effect of wells on gene expression in [OEFinder](http://dx.doi.org/10.1101/025437).

Note that OEFinder relies on a series of permutations. We had to run OEFinder on a cluster-based computing system.

The results indicated that there are 549 OE genes (out of 20419 genes). GO analysis (DAVID) of these 549 showed that their functions related to membrane lumen (organelle lumen), ribosomal protein, and RNA processing (splicesome), suggesting that they might be all highly expressed genes. 


## Setup

```{r}
source("functions.R")
library(edgeR)
library(ggplot2)
theme_set(theme_bw(base_size = 16))
```



## Prepare single cell data before filtering

Input annotation

```{r}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input read counts.

```{r}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts

```{r}
molecules <- read.table("../data/molecules.txt", header = TRUE, stringsAsFactors = FALSE)
```

Remove bulk samples

```{r}
single_samples <- anno$well != "bulk"
anno_single <- anno[ which(single_samples), ]
molecules_single <- molecules[ , which(single_samples)]
reads_single <- reads[ , which(single_samples)]
stopifnot(ncol(molecules_single) == nrow(anno_single),
          colnames(molecules_single) == anno_single$sample_id)
```


Output single cell samples to txt files.

```{r, eval = FALSE}
if (!file.exists("../data/molecules-single.txt")) {
  write.table(molecules_single, 
              file = "../data/molecules-single.txt",
              quote = FALSE,
              col.names = TRUE, row.names = TRUE)
}
```


Prepare capture site identification file. A txt file with one column of capture site ID (A, B, C, ..., H).

```{r}
require(stringr)
capture_site <- str_extract(anno_single$well, "[aA-zZ]+")
table(capture_site)
```


Save capture_site to a txt file.

```{r}
if (!file.exists("../data/capture-site.txt")) {
write.table(data.frame(site = capture_site), 
            file = "../data/capture-site.txt",
            quote = FALSE,
            col.names = FALSE, row.names = FALSE)
}
```


## OEFinder

For unnormalized data, OEFinder defaults the DESeq normalization method. 

Upload *molecules-single-ENSG.txt* and *capture-site.txt* to OEFinder Shiny GUI interface.

Output to *singleCellSeq/data/OEFinder*.


* Run OEFinder

```{r, eval = FALSE}
# Packages required to start OEFinder
library(shiny)
library(gdata)
library(shinyFiles)
library(EBSeq)

runGitHub("OEFinder", "lengning")

```

Load OEFinder outputted genes.

```{r}
OE_raw <- read.csv("../data/OEFinder/all-genes-OEgenes.csv",
                     stringsAsFactors = FALSE,
                     quote = "\"", sep = ",", header = TRUE)
colnames(OE_raw) <- c("genes", "pvalue")
head(OE_raw)
str(OE_raw)
```


2 ERCC genes in the Overexpressed genes

```{r}
grep("ERCC", OE_raw$genes)
```


Create an indicator variable for the OE genes

```{r}
oefinder_raw <- rownames(molecules_single) %in% as.character(OE_raw$genes)
table(oefinder_raw)
```


## Expression level of OE genes

To answer the question if OE genes are more highly expressed, the average molecule CPM of OE genes or of all genes in each cells were plotted. If the OE genes are highly expressed, it is possible that this position effect is present in all genes but can only be observed with certain amount of expressing level.

```{r}
## remove gene with 0 counts
expressed_single <- rowSums(molecules_single) > 0
molecules_single <- molecules_single[expressed_single, ]
dim(molecules_single)

## remove gene with molecule count larger than 1024 (15 if them)
overexpressed_genes <- rownames(molecules_single)[apply(molecules_single, 1,
                                                        function(x) any(x >= 1024))]
molecules_single <- molecules_single[!(rownames(molecules_single) %in% overexpressed_genes), ]

## collision probability and cpm molecule counts
molecules_single_collision <- -1024 * log(1 - molecules_single / 1024)
molecules_single_cpm <- cpm(molecules_single_collision, log = TRUE)

## select for OE genes 
molecules_single_OE <- molecules_single_cpm[rownames(molecules_single_cpm) %in% as.character(OE_raw$genes),]

## boxplot
anno_single$OE_ave <- apply(molecules_single_OE, 2, mean)
anno_single$all_gene_ave <- apply(molecules_single_cpm, 2, mean)

ggplot(anno_single, aes(x = as.factor(individual))) + geom_boxplot(aes( y = OE_ave, fill = as.factor(batch), alpha = 0.25)) + geom_boxplot(aes( y = all_gene_ave, fill = as.factor(batch))) +  labs (x = "individual", y = "avarage molecule cpm", alpha = "OE", fill = "batch")
```


## Distribution of OE genes

```{r}
cutoffs <- seq(1001, nrow(molecules_single_cpm), by = 1000)
cutoffs <- c(cutoffs, nrow(molecules_single_cpm))
top_genes_count <- lapply(1:length(cutoffs), function(cut) {
                        per_cutoff <- cutoffs[cut]
                        cell_across_order <- order(rowSums(molecules_single_cpm), decreasing = TRUE)
                        top_genes <- rownames(molecules_single_cpm)[cell_across_order < per_cutoff]
                        sum(OE_raw$genes %in% top_genes)
                        })
top_genes_count <- do.call(c, top_genes_count)

ggplot(data.frame(top_count = top_genes_count,
                  cutoffs = cutoffs), 
       aes(x = as.factor(cutoffs), y = top_count)) + geom_point() +
       labs(x = "Top X genes", y = "Number of OEFinder OE genes")
```

OE genes identified by OEFinder were not limited to the top 1000 genes. On the contrary, we found
OE genes at all levels of gene expression (averaged acrosss cells).




## Session information

```{r info}
sessionInfo()
```
