---
title: "Build a cell-cycle classifer using FACS sorted data"
author: "Joyce Hsiao"
date: 2016-02-03  
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")

library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE)
```



## Background and objectives


## Conclusions


## Set up

```{r, message=FALSE, warning=FALSE}
library("matrixStats")
library("glmnet")
library("knitr")
library("mygene")
```


## Prepare data

Import processed counts.


```{r import-data}
data <- read.csv("../data/GSE64016_H1andFUCCI_normalized_EC.csv")
gene_names <- data[,1]
data <- data[,-1]
rownames(data) <- gene_names

## The gene IDs provided, not Ensembl IDs
gene_names[1:10]

## Extract the cell labels
cell_phases <- sapply(colnames(data), function(x) strsplit(x,"_")[[1]][1]);

table(cell_phases)

cycle_data <- data[ ,cell_phases != "H1"]
dim(cycle_data)
cycle_vector <- factor(cell_phases[cell_phases != "H1"], 
                       levels = c("G1", "S", "G2"))
cycle_vector <- as.integer(cycle_vector)
table(cycle_vector)
```


Check column sums. These are EBseq normalized counts. 

*[chunk not evaluated]*
```{r check-properties, eval = FALSE}
par(mfrow = c(1,1))
boxplot(log2(cycle_data))
```


## Framework

Let $K_{g,i(s)}$ be the number of counts mapped to gene $g$ in cell $s$ sampled from
cell-cycle phase $k$. We assume that 


## ungrouped lasso

```{r}
fit_ungrouped <- glmnet(x = t(cycle_data) , y = cycle_vector, alpha = 1, 
                        type.multinomial = "ungrouped", 
                        family = "multinomial")
head(print(fit_ungrouped))
```

```{r}
plot(fit_ungrouped, xvar = "lambda", label = TRUE, type.coef = "2norm")

plot(log10(fit_ungrouped$lambda),
     fit_ungrouped$dev.ratio,
     ylab = "percent of null variance explained",
     xlab = "log10 lambda")
```

cross-validation on the glmnet fit

```{r}
cvfit <- cv.glmnet(x = t(cycle_data), 
                   y = cycle_vector, 
                   family="multinomial", 
                   type.multinomial = "ungrouped", parallel = TRUE)
plot(cvfit)
```

```{r}
df_coef <- data.frame(v1 = matrix(coef(cvfit, s = "lambda.min")[[1]]),
                      v2 = matrix(coef(cvfit, s = "lambda.min")[[2]]),
                      v3 = matrix(coef(cvfit, s = "lambda.min")[[3]]) )
length(which(rowSums(df_coef != 0) > 0))
```


```{r}
as.character(gene_names[which(rowSums(df_coef != 0) > 0)])
```


```{r}
query_genes <- queryMany(c(as.character(gene_names[which(rowSums(df_coef != 0) > 0)])),  
                         scopes="symbol", 
                         fields=c("name","summary"), species="human")
kable(query_genes)
```


## Investigate selected genes

```{r}
df_coef_selected <- df_coef[which(rowSums(df_coef != 0) > 0), ]
df_coef_relative_S <- data.frame(
    G1_to_S = df_coef_selected[ ,1] - df_coef_selected[ ,2], 
    G2_to_S = df_coef_selected[ ,3] - df_coef_selected[ ,2])
#round(df_coef_relative_S, 4)
```


## Session information

```{r info}
sessionInfo()
```
