---
title: "Additional analysis for the manuscript"
date: 2016-05-02
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(fig.width = 8, fig.height = 8)
```


```{r packages, message=FALSE}
library("dplyr")
library("ggplot2")
library("cowplot")
library("lmtest")
library("lme4")
source("functions.R")
```

## Basic framework




## Questions

1. Concentration vs. total molecule-count (ENSG)

(Our study design allows us to investigate...)

total molecule-count ~ Concentration 

[previous work](http://jdblischak.github.io/singleCellSeq/analysis/compare-reads-v-molecules-per-batch-test.html)

2. Molecule-to-read conversion rate (ENSG, ERCC)

(We explored...)

total molecule-count ~ read

3. total ERCC molecule-count and total ENSG molecule-count

(Could we account...)

total ENSG molecule-count ~ total ERCC molecule-count

4. Percent variation explained by individual and replicate effect in ENSG and ERCC

(As a first step...)

[previous work](http://jdblischak.github.io/singleCellSeq/analysis/compare-reads-v-molecules-per-batch-test.html)


## Input

Input filtered annotation.

```{r input-annotation-filter}
anno_filter <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_filter)
```

Input filtered molecule counts.

```{r input-molecule-counts-filter}
molecules_filter <- read.table("../data/molecules-filter.txt", header = TRUE,
                               stringsAsFactors = FALSE)
molecules_filter_ENSG <- molecules_filter[grep("ERCC", rownames(molecules_filter), invert = TRUE), ]
  
stopifnot(ncol(molecules_filter) == nrow(anno_filter),
          colnames(molecules_filter) == anno_filter$sample_id)
```

Input filtered read counts

```{r}
reads_filter <- read.table("../data/reads-filter.txt", header = TRUE,
                    stringsAsFactors = FALSE)
reads_filter_ENSG <- reads_filter[grep("ERCC", rownames(reads_filter), invert = TRUE), ]

stopifnot(all.equal(colnames(reads_filter_ENSG), 
                    colnames(molecules_filter_ENSG)))
```

Input quality control file. Filter cells to match cells in molecules_filter.

```{r}
qc <- read.table("../data/qc-ipsc.txt", header = TRUE,
                 stringsAsFactors = FALSE)
qc$sample_id <- with(qc, paste0(individual, ".", replicate, ".", well))

qc_filter <- qc[match(anno_filter$sample_id, qc$sample_id), ]

stopifnot(all.equal(qc_filter$sample_id, anno_filter$sample_id))
```

Input standardized molecule counts.

```{r input-molecule-counts-cpm}
molecules_cpm <- read.table("../data/molecules-cpm.txt", header = TRUE,
                            stringsAsFactors = FALSE)
stopifnot(ncol(molecules_cpm) == nrow(anno_filter),
          colnames(molecules_cpm) == anno_filter$sample_id)
```

Input Poisson GLM transformed molecule counts per million.

```{r input-molecule-counts-trans}
molecules_cpm_trans <- read.table("../data/molecules-cpm-trans.txt", header = TRUE,
                               stringsAsFactors = FALSE)
stopifnot(ncol(molecules_cpm_trans) == nrow(anno_filter),
          colnames(molecules_cpm_trans) == anno_filter$sample_id)
```

Input final batch-corrected molecule counts per million.

```{r input-molecule-counts-final}
molecules_final <- read.table("../data/molecules-final.txt", header = TRUE,
                              stringsAsFactors = FALSE)
stopifnot(ncol(molecules_final) == nrow(anno_filter),
          colnames(molecules_final) == anno_filter$sample_id)
```





## Concentration vs. total molecule-count (ENSG)

As we try to understand the general relationships between sequencing results and cellular mRNA content, we remove outlier batches. NA19098 replicate 1 failed the quantification of the concentration of the single cells and was hence removed. Because NA19098 concentration is only quantified in one replicate, we removed NA19098 from analysis involving batch differences and concentration.

```{r filter-outliers}
anno_single <- anno_filter
ercc_index <- grepl("ERCC", rownames(molecules_filter))
anno_single$total_molecules_gene = colSums(molecules_filter[!ercc_index, ])
anno_single$total_molecules_ercc = colSums(molecules_filter[ercc_index, ])
anno_single$total_molecules = colSums(molecules_filter)
anno_single$num_genes = apply(molecules_filter[!ercc_index, ], 2, function(x) sum(x > 0))
anno_single$concentration <- qc_filter$concentration[match(anno_single$sample_id, qc_filter$sample_id)]

anno_single <- anno_single %>% filter(individual != "NA19098")
anno_single$individual <- as.factor(anno_single$individual)
anno_single$replicate <- as.factor(anno_single$replicate)
```

Correlation between total molecule-count and concentration. 

```{r}
with(anno_single, 
     cor.test(total_molecules_gene /(10^3), concentration, 
              method = "spearman"))

with(anno_single[anno_single$individual == "NA19101",], 
     cor.test(total_molecules_gene /(10^3), concentration, 
              method = "spearman"))

with(anno_single[anno_single$individual == "NA19239",], 
     cor.test(total_molecules_gene /(10^3), concentration, 
              method = "spearman"))

# per replicate
sapply(unique(anno_single$batch), function(batch) {
with(anno_single[anno_single$batch == batch,], 
     cor(total_molecules_gene /(10^3), concentration, 
              method = "spearman"))
})
```

We take total molecule-count divided by 1000.

```{r}
fit <- lmer(total_molecules_gene /(10^3)~ concentration + individual + 
                (1|individual:replicate),
              data = anno_single)
summary(fit)

fit_1 <- lm(total_molecules_gene /(10^3) ~ concentration + individual, 
            data = anno_single)
fit_2 <- lmer(total_molecules_gene /(10^3) ~ concentration +
                (1|individual:replicate), data = anno_single)

# significance of individual effect
lrtest(fit_2, fit)

# significance of replicate effect
anova(fit, fit_1)
```


## Reads to molecule conversion efficiency 

Prepare ERCC data

```{r}
reads_ERCC <- reads_filter[grep("ERCC", rownames(reads_filter), 
                                invert = FALSE), ]
molecules_ERCC <- molecules_filter[grep("ERCC", rownames(molecules_filter), 
                                invert = FALSE), ]

total_counts_ERCC <- data.frame(total_reads = colSums(reads_ERCC),
                                total_molecules = colSums(molecules_ERCC))
total_counts_ERCC$conversion <- with(total_counts_ERCC,
                                     total_molecules/total_reads)
total_counts_ERCC$individual <- as.factor(anno_filter$individual[match(rownames(total_counts_ERCC),
                          anno_filter$sample_id)])
total_counts_ERCC$replicate <- as.factor(anno_filter$replicate[match(rownames(total_counts_ERCC),
                          anno_filter$sample_id)])
```


Prepare ENSG data

```{r}
reads_ENSG <- reads_filter[grep("ERCC", rownames(reads_filter), 
                                invert = TRUE), ]
molecules_ENSG <- molecules_filter[grep("ERCC", rownames(molecules_filter), 
                                invert = TRUE), ]

total_counts_ENSG <- data.frame(total_reads = colSums(reads_ENSG),
                                total_molecules = colSums(molecules_ENSG))
total_counts_ENSG$conversion <- with(total_counts_ENSG,
                                     total_molecules/total_reads)
total_counts_ENSG$individual <- as.factor(anno_filter$individual[match(rownames(total_counts_ENSG),
                          anno_filter$sample_id)])
total_counts_ENSG$replicate <- as.factor(anno_filter$replicate[match(rownames(total_counts_ENSG),
                          anno_filter$sample_id)])
```


Concentration distribution is skewed for both ERCC and ENSG genes. Hence, we analyze the log2 conversion data (log2 base was taken so that our log transformation base is consistent throughout the paper).

```{r}
par(mfrow = c(2,2))
hist(total_counts_ERCC$conversion,
     main = "ERCC spike-in",
     xlab = "Conversion")
hist(log2(total_counts_ERCC$conversion),
     main = "ERCC spike-in",
     xlab = "log2 Conversion")
hist(total_counts_ENSG$conversion,
     main = "Endogenous genes",
     xlab = "Conversion")
hist(log2(total_counts_ENSG$conversion),
     main = "Endogenous genes",
     xlab = "log2 Conversion")
```


ERCC Model fitting

```{r}
fit <- lmer(log2(conversion) ~ individual + 
              (1|individual:replicate), 
            data = total_counts_ERCC)
fit_1 <- lm(log2(conversion) ~ individual,
            data = total_counts_ERCC)
fit_2 <- lmer(log2(conversion) ~ 1 + 
              (1|individual:replicate), 
            data = total_counts_ERCC)

# significance of individual effect
lrtest(fit_2, fit)

# significance of replicate effect
anova(fit, fit_1)
```



ENSG Model fitting

```{r}
fit <- lmer(log2(conversion) ~ individual + 
              (1|individual:replicate), 
            data = total_counts_ENSG)
fit_1 <- lm(log2(conversion) ~ individual,
            data = total_counts_ENSG)
fit_2 <- lmer(log2(conversion) ~ 1 + 
              (1|individual:replicate), 
            data = total_counts_ENSG)

# significance of individual effect
lrtest(fit_2, fit)

# significance of replicate effect
anova(fit, fit_1)
```


## total ENSG molecule-count and total ERCC molecule-count

Prepare data

```{r}
anno_temp <- anno_filter

anno_temp$ensg_total_count <- 
  colSums(molecules_filter[grep("ERCC", 
                           rownames(molecules_filter), invert = FALSE), ])
anno_temp$ercc_total_count <- 
  colSums(molecules_filter[grep("ERCC", 
                           rownames(molecules_filter), invert = TRUE), ])
```

Check if total count ERCC is normal distribution 

```{r}
par(mfrow = c(2,2))
hist(anno_temp$ercc_total_count,
     main = "ERCC spike-in",
     xlab = "Total counts")
hist(log2(anno_temp$ercc_total_count),
     main = "ERCC spike-in",
     xlab = "log2 Total counts")
hist(anno_temp$ensg_total_count,
     main = "Endogenous genes",
     xlab = "Total counts")
hist(log2(anno_temp$ensg_total_count),
     main = "Endogenous genes",
     xlab = "log2 Total counts")
```


First, we assess total ERCC molecule-count beteween individuals and replicates.

```{r}
fit <- lmer(log2(ercc_total_count) ~ individual + 
              (1|individual:replicate), 
            data = anno_temp)
fit_1 <- lm(log2(ercc_total_count) ~ individual,
            data = anno_temp)
fit_2 <- lmer(log2(ercc_total_count) ~ 1 + 
              (1|individual:replicate), 
            data = anno_temp)

# significance of individual effect
lrtest(fit_2, fit)

# significance of replicate effect
anova(fit, fit_1)
```


Second, we perform the same analysis on total ENSG molecule-counts.


```{r}
fit <- lmer(log2(ensg_total_count) ~ individual + 
              (1|individual:replicate), 
            data = anno_temp)
fit_1 <- lm(log2(ensg_total_count) ~ individual,
            data = anno_temp)
fit_2 <- lmer(log2(ensg_total_count) ~ 1 + 
              (1|individual:replicate), 
            data = anno_temp)

# significance of individual effect
lrtest(fit_2, fit)

# significance of replicate effect
anova(fit, fit_1)
```


Second, we include total ENSG molecule-count in the model of total ERCC molecule-count in addition to individual and replicate factors.

ERCC ~ ENSG

```{r}
fit <- lmer(log2(ercc_total_count) ~ log2(ensg_total_count) + individual + 
              (1|individual:replicate), 
            data = anno_temp)
fit_1 <- lm(log2(ercc_total_count) ~ log2(ensg_total_count) + individual,
            data = anno_temp)
fit_2 <- lmer(log2(ercc_total_count) ~ log2(ensg_total_count) + 
              (1|individual:replicate), 
            data = anno_temp)

# significance of individual effect
lrtest(fit_2, fit)

# significance of replicate effect
anova(fit, fit_1)
```


## Percent variation per gene

Load model fitting code - a wrapper of the blmer function that fits a bayesian nested model for one gene at a time. 

```{r percent-var-code, eval = TRUE}
#' Per gene variance component model
#'
#' @param xx Matrix of expression measurements on log scale.
#' @param annotation Meta-data matrix of each column of xx.

gene_variation <- function(xx, annotation) {
  
  individual <- as.factor(annotation$individual)
  replicate <- as.factor(annotation$replicate)
  
  ## fit bayesian GLM one gene at a time
  
  blme_fit <- lapply( 1:NROW(counts), function(i) {
      
      value <- unlist(counts[i,])
      
      fit_try <- tryCatch( fit <- blme::blmer(value ~ 1|individual/replicate, 
                                      cov.prior = wishart,
                                      resid.prior = invgamma),
                             condition = function(c) c)
      if(inherits(fit_try, "condition")){
      var_foo <- rep(NA, 3)
      return(var_foo)
      }
      if(!inherits(fit_try, "condition")){
        var_foo <- as.data.frame(VarCorr(fit_try))[,4]
        var_foo <- var_foo[c(2,1,3)]
        var_foo
      }
    })
  blme_fit <- do.call(rbind, blme_fit)
  rownames(blme_fit) <- rownames(counts)
  colnames(blme_fit) <- c("individual","replicate","residual")
  blme_fit
}
```


```{r, eval = FALSE}
individual <- as.factor(anno_filter$individual)
replicate <- as.factor(anno_filter$replicate)

## endogenous molecule-count
blme_raw <- gene_variation(xx = log2(molecules_filter_ENSG+1),
                           annotation = anno_filter)

## ERCC molecule-count
blme_ercc <- gene_variation(xx = log2(molecules_filter_ERCC+1),
                            annotation = anno_filter)

## ENSG CPM
blme_cpm <- gene_variation(xx = molecules_cpm, 
                           annotation = anno_filter)

## ENSG CPM Poisson
blme_cpm_trans <- gene_variation(xx = molecules_cpm_trans, 
                           annotation = anno_filter)

## ENSG CPM Poisson
blme_final <- gene_variation(xx = molecules_final, 
                           annotation = anno_filter)

save(blme_raw, blme_ercc, blme_cpm, blme_cpm_trans,
     blme_final,
     file = "../data/blme-variance.rda")
```

Standardardize variance components such that for each gene the variance 
component estimates sum up to 1. Because variance of sequencing counts depend on the dynamic range, we avoid biased comparison of variance estimates when comparing across genes, by the action of standarizing per gene variance components by the total of variance estimates.

*Endogenous genes*

```{r}
load("../data/blme-variance.rda")
blme_raw_norm <- (blme_raw/rowSums(blme_raw))

par(mfrow = c(2,2))
plot(blme_raw_norm[,1], blme_raw_norm[,2],
     xlab = "Individual",
     ylab = "Replicate")
abline(0, 1, col = "red")
plot(blme_raw_norm[,3], blme_raw_norm[,1],
     xlab = "Residual",
     ylab = "Individual")
plot(blme_raw_norm[,3], blme_raw_norm[,2],
     xlab = "Residual",
     ylab = "Replicate")
boxplot(blme_raw_norm)
title(main = "Proportion of per-gene variance",
      outer = TRUE, line = -1)
```

Compare individual versus replicate variance estimates.

```{r}
summary(blme_raw_norm)

# Kruskal wallis rank sum test to compare
# proportions of variance explained due to individual
# versus due to replicate
kruskal.test(c(blme_raw_norm[,1], blme_raw_norm[,2]) ~
               rep(c(1,2), each = NROW(blme_raw_norm)) )
```


*ERCC spike-in genes*

```{r}
load("../data/blme-variance.rda")
blme_ercc_norm <- (blme_ercc/rowSums(blme_ercc))

par(mfrow = c(2,2))
plot(blme_ercc_norm[,1], blme_ercc_norm[,2],
     xlab = "Individual",
     ylab = "Replicate")
abline(0, 1, col = "red")
plot(blme_ercc_norm[,3], blme_ercc_norm[,1],
     xlab = "Residual",
     ylab = "Individual")
plot(blme_ercc_norm[,3], blme_ercc_norm[,2],
     xlab = "Residual",
     ylab = "Replicate")
boxplot(blme_ercc_norm)
title(main = "Proportion of per-gene variance")
```

Compare individual versus replicate variance estimates.

```{r}
summary(blme_ercc_norm)

# Kruskal wallis rank sum test to compare
# proportions of variance explained due to individual
# versus due to replicate
kruskal.test(c(blme_ercc_norm[,1], blme_ercc_norm[,2]) ~
               rep(c(1,2), each = NROW(blme_ercc_norm)) )
```


*Per-gene variance estimates after each normalization step*


```{r, fig.width=9, fig.height=9}
# compute percent variation
blme_raw_norm <- data.frame(blme_raw/rowSums(blme_raw))
blme_cpm_norm <- data.frame(blme_cpm/rowSums(blme_cpm))
blme_cpm_trans_norm <- data.frame(blme_cpm_trans/rowSums(blme_cpm_trans))
blme_final_norm <- data.frame(blme_final/rowSums(blme_final))

library(ggplot2)
cowplot::plot_grid(
  ggplot(data.frame(
    prop = c(blme_cpm_norm$individual/blme_raw_norm$individual,
             blme_cpm_trans_norm$individual/blme_cpm_norm$individual,
             blme_final_norm$individual/blme_cpm_trans_norm$individual),
    label = factor(rep(c(1:3), 
                each = NROW(blme_raw_norm)),
                levels = c(1:3),
                labels = c("CPM-Raw", "Poisson/CPM", "Final/Poisson"))),
       aes(x = label, y = prop)) +
    geom_violin() + geom_boxplot(width = .2, aes(colour = label)) +
      xlab("") +
      ylab("Individual percent-variance ratio") +
      ggtitle("Individual percent-variance ratio (to the previous step)") +
      theme(legend.position = "none"),
  ggplot(data.frame(
    prop = c(blme_cpm_norm$individual/blme_raw_norm$individual,
             blme_cpm_trans_norm$individual/blme_raw_norm$individual,
             blme_final_norm$individual/blme_raw_norm$individual),
    label = factor(rep(c(1:3), 
                each = NROW(blme_raw_norm)),
                levels = c(1:3),
                labels = c("CPM/Raw", "Poisson/Raw", "Final/Raw"))),
       aes(x = label, y = prop)) +
    geom_violin() + geom_boxplot(width = .2, aes(colour = label)) +
      xlab("") +
      ylab("Individual percent-variance ratio") +
      ggtitle("Individual percent-variance ratio (to Step Raw)") +
      theme(legend.position = "none"),
  ggplot(data.frame(
    prop = log(c(blme_raw_norm$individual/blme_raw_norm$replicate,
             blme_cpm_norm$individual/blme_cpm_norm$replicate,
             blme_cpm_trans_norm$individual/blme_cpm_trans_norm$replicate,
             blme_final_norm$individual/blme_final_norm$replicate)),
    label = factor(rep(c(1:4), 
                each = NROW(blme_raw_norm)),
                levels = c(1:4),
                labels = c("Raw", "CPM", "Poisson", "Final"))),
       aes(x = label, y = prop)) +
    geom_violin() + geom_boxplot(width = .2, aes(colour = label)) +
      xlab("") +
      ylab("log ratio of percent variance") +
      ggtitle("Individual to replicate log-ratio") +
      theme(legend.position = "none"),
  labels = LETTERS[1:4])
```

Per-gene variation estimates between steps fail to reveal changes induced in the expression profiles between steps, if any. 

*Multivariate distance between C1 preparations*: Friedman-Rafsky multivariate run test.

```{r distance-code}
if (library("flowMap", quietly = TRUE, logical.return = TRUE) == FALSE) {
  devtools::install_github("jhsiao999/flowMap")
} else {
  library(flowMap)
}
compute_fr <- function(xx, annotation) {
  batches <- unique(annotation$batch)
  fr_pval <- fr_run <- matrix(0, nrow = length(batches),
                       ncol = length(batches))
  indices <- which(upper.tri(fr_dist, diag = FALSE), arr.ind = TRUE)
  for (i in 1:NROW(indices)) {
    ind_row <- indices[i, 1]
    ind_col <- indices[i, 2]
    fr_res <- 
          getFR(t(as.matrix(xx[,annotation$batch == batches[ind_row]])),
            t(as.matrix(xx[,anno_filter$batch == batches[ind_col]])))
    fr_pval[ind_row, ind_col]  <- fr_res$pNorm
    fr_run[ind_row, ind_col]  <- fr_res$ww
  }
  fr_pval <- Matrix::forceSymmetric(fr_pval)
  diag(fr_pval) <- 1
  return(list(fr_pval = fr_pval,
              fr_run = fr_run))
}
```

Compute multivariate distance for matrices after each step of transformation.
```{r, eval = F}
fr_raw <- compute_fr(xx = molecules_filter_ENSG,
                     annotation = anno_filter)
fr_cpm <- compute_fr(xx = molecules_cpm,
                     annotation = anno_filter)
fr_cpm_trans <- compute_fr(xx = molecules_cpm_trans,
                     annotation = anno_filter)
fr_final <- compute_fr(xx = molecules_final,
                     annotation = anno_filter)
save(fr_raw, fr_cpm, fr_cpm_trans, fr_final,
     file = "../data/fr-distance.rda")
```

Make p-value heatmaps

```{r}
load(file = "../data/fr-distance.rda")
library(gplots)
heatmap.2(as.matrix(fr_final$fr_pval), 
          dendrogram = "column",
          trace = "none", Rowv = FALSE,
          labRow = unique(anno_filter$batch),
          labCol = unique(anno_filter$batch),
          key = TRUE, main = "Raw")
heatmap.2(as.matrix(fr_cpm$fr_pval), 
          dendrogram = "column",
          trace = "none", Rowv = FALSE,
          labRow = unique(anno_filter$batch),
          labCol = unique(anno_filter$batch),
          key = TRUE, main = "CPM")
heatmap.2(as.matrix(fr_cpm_trans$fr_pval), 
          dendrogram = "column",
          trace = "none", Rowv = FALSE,
          labRow = unique(anno_filter$batch),
          labCol = unique(anno_filter$batch),
          key = TRUE, main = "Poisson")
heatmap.2(as.matrix(fr_final$fr_pval), 
          dendrogram = "column",
          trace = "none", Rowv = FALSE,
          labRow = unique(anno_filter$batch),
          labCol = unique(anno_filter$batch),
          key = TRUE, main = "Final")
```


## Correlation between and within batches

Code for computing correlation between cells within batches and between batches.

```{r compute-corr-batch, eval = TRUE}
#' molecules_input <- molecules_filter_ENSG
#' annotation <- anno_filter
compute_corr_batch <- function(molecules_input, annotation) {

  cor_mat <- cor(molecules_input, method = "spearman")
  batch <- unique(annotation$batch)
  individual <- unique(annotation$individual)

  # same individual, within batch
  corr_same_ind_within_batch <-
    lapply(1:length(individual), function(i) {
      batch <-
        unique(annotation$batch[annotation$individual == individual[i]])

      corr_batch <- lapply(1:length(batch), function(i) {
        df <- cor_mat[annotation$batch == batch[i],
                      annotation$batch == batch[i]]
        df[upper.tri(df, diag = FALSE)]
      })
      unlist(corr_batch)
    })

  # same individual, between replicates
  corr_same_ind_between_batch <-
    lapply(1:length(individual), function(i) {
      batch <-
        unique(annotation$batch[annotation$individual == individual[i]])
      submat <- lapply(1:(length(batch)-1), function(i) {
        submat0 <- lapply(2:length(batch), function(j) {
          df <- cor_mat[annotation$batch == batch[i],
                        annotation$batch == batch[j]]
          df[upper.tri(df, diag = FALSE)]
        })
        unlist(submat0)
      })
      unlist(submat)
    })

  # different individual
  corr_diff_ind_between_batch <-
    lapply(1:(length(individual)-1), function(i) {
      if (i == 1) {
        batch <-
          unique(annotation$batch[annotation$individual == individual[i]])
        batch_other <-
          unique(annotation$batch[annotation$individual != individual[i+1]])
      }
      if (i == 2) {
        batch <-
          unique(annotation$batch[annotation$individual == individual[i]])
        batch_other <-
          unique(annotation$batch[annotation$individual == individual[i+1]])
      }

      submat <- lapply(1:length(batch), function(i) {
        submat0 <- lapply(1:length(batch_other), function(j) {
          df <- cor_mat[annotation$batch == batch[i],
                        annotation$batch == batch_other[j]]
          df[upper.tri(df, diag = FALSE)]
        })
        unlist(submat0)
      })
      unlist(submat)
    })
  corr_diff_ind_between_batch <- unlist(corr_diff_ind_between_batch)

  return( list(corr_same_ind_within_batch = corr_same_ind_within_batch,
               corr_same_ind_between_batch = corr_same_ind_between_batch,
               corr_diff_ind_between_batch = corr_diff_ind_between_batch) )

}
```


Compute correlation for molecule-count data after filtering.

```{r}
corr_filter <- compute_corr_batch(molecules_filter_ENSG, anno_filter)
par(mfrow = c(1,1))
boxplot(cbind(corr_filter[[1]][[1]],
              corr_filter[[1]][[2]],
              corr_filter[[1]][[3]],
              corr_filter[[2]][[1]],
              corr_filter[[2]][[2]],
              corr_filter[[2]][[3]],
              corr_filter[[3]]),
        main = "Counts",
        axes = F)
axis(1, at = c(1:7),
     labels = c("within-batch-19098",
                      "within-batch-19101",
                      "within-batch-19239",
                      "between-batch-19098",
                      "between-batch-19101",
                      "between-batch-19239",
                      "across-batches"))
axis(2)
```

Kruskal wallis comparing all between-batch correlations with
all within-batch correlations


```{r}
df <- data.frame(corrs = c(unlist(corr_filter[[1]]),
                                unlist(corr_filter[[2]])),
                        label = c(rep(1, length(unlist(corr_filter[[1]]))),
                                  rep(2, length(unlist(corr_filter[[2]]))))) 
kruskal.test(df$corrs ~ df$label)

# summary statistics of correlations within-batches
# of all three individuals
summary(unlist(corr_filter[[1]]))

# summary statistics of correlations between-batches
# of all three individuals
summary(unlist(corr_filter[[2]]))
```


## Session information

```{r info}
sessionInfo()
```


