---
title: "PCA quantification"
date: 2016-04-23
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(fig.width = 8, fig.height = 8)
```

```{r packages, message=FALSE}
library("HDMD")
library("ggplot2")
theme_set(theme_bw(base_size = 12))
library("tidyr")
```

Needs to be run after the [data transformation analysis][data-transformation.html] because it uses the PCA results.
We want to demonstrate the effect of our normalization steps on the technical variance as observed in the PCA plots.

## Input

Input filtered annotation.

```{r input-annotation-filter}
anno_filter <- read.table("../data/annotation-filter.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno_filter)
```

Input PCA results from [data transformation analysis][data-transformation.html].

```{r input-pca}
pca_raw <- read.table("../data/pca-molecules-filter.txt",
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
pca_cpm <- read.table("../data/pca-molecules-cpm.txt",
                      header = TRUE, sep = "\t", stringsAsFactors = FALSE)
pca_pois <- read.table("../data/pca-molecules-cpm-trans.txt",
                       header = TRUE, sep = "\t", stringsAsFactors = FALSE)
pca_final <- read.table("../data/pca-final.txt",
                       header = TRUE, sep = "\t", stringsAsFactors = FALSE)
stopifnot(pca_raw$Obs.batch == anno_filter$batch,
          pca_cpm$Obs.batch == anno_filter$batch,
          pca_pois$Obs.batch == anno_filter$batch,
          pca_final$Obs.batch == anno_filter$batch)
```

## Quantifying PCA results

First visualizing how the pairwise differences between individuals and between replicates of the same individual change after each normalization step.

```{r pca-quant}
pca_results <- list(raw = pca_raw[, 3:4],
                    cpm = pca_cpm[, 3:4],
                    poisson = pca_pois[, 3:4],
                    final = pca_final[, 3:4])
distance_results <- NULL
for (analysis in names(pca_results)) {
  d <- pairwise.mahalanobis(x = pca_results[[analysis]],
                            grouping = 1:nrow(pca_results[[analysis]]))$distance
  rownames(d) <- anno_filter$sample_id
  colnames(d) <- anno_filter$sample_id
  d_long <- gather(as.data.frame(d), key = "sample1", value = "distance")
  d_long$sample2 <- colnames(d)
  stopifnot(d_long$d[d_long$sample1 == d_long$sample2] == 0)
  d_long$analysis <- analysis
  distance_results <- rbind(distance_results, d_long)
}
distance_results <- separate(distance_results, sample1, into = c("individual1", "replicate1", "well1"), sep = "\\.", remove = FALSE)
distance_results <- separate(distance_results, sample2, into = c("individual2", "replicate2", "well2"), sep = "\\.", remove = FALSE)
# Remove self-comparisons
distance_results <- distance_results[distance_results$sample1 !=
                                     distance_results$sample2, ]
# Classify by comparison
distance_results$same_ind <- distance_results$individual1 == distance_results$individual2
distance_results$same_rep <- distance_results$replicate1 == distance_results$replicate2
distance_results$comparison <- NA
distance_results$comparison[distance_results$same_ind &
                            distance_results$same_rep] <- "Same individual,\nsame replicate"
distance_results$comparison[distance_results$same_ind &
                            !distance_results$same_rep] <- "Same individual,\ndifferent replicate"
distance_results$comparison[!distance_results$same_ind] <- "different individual"
stopifnot(!is.na(distance_results$comparison))
table(distance_results$comparison, useNA = "ifany")
```

Order the factors.

```{r order-factors}
distance_results$comparison <- factor(distance_results$comparison,
                                      levels = c("different individual",
                                                 "Same individual,\ndifferent replicate",
                                                 "Same individual,\nsame replicate"))
distance_results$analysis <- factor(distance_results$analysis,
                                    levels = c("raw", "cpm", "poisson", "final"))
```

Plot.

```{r pca-quant-plot}
ggplot(distance_results, aes(x = comparison, y = distance, fill = analysis)) +
  geom_boxplot()
```

## Session information

```{r info}
sessionInfo()
```
