---
title: "Quality control for iPSC cells"
date: 2015-05-27
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

## Input

```{r packages, message=FALSE}
library("dplyr")
library("ggplot2")
```

Summary counts from featureCounts.
Created with [gather-summary-counts.py](https://github.com/jdblischak/singleCellSeq/blob/master/code/gather-summary-counts.py).

```{r input-summary-counts}
summary_counts <- read.table("../data/summary-counts.txt", header = TRUE,
                             stringsAsFactors = FALSE)
```

Using only the sickle-trimmed data, sum the counts across all the sequencing lanes for a given sample.

```{r sum-per-sample}
summary_per_sample <- summary_counts %>%
  filter(sickle == "quality-trimmed") %>%
  select(-index, -lane, -flow_cell, -sickle) %>%
  group_by(individual, batch, well, rmdup) %>%
  summarise_each(funs(sum)) %>%
  ungroup %>%
  as.data.frame
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input read counts.

```{r input-read-counts}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input single cell observational quality control data.

```{r input-qc}
qc <- read.table("../data/qc-ipsc.txt", header = TRUE,
                 stringsAsFactors = FALSE)
head(qc)
```

## Quality control at cell level

Looking at the unmapped ratio and ERCC ratio of each cell.

1. low number of mapped reads might be caused by cell apoptosis, RNA degradation, insufficient cell lysis, bad Tn5 transposase quality, or contamination
2. high proportion of ERCC spike-in indicates low RNA quantity from iPSCs which might cell death

```{r reads}
#reads per sample
summary_per_sample_reads <- summary_per_sample %>% filter(rmdup == "reads")

#create unmapped ratios
summary_per_sample_reads$unmapped.ratios <- summary_per_sample_reads[,9]/apply(summary_per_sample_reads[,5:13],1,sum)

#create total mapped reads
summary_per_sample_reads$total.mapped <- apply(summary_per_sample_reads[,5:8],1,sum)

#plot
ggplot(summary_per_sample_reads, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), shape = as.factor(batch))) + geom_point(size = 3, alpha = 0.5)

#plot the sum of reads and 'Assigned'
plot(apply(reads,2,sum),summary_per_sample_reads[,5])

#creat ERCC ratios
summary_per_sample_reads$ERCC.ratios <- apply(reads[grep("ERCC", rownames(reads)), ],2,sum)/apply(summary_per_sample_reads[,5:8],1,sum)

#plot
ggplot(summary_per_sample_reads, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), shape = as.factor(batch))) + geom_point(size = 3, alpha = 0.5)

#remove bulk keep single cell
summary_per_sample_reads_single <- summary_per_sample_reads[summary_per_sample_reads$well!="bulk",]

ggplot(summary_per_sample_reads_single, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), shape = as.factor(batch))) + geom_point(size = 3, alpha = 0.5)

ggplot(summary_per_sample_reads_single, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), shape = as.factor(batch))) + geom_point(size = 3, alpha = 0.5)

ggplot(summary_per_sample_reads_single, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), shape = as.factor(batch))) + geom_point(size = 3, alpha = 0.5) + facet_grid(individual ~ batch)

ggplot(summary_per_sample_reads_single, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), shape = as.factor(batch))) + geom_point(size = 3, alpha = 0.5) + facet_grid(individual ~ batch)

```

QC by cell number, unmapped ratio, and ERCC ratio
```{r cell-qc}
#add cell number per well by merging qc file
summary_per_sample_reads_single_qc <- merge(summary_per_sample_reads_single,qc,by=c("individual","batch","well"))

ggplot(summary_per_sample_reads_single_qc, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

ggplot(summary_per_sample_reads_single_qc, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

ggplot(summary_per_sample_reads_single_qc, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3) + facet_grid(individual ~ batch)

ggplot(summary_per_sample_reads_single_qc, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3) + facet_grid(individual ~ batch)

```

Given that each plates/batch was collected seperately and independently, it would be reasonable to remove the bad quality cell by batch (using different cutoffs for each C1 collection). For exmaple, 19098 batch 2 have more ERCC than all the others. Using the same quiteria would end up with no cell.

```{r by-C1}
#individual 19098 batch1
C1_19098_1 <- summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$individual == "19098" & summary_per_sample_reads_single_qc$batch =="1",]

ggplot(C1_19098_1, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

ggplot(C1_19098_1, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

C1_19098_1_retained <- C1_19098_1[C1_19098_1$unmapped.ratios< 0.4 & C1_19098_1$ERCC.ratios< 0.025 & C1_19098_1$cell_number == 1,]

dim(C1_19098_1_retained)

ggplot(C1_19098_1_retained, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

#individual 19098 batch2
C1_19098_2 <- summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$individual == "19098" & summary_per_sample_reads_single_qc$batch =="2",]

ggplot(C1_19098_2, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

ggplot(C1_19098_2, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

C1_19098_2_retained <- C1_19098_2[C1_19098_2$unmapped.ratios< 0.6 & C1_19098_2$ERCC.ratios< 0.35 & C1_19098_2$cell_number == 1,]

dim(C1_19098_2_retained)

ggplot(C1_19098_2_retained, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

#individual 19098 batch3
C1_19098_3 <- summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$individual == "19098" & summary_per_sample_reads_single_qc$batch =="3",]

ggplot(C1_19098_3, aes(x = total.mapped, y = unmapped.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

ggplot(C1_19098_3, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)

C1_19098_3_retained <- C1_19098_3[C1_19098_3$unmapped.ratios< 0.45 & C1_19098_3$ERCC.ratios< 0.08 & C1_19098_3$cell_number == 1,]

dim(C1_19098_3_retained)

ggplot(C1_19098_3_retained, aes(x = total.mapped, y = ERCC.ratios, col = as.factor(individual), label = as.character(cell_number))) + geom_text(fontface=3)
```

## Effect of cell QC data on sequencing results

### Cell number

```{r cell-num-counts}
table(summary_per_sample_reads_single_qc$cell_number)
```

```{r cell-num-ercc}
ggplot(summary_per_sample_reads_single_qc,
       aes(x = as.factor(cell_number), y = ERCC.ratios)) +
  geom_boxplot()
```

```{r cell-num-unmapped}
ggplot(summary_per_sample_reads_single_qc,
       aes(x = as.factor(cell_number), y = unmapped.ratios)) +
  geom_boxplot()
```

### Concentration

```{r concentration-sum}
summary(summary_per_sample_reads_single_qc$concentration)
```

```{r conc-ercc}
ggplot(summary_per_sample_reads_single_qc,
       aes(x = concentration, y = ERCC.ratios)) +
  geom_point() +
  geom_smooth()
```

```{r conc-unmapped}
ggplot(summary_per_sample_reads_single_qc,
       aes(x = concentration, y = unmapped.ratios)) +
  geom_point() +
  geom_smooth()
```

### Tra-1-60

Tra-1-60 is a [stem cell marker][marker].

[marker]: https://en.wikipedia.org/wiki/Stem_cell_marker

```{r tra-sum}
table(summary_per_sample_reads_single_qc$tra1.60)
```

```{r tra-ercc}
ggplot(summary_per_sample_reads_single_qc,
       aes(x = as.factor(tra1.60), y = ERCC.ratios)) +
  geom_boxplot()
```

```{r tra-unmapped}
ggplot(summary_per_sample_reads_single_qc,
       aes(x = as.factor(tra1.60), y = unmapped.ratios)) +
  geom_boxplot()
```

### Filtering by cell number and Tra-1-60

Simply removing all the cells with bad cell number or tra-1-60 does not remove all the low quality samples.

```{r cell-num-tra-sum}
table(summary_per_sample_reads_single_qc$cell_number,
      summary_per_sample_reads_single_qc$tra1.60)
```

```{r qc-filtered-by-cell-num-tra}
ggplot(subset(summary_per_sample_reads_single_qc, cell_number == 1 & tra1.60 == 1),
       aes(x = total.mapped, y = ERCC.ratios)) +
  geom_point()
ggplot(subset(summary_per_sample_reads_single_qc, cell_number == 1 & tra1.60 == 1),
       aes(x = total.mapped, y = unmapped.ratios)) +
  geom_point()
```

## Session information

```{r info}
sessionInfo()
```
