---
title: "Quality control for iPSC cells"
date: 2015-05-27
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

## Input

```{r packages}
library("dplyr")
```

Summary counts from featureCounts.
Created with [gather-summary-counts.py](https://github.com/jdblischak/singleCellSeq/blob/master/code/gather-summary-counts.py).

```{r input-summary-counts}
summary_counts <- read.table("../data/summary-counts.txt", header = TRUE,
                             stringsAsFactors = FALSE)
```

Using only the sickle-trimmed data, sum the counts across all the sequencing lanes for a given sample.

```{r sum-per-sample}
summary_per_sample <- summary_counts %>%
  filter(sickle == "quality-trimmed") %>%
  select(-index, -lane, -flow_cell, -sickle) %>%
  group_by(individual, batch, well, rmdup) %>%
  summarise_each(funs(sum)) %>%
  ungroup %>%
  as.data.frame
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input read counts.

```{r input-read-counts}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input single cell observational quality control data.

```{r input-qc}
qc <- read.table("../data/qc-ipsc.txt", header = TRUE,
                 stringsAsFactors = FALSE)
head(qc)
```

```{r}
#reads per sample
summary_per_sample_reads <- summary_per_sample %>% filter(rmdup == "reads")

#create unmapped ratios
summary_per_sample_reads$unmapped.ratios <- summary_per_sample_reads[,9]/apply(summary_per_sample_reads[,5:13],1,sum)

#create total mapped reads
summary_per_sample_reads$total.mapped <- apply(summary_per_sample_reads[,5:8],1,sum)

#creat ERCC ratios
ERCC.ratios <- apply(reads[grep("ERCC", rownames(reads)), ],2,sum)/apply(summary_per_sample_reads[,5:8],1,sum)

#how come the sum of reads are larger than 'Assigned'
(apply(reads,2,sum)/summary_per_sample_reads[,5])
#plot the sum of reads and 'Assigned'
plot(apply(reads,2,sum),summary_per_sample_reads[,5])
```



## Session information

```{r info}
sessionInfo()
```
