---
title: "QC of single cell libraries"
author: "PoYuan Tung"
date: 2015-10-23
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(fig.width = 8, fig.height = 8)
```

## Input

```{r packages, message=FALSE}
library("dplyr")
library("edgeR")
library("ggplot2")
library("cowplot")
theme_set(theme_bw(base_size = 16))
source("functions.R")
```

Summary counts from featureCounts.
Created with [gather-summary-counts.py](https://github.com/jdblischak/singleCellSeq/blob/master/code/gather-summary-counts.py).
These data were collected from the summary files of the full combined samples.

```{r input-summary-counts}
summary_counts <- read.table("../data/summary-counts.txt", header = TRUE,
                             stringsAsFactors = FALSE)
```

Currently this file only contains data from sickle-trimmed reads, so the code below simply ensures this and then removes the column.

```{r clean-summary-counts}
summary_per_sample <- summary_counts %>%
  filter(sickle == "quality-trimmed") %>%
  dplyr :: select(-sickle) %>%
  arrange(individual, batch, well, rmdup) %>%
  as.data.frame

summary_per_sample_single <- summary_per_sample[summary_per_sample$well != "bulk", ]
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
anno_single <- anno[anno$well != "bulk",]
```

Input read counts.

```{r input-read-counts}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input single cell observational quality control data.

```{r input-qc}
qc <- read.table("../data/qc-ipsc.txt", header = TRUE,
                 stringsAsFactors = FALSE)
head(qc)
```

## Total ERCC and removal of 19098.r2

Show the evidence that removing 19098 batch 2 is the first thing to do.

```{r sad-ercc}
summary_per_sample_single_reads <- summary_per_sample_single[summary_per_sample_single$rmdup == "reads",]

summary_per_sample_single_reads$sample_id <- anno_single$sample_id
summary_per_sample_single_reads$batch_9 <- anno_single$batch
stopifnot(colnames(reads) == summary_per_sample_single_reads$sample_id )

summary_per_sample_single_reads$ERCC_reads <- apply(reads[grep("ERCC", rownames(reads)), ],2,sum)
summary_per_sample_single_reads$ERCC_molecules <- apply(molecules[grep("ERCC", rownames(molecules)), ],2,sum)
```

```{r sad-ercc-plot, fig.width=12, fig.height=8}
## create a color palette with one color per individual and different shades for repplicates
great_color <- c("#CC3300", "#FF9966", "#FFCC99", "#006633", "#009900", "#99FF99", "#3366FF", "#6699FF", "#66CCFF")
great_color_8 <- c("#CC3300", "#FF9966", "#006633", "#009900", "#99FF99", "#3366FF", "#6699FF", "#66CCFF")

ercc_reads_plot <- ggplot(summary_per_sample_single_reads,
                   aes(x = factor(batch_9), y = ERCC_reads,
                   fill = factor(batch_9)), height = 600, width = 2000) +
                   geom_violin(alpha = .5) + 
                   geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
                   scale_fill_manual(values = great_color) + 
                   labs(x = "batch", y = "ERCC reads", title = "Total ERCC read counts per cell") +
                   theme(axis.text.x = element_text(hjust=1, angle = 45))

ercc_molecule_plot <- ggplot(summary_per_sample_single_reads,
                   aes(x = factor(batch_9), y = ERCC_molecules,
                   fill = factor(batch_9)), height = 600, width = 2000) +
                   geom_violin(alpha = .5) + 
                   geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
                   scale_fill_manual(values = great_color) + 
                   labs(x = "batch", y = "ERCC molecules", title = "Total ERCC molecule counts per cell") +
                   theme(axis.text.x = element_text(hjust=1, angle = 45))

theme_set(theme_bw(base_size = 8))
plot_grid(ercc_reads_plot + theme(legend.position = "none"),
          ercc_molecule_plot + theme(legend.position = "none"),
          labels = LETTERS[1:2])
```

Remove 19098r2 for all the following analysis
```{r remove-19098r2}
remove_19098r2 <- anno_single$batch != "NA19098.r2"
anno_single_rm <- anno_single[remove_19098r2,]
summary_per_sample_single_reads_rm <- summary_per_sample_single_reads[remove_19098r2,]
reads_rm <- reads[, remove_19098r2]
molecules_rm <- molecules[, remove_19098r2]

stopifnot(summary_per_sample_single_reads_rm$sample_id == colnames(reads_rm))
```

## Total mapped reads reads

```{r total reads}
## add cell number per well by merging qc file
summary_per_sample_reads_single_qc <- merge(summary_per_sample_single_reads_rm,qc,by=c("individual","batch","well"))

## calculate total mapped reads per sample
summary_per_sample_reads_single_qc$total_mapped <- apply(summary_per_sample_reads_single_qc[,5:8],1,sum)

## cut off 
cut_off_reads <- quantile(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0,"total_mapped"], 0.95)

cut_off_reads

summary_per_sample_reads_single_qc$cut_off_reads <- summary_per_sample_reads_single_qc$total_mapped > cut_off_reads

## numbers of cells 
sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "total_mapped"] > cut_off_reads)

sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "total_mapped"] <= cut_off_reads)

## density plots
plot_reads <- ggplot(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0 |
                                           summary_per_sample_reads_single_qc$cell_number == 1 , ],
       aes(x = total_mapped, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_reads, colour="grey", linetype = "longdash") +
       labs(x = "total mapped reads", title = "Total mappred reads")
```

## Unmapped ratios
```{r unmapped ratios}
## calculate unmapped ratios
summary_per_sample_reads_single_qc$unmapped_ratios <- summary_per_sample_reads_single_qc[,9]/apply(summary_per_sample_reads_single_qc[,5:13],1,sum)

## cut off 
cut_off_unmapped <- quantile(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0,"unmapped_ratios"], 0.05)

cut_off_unmapped

summary_per_sample_reads_single_qc$cut_off_unmapped <- summary_per_sample_reads_single_qc$unmapped_ratios < cut_off_unmapped

## numbers of cells 
sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "unmapped_ratios"] >= cut_off_unmapped)

sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "unmapped_ratios"] < cut_off_unmapped)

## density plots
plot_unmapped <- ggplot(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0 |
                                           summary_per_sample_reads_single_qc$cell_number == 1 , ],
       aes(x = unmapped_ratios *100, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_unmapped *100, colour="grey", linetype = "longdash") +
       labs(x = "unmapped reads/ total reads", title = "Unmapped reads percentage")
```

## ERCC percentage 
```{r ercc-percentage}
## calculate ercc reads percentage
summary_per_sample_reads_single_qc$ercc_percentage <- apply(reads_rm[grep("ERCC", rownames(reads_rm)), ],2,sum)/apply(summary_per_sample_reads_single_qc[,5:8],1,sum) 

## cut off 
cut_off_ercc <- quantile(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0,"ercc_percentage"], 0.05)

cut_off_ercc

summary_per_sample_reads_single_qc$cut_off_ercc <- summary_per_sample_reads_single_qc$ercc_percentage < cut_off_ercc

## numbers of cells 
sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "ercc_percentage"] >= cut_off_ercc)

sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "ercc_percentage"] < cut_off_ercc)

## density plots
plot_ercc <- ggplot(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0 |
                                           summary_per_sample_reads_single_qc$cell_number == 1 , ],
       aes(x = ercc_percentage *100, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_ercc *100, colour="grey", linetype = "longdash") +
       labs(x = "ERCC reads / total mapped reads", title = "ERCC reads percentage")
```

## Number of genes detected 
```{r gene-number}
## endogenous genes
reads_rm_gene <- reads_rm[grep("ENSG", rownames(reads_rm)), ]

## number of genes detected 
summary_per_sample_reads_single_qc$gene_number <- colSums(reads_rm_gene >= 1)

## cut off 
cut_off_genes <- quantile(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0,"gene_number"], 0.95)

cut_off_genes

summary_per_sample_reads_single_qc$cut_off_genes <- summary_per_sample_reads_single_qc$gene_number > cut_off_genes

## numbers of cells 
sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "gene_number"] > cut_off_genes)

sum(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1, "gene_number"] <= cut_off_genes)

## density plots
plot_gene <- ggplot(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 0 |
                                           summary_per_sample_reads_single_qc$cell_number == 1 , ],
       aes(x = gene_number, fill = as.factor(cell_number))) + 
       geom_density(alpha = 0.5) +
       geom_vline(xintercept = cut_off_genes, colour="grey", linetype = "longdash") +
       labs(x = "gene numbers", title = "Numbers of detected genes")
```

```{r density plots}
theme_set(theme_bw(base_size = 10))
plot_grid(plot_reads + theme(legend.position = "none"), 
          plot_unmapped + theme(legend.position = "none"),
          plot_ercc + theme(legend.position = "none"), 
          plot_gene + theme(legend.position = "none"),
          labels = LETTERS[1:4])
```

## Total molecule counts
```{r total-mol}
## calculate total gene molecule counts
summary_per_sample_reads_single_qc$total_gene_molecule <- colSums(molecules_rm[grep("ENSG", rownames(molecules_rm)),])

## look for outiers
ggplot(summary_per_sample_reads_single_qc, aes(x = concentration, y = total_gene_molecule / 10^3,
  color = as.factor(cell_number))) +
  geom_text(aes(label = cell_number)) +
  labs(x = "Concentration", y = "Gene molecules (thousands)") +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")

outliers <- summary_per_sample_reads_single_qc %>% filter(cell_number == 1, concentration < 1.25, concentration > .15,
                                   total_gene_molecule > 140000)

outliers %>% dplyr::select(sample_id)

summary_per_sample_reads_single_qc$molecule_outlier <- summary_per_sample_reads_single_qc$cell_number == 1 &
  summary_per_sample_reads_single_qc$concentration < 1.25 &
  summary_per_sample_reads_single_qc$concentration > .15 &
  summary_per_sample_reads_single_qc$total_gene_molecule > 140000
```

### Linear Discriminat Analysis
```{r lda}
library(MASS)
## create 3 groups according to cell number
group_3 <- rep("two",dim(summary_per_sample_reads_single_qc)[1])
         group_3[grep("0", summary_per_sample_reads_single_qc$cell_number)] <- "no"
         group_3[grep("1", summary_per_sample_reads_single_qc$cell_number)] <- "one"

## create data frame
data_lda <- data.frame(anno_single_rm,
                       cell_number = summary_per_sample_reads_single_qc$cell_number,
                       concentration = summary_per_sample_reads_single_qc$concentration,
                       total_gene_molecule = summary_per_sample_reads_single_qc$total_gene_molecule,
                       group = group_3)

## remove 19098.r1
data_con <- data_lda %>% filter(batch != "NA19098.r1")
plot_before <- ggplot(data_con, aes(x = concentration, y = total_gene_molecule / 10^3,
               color = as.factor(group))) +
               geom_text(aes(label = cell_number)) +
               labs(x = "Concentration", y = "Gene molecules (thousands)") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")

## perform lda
data_con_lda <- lda(group ~ concentration + total_gene_molecule,
                    data = data_con)
data_con_lda_p <- predict(data_con_lda, 
                          newdata = data_con[,c("concentration", "total_gene_molecule")])$class

## determine how well the model fix
table(data_con_lda_p, data_con[, "group"])
data_con$data_con_lda_p <- data_con_lda_p

plot_after <- ggplot(data_con, aes(x = concentration, y = total_gene_molecule / 10^3,
               color = as.factor(data_con_lda_p))) +
               geom_text(aes(label = cell_number)) +
               labs(x = "Concentration", y = "Gene molecules (thousands)") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")

## identify the outlier 
outliers_lda <- data_con %>% filter(cell_number == 1, data_con_lda_p == "two")
outliers_lda$sample_id

## the results are the same!!
stopifnot(outliers %>% dplyr::select(sample_id) == outliers_lda$sample_id)

theme_set(theme_bw(base_size = 10))
plot_grid(plot_before + theme(legend.position = "none"), 
          plot_after + theme(legend.position = "none"),
          labels = LETTERS[1:2])
```

## Reads to molecule conversion
```{r convertion}
## calculate convertion
summary_per_sample_reads_single_qc$ERCC_conversion <- summary_per_sample_reads_single_qc$ERCC_molecules / summary_per_sample_reads_single_qc$ERCC_reads

summary_per_sample_reads_single_qc$conversion <- summary_per_sample_reads_single_qc$total_gene_molecule /  colSums(reads_rm[grep("ENSG", rownames(reads_rm)),])

ggplot(summary_per_sample_reads_single_qc, aes(x = ERCC_conversion, y = conversion,
  color = as.factor(cell_number))) +
  geom_text(aes(label = cell_number)) +
  labs(x = "Convertion of ERCC", y = "Conversion of genes") +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")

out_ercc_con <- summary_per_sample_reads_single_qc %>% filter(cell_number == "1", ERCC_conversion > .094)

## try lda
data_lda$conversion <- summary_per_sample_reads_single_qc$conversion
data_ercc_lda <- lda(group ~ ERCC_conversion + conversion,
                    data = data_lda)
data_ercc_lda_p <- predict(data_ercc_lda, 
                          newdata = data_lda[,c("ERCC_conversion", "conversion")])$class
table(data_con_lda_p, data_con[, "group"])

data_lda$data_ercc_lda_p <- data_ercc_lda_p

## identify the outlier 
outliers_ercc <- data_lda %>% filter(cell_number == 1, data_ercc_lda_p == "two")
outliers_ercc$sample_id

## cutoff
out_ercc_con <- summary_per_sample_reads_single_qc %>% filter(cell_number == "1", ERCC_conversion > .094)
stopifnot(out_ercc_con %>% dplyr::select(sample_id) == outliers_ercc$sample_id)

summary_per_sample_reads_single_qc$conversion_outlier <- summary_per_sample_reads_single_qc$cell_number == 1 &
                                                         summary_per_sample_reads_single_qc$ERCC_conversion > .094

plot_ercc_before <- ggplot(data_lda, aes(x = ERCC_conversion, y = conversion,
               color = as.factor(group))) +
               geom_text(aes(label = cell_number)) +
               labs(x = "Convertion of ERCC", y = "Conversion of genes") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")

plot_ercc_after <- ggplot(data_lda, aes(x = ERCC_conversion, y = conversion,
               color = as.factor(data_ercc_lda_p))) +
               geom_text(aes(label = cell_number)) +
               labs(x = "Convertion of ERCC", y = "Conversion of genes") +
               scale_color_brewer(palette = "Dark2") +
               theme(legend.position = "none")

theme_set(theme_bw(base_size = 10))
plot_grid(plot_ercc_before + theme(legend.position = "none"), 
          plot_ercc_after + theme(legend.position = "none"),
          labels = LETTERS[1:2])
```

## Mitochondrial genes
```{r mito-gene}
## create a list of mitochondrial genes (13 protein-coding genes)
## MT-ATP6, MT-CYB, MT-ND1, MT-ND4, MT-ND4L, MT-ND5, MT-ND6, MT-CO2, MT-CO1, MT-ND2, MT-ATP8, MT-CO3, MT-ND3
mtgene <- c("ENSG00000198899", "ENSG00000198727", "ENSG00000198888", "ENSG00000198886", "ENSG00000212907", "ENSG00000198786", "ENSG00000198695", "ENSG00000198712", "ENSG00000198804", "ENSG00000198763","ENSG00000228253", "ENSG00000198938", "ENSG00000198840")

## reads of mt genes in single cells
mt_reads <- reads_rm_gene[mtgene,]
dim(mt_reads)
stopifnot(colnames(reads_rm) == rownames(summary_per_sample_reads_single_qc$sample_id))

## mt ratio of single cell
summary_per_sample_reads_single_qc$mt_reads <- apply(mt_reads, 2, sum)
summary_per_sample_reads_single_qc$mt_reads_ratio <- summary_per_sample_reads_single_qc$mt_reads /summary_per_sample_reads_single_qc$total_mapped

## vs. number of genes detected
ggplot(summary_per_sample_reads_single_qc,
       aes(x = gene_number, y = mt_reads_ratio, 
       color = as.factor(cell_number))) +
       geom_text(aes(label = cell_number)) +
       labs(x = "Number of genes", y = "Mitochondrial ratio") +
       scale_color_brewer(palette = "Dark2") +
       theme(legend.position = "none")
```

## Filter

### Final list
```{r filter}
## all filter
summary_per_sample_reads_single_qc$filter_all <- summary_per_sample_reads_single_qc$cell_number == 1 &
                                                 summary_per_sample_reads_single_qc$cut_off_reads &
                                                 summary_per_sample_reads_single_qc$cut_off_unmapped &
                                                 summary_per_sample_reads_single_qc$cut_off_ercc &
                                                 summary_per_sample_reads_single_qc$cut_off_genes &
                                                 summary_per_sample_reads_single_qc$molecule_outlier == "FALSE" &
                                                 summary_per_sample_reads_single_qc$conversion_outlier == "FALSE"
                                                 
table(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$filter_all,
                           c("individual", "batch")])    
```

```{r output}
stopifnot(nrow(summary_per_sample_reads_single_qc) == nrow(anno_single_rm))
quality_single_cells <- anno_single_rm %>%
  filter(summary_per_sample_reads_single_qc$filter_all) %>%
  dplyr :: select(sample_id)

write.table(quality_single_cells,
            file = "../data/quality-single-cells.txt", quote = FALSE,
            sep = "\t", row.names = FALSE, col.names = FALSE)

```

### Mito ratios
```{r mito-filter}
ggplot(summary_per_sample_reads_single_qc,
       aes(x = gene_number, y = mt_reads_ratio, 
       color = as.factor(filter_all))) +
       geom_text(aes(label = cell_number)) +
       labs(x = "Number of genes", y = "Mitochondrial ratio") +
       theme(legend.position = "none")

ggplot(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$cell_number == 1,],
       aes(x = factor(filter_all), y = mt_reads_ratio,
           fill = factor(filter_all)), height = 600, width = 2000) +
geom_violin(alpha = .5) + 
geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
labs(x = "Quality", y = "Mitochonrial ratio", title = "Mitochondrial ratio of libraries with 1 cell")

## check the batch of those outliers
mito_outliers <- summary_per_sample_reads_single_qc %>% filter(filter_all == "TRUE", mt_reads_ratio > .15)
mito_outliers  %>% dplyr::select(sample_id, mt_reads_ratio)

## check if 19098 have high mt genes
ggplot(summary_per_sample_reads_single_qc[summary_per_sample_reads_single_qc$filter_all == "TRUE",],
                   aes(x = factor(batch_9), y = mt_reads_ratio,
                   fill = factor(batch_9)), height = 600, width = 2000) +
                   geom_violin(alpha = .5) + 
                   geom_boxplot(alpha = .01, width = .2, position = position_dodge(width = .9)) +
                   scale_fill_manual(values = great_color_8) +  
                   labs(x = "batch", y = "Mitochondrial ratio") +
                   theme(axis.text.x = element_text(hjust=1, angle = 45))
```

### plots
```{r plots}
## for figure 1
genes_unmapped <-  ggplot(summary_per_sample_reads_single_qc,
                   aes(x = gene_number, y = unmapped_ratios * 100,
                   col = as.factor(individual), shape = as.factor(batch), height = 600, width = 2000)) +
                   geom_point(size = 3, alpha = 0.3) + 
                   geom_vline(xintercept = cut_off_genes, colour="grey", linetype = "longdash") +
                   geom_hline(yintercept = cut_off_unmapped * 100, colour="grey", linetype = "longdash") +
                   labs(x = "Number of genes", y = "Unmapped reads percentage (%)") 

genes_ercc <-  ggplot(summary_per_sample_reads_single_qc,
                   aes(x = gene_number, y = ercc_percentage * 100,
                   col = as.factor(individual), shape = as.factor(batch), height = 600, width = 2000)) +
                   geom_point(size = 3, alpha = 0.3) +
                   geom_vline(xintercept = cut_off_genes, colour="grey", linetype = "longdash") +
                   geom_hline(yintercept = cut_off_ercc * 100, colour="grey", linetype = "longdash") +
                   labs(x = "Number of genes", y = "ERCC reads percentage (%)") 

reads_unmapped_num <-  ggplot(summary_per_sample_reads_single_qc,
                   aes(x = total_mapped, y = unmapped_ratios * 100,
                   col = as.factor(individual), label = as.character(cell_number), height = 600, width = 2000)) +
                   geom_text(fontface = 3, alpha = 0.5) + 
                   geom_vline(xintercept = cut_off_reads, colour="grey", linetype = "longdash") +
                   geom_hline(yintercept = cut_off_unmapped * 100, colour="grey", linetype = "longdash") +
                   labs(x = "Total mapped reads", y = "Unmapped reads percentage (%)") 

reads_ercc_num <-  ggplot(summary_per_sample_reads_single_qc,
                   aes(x = total_mapped, y = ercc_percentage * 100,
                   col = as.factor(individual), label = as.character(cell_number), height = 600, width = 2000)) +
                   geom_text(fontface = 3, alpha = 0.5) + 
                   geom_vline(xintercept = cut_off_reads, colour="grey", linetype = "longdash") +
                   geom_hline(yintercept = cut_off_ercc * 100, colour="grey", linetype = "longdash") +
                   labs(x = "Total mapped reads", y = "ERCC reads percentage (%)") 

theme_set(theme_bw(base_size = 10))
plot_grid(genes_unmapped + theme(legend.position = "none"), 
          genes_ercc + theme(legend.position = "none"),
          reads_unmapped_num + theme(legend.position = "none"), 
          reads_ercc_num + theme(legend.position = "none"),
          labels = LETTERS[1:4])
```


## Session information

```{r info}
sessionInfo()
```

