---
title: "Subsample: Cell to cell variance in expression"
date: 2015-07-16
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
opts_chunk$set(warning = FALSE)
```

How does cell to cell variance in gene expression varies with subsampling both sequencing depth and number of cells?
To do this I calculated the variance for each gene using all the single cells available for a given individual at a given sequencing depth and then compared these to the estimates from a random subsample of cells.
I log-tranformed the values (plus a pseudocount of 0.25 to avoid dividing by zero) and then calculated the Pearson's r and the [root-mean-square error](https://en.wikipedia.org/wiki/Root-mean-square_deviation) (RMSE).

```{r message=FALSE}
library("dplyr")
library("ggplot2")
theme_set(theme_bw(base_size = 14))
```

## Batch process each subsampled data set

Run 10 iterations for each individual for each sequencing depth for each subsample of cells.
The analysis is performed by [subsample-variance.R](https://github.com/jdblischak/singleCellSeq/blob/master/code/subsample-variance.R).

```bash
cd $ssd/subsampled
mkdir -p variance
mkdir -p ~/log/subsample-variance.R
for IND in 19098 19101 19239
do
  for NUM in 200000 400000 1000000 2000000 3000000 4000000
  do
    for CELLS in 5 10 15 20 25 50 75 100 125 150
    do
      for SEED in {1..10}
      do
        # Molecules
        CMD="subsample-variance.R $CELLS $SEED molecule-counts-$NUM.txt --individual=$IND --min_count=1 --min_cells=5 --good_cells=/mnt/lustre/home/jdblischak/singleCellSeq/data/quality-single-cells.txt"
        DEST="variance/molecule-$IND-$CELLS-$SEED-$NUM.txt"
        echo "$CMD > $DEST" | qsub -l h_vmem=2g -cwd -V -N variance-molecule-$IND-$CELLS-$SEED-$NUM -j y -o ~/log/subsample-variance.R -l 'hostname=!bigmem01'
        sleep .01s
        # Reads
        CMD="subsample-variance.R $CELLS $SEED read-counts-$NUM.txt --individual=$IND --min_count=10 --min_cells=5 --good_cells=/mnt/lustre/home/jdblischak/singleCellSeq/data/quality-single-cells.txt"
        DEST="variance/read-$IND-$CELLS-$SEED-$NUM.txt"
        echo "$CMD > $DEST" | qsub -l h_vmem=2g -cwd -V -N variance-read-$IND-$CELLS-$SEED-$NUM -j y -o ~/log/subsample-variance.R -l 'hostname=!bigmem01'
        sleep .01s
      done
    done
  done
done
```

Convert to one file using Python.
Run from `$ssd/subsampled`.

```python
import os
import glob
files = glob.glob("variance/*txt")
out = open("variance.txt", "w")
out.write("type\tind\tdepth\tnum_cells\tseed\ttotal_cells\tr\trmse\n")
for fname in files:
    fname_parts = os.path.basename(fname).rstrip(".txt").split("-")
    type = fname_parts[0]
    ind = fname_parts[1]
    depth = fname_parts[4]
    f = open(fname, "r")
    out.write(type + "\t" + ind + "\t" + depth + "\t" + f.read())
    f.close()

out.close()
```

## Input data

```{r input}
var_data <- read.table("/mnt/gluster/data/internal_supp/singleCellSeq/subsampled/variance.txt",
                         header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

Calculate the mean and standard error of the mean (sem) for each of the 10 iterations.

```{r calculate-mean-and-sem}
var_data_plot <- var_data %>%
  group_by(type, ind, depth, num_cells, total_cells) %>%
  summarize(r_mean = mean(r), r_sem = sd(r) / sqrt(length(r)),
            rmse_mean = mean(rmse), rmse_sem = sd(rmse) / sqrt(length(rmse)))
```

For the analysis of read counts, a gene was included if it had greater than 10 reads in at least 5 cells.
For the analysis of molecule counts, a gene was included if it had greater than 1 molecule in at least 5 cells.

## Pearson's r

```{r subsample-variance-r, fig.width=10, fig.height=10}
p_r <- ggplot(var_data_plot, aes(x = num_cells, y = r_mean, color = as.factor(depth))) +
  geom_line() +
  geom_errorbar(aes(ymin = r_mean - r_sem, ymax = r_mean + r_sem), width = 1) +
  facet_grid(type~ind) +
  labs(x = "Number of subsampled cells",
       y = "Pearson's r",
       color = "Depth",
       title = "Subsample: Correlation of variance estimates")
p_r
```

## Root-mean-square error

```{r subsample-variance-rmse, fig.width=10, fig.height=10}
p_rmse <- ggplot(var_data_plot, aes(x = num_cells, y = rmse_mean, color = as.factor(depth))) +
  geom_line() +
  geom_errorbar(aes(ymin = rmse_mean - rmse_sem, ymax = rmse_mean + rmse_sem), width = 1) +
  facet_grid(type~ind) +
  labs(x = "Number of subsampled cells",
       y = "Root-mean-square error",
       color = "Depth",
       title = "Subsample: RMSE of variance estimates")
p_rmse
```

## Effect of read depth on total number of cells

One thing to keep in mind is that the total number of cells used to calculate the "true" variance varies by sequencing depth.

```{r total-cells-by-depth}
ggplot(var_data_plot, aes(x = depth, y = total_cells)) +
  geom_point() +
  facet_grid(~ind) +
  labs(x = "Sequencing depth",
       y = "Total number of cells",
       title = "Effect of read depth on total number of cells")
```

Also need to keep in mind that the [number of genes included changes due to sequencing depth and number of cells subsampled](detect-genes.html).

## Session information

```{r info}
sessionInfo()
```
