---
output:
  html_document:
    toc: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

At lower number of sampled cells, reads appear to perform better with lower sequencing.
On the other hand, the molecules look as expected, higher sequencing depth leads to increased or similar performance.
This is observed both for endogenous and ERCC genes.
Maybe there is less bias with fewer reads?
I re-did this analysis with more extreme differences in sequencing depth, from a low of 50,000 to a high of 4 million.

The key is separating the low and high expressed genes.
The reads do a poor job of capturing the lowly expressed genes, so with more sequencing, this relationship gets less linear.
This is why Spearman correlation works better for the reads.
The molecules are less affected because they remove the amplification bias of the highly expressed genes.

## Setup

I haven't correctly formatted the output for when there are not enough cells to be subsampled, so I removed these.

```
$ cat subsampling-results.txt | awk '$10 >= 0' > subsampling-results8.txt
```

```{r packages, message=FALSE}
library("knitr")
opts_chunk$set(fig.width = 8, fig.height = 8, fig.pos = "center")
library("dplyr")
library("ggplot2")
theme_set(theme_bw())
```

```{r}
d <- read.table("/mnt/gluster/home/jdblischak/ssd/subsampled/subsampling-results8.txt",
                header = TRUE, sep = "\t", stringsAsFactors = FALSE)
head(d)
summary(d)
```

```{r}
d_grouped <- d %>%
  group_by(type, depth, gene_subset, subsampled_cells,
           individual, potential_cells, available_cells,
           lower_q, upper_q, available_ensg, used_ensg,
           available_ercc, used_ercc) %>%
  summarize(mean_pearson_ensg = mean(pearson_ensg),
            sem_pearson_ensg = sd(pearson_ensg) / sqrt(length(pearson_ensg)),
            mean_pearson_ercc = mean(pearson_ercc),
            sem_pearson_ercc = sd(pearson_ercc) / sqrt(length(pearson_ercc)),
            mean_spearman_ensg = mean(spearman_ensg),
            sem_spearman_ensg = sd(spearman_ensg) / sqrt(length(spearman_ensg)),
            mean_spearman_ercc = mean(spearman_ercc),
            sem_spearman_ercc = sd(spearman_ercc) / sqrt(length(spearman_ercc)))
head(d_grouped)
dim(d_grouped)
```

## Endogenous Pearson NA19098

```{r}
ensg_pearson_plot <- ggplot(d_grouped[d_grouped$individual == "NA19098", ],
                 aes(x = subsampled_cells, y = mean_pearson_ensg,
                     color = as.factor(depth))) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_pearson_ensg - sem_pearson_ensg,
                    ymax = mean_pearson_ensg + sem_pearson_ensg),
                width = 1) +
  facet_grid(gene_subset~type)
ensg_pearson_plot
```

## Endogenous Pearson NA19101

```{r}
ensg_pearson_plot %+% d_grouped[d_grouped$individual == "NA19101", ]
ensg_pearson_plot %+% d_grouped[d_grouped$individual == "NA19101" &
                                d_grouped$depth %in% c(50000, 4000000), ]
```

## Endogenous Pearson NA19239

```{r}
ensg_pearson_plot %+% d_grouped[d_grouped$individual == "NA19239", ]
```

## ERCC Pearson NA19098

```{r}
ercc_pearson_plot <- ggplot(d_grouped[d_grouped$individual == "NA19098", ],
                 aes(x = subsampled_cells, y = mean_pearson_ercc,
                     color = as.factor(depth))) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_pearson_ercc - sem_pearson_ercc,
                    ymax = mean_pearson_ercc + sem_pearson_ercc),
                width = 1) +
  facet_grid(gene_subset~type)
ercc_pearson_plot
```

## ERCC Pearson NA19101

```{r}
ercc_pearson_plot %+% d_grouped[d_grouped$individual == "NA19101", ]
ercc_pearson_plot %+% d_grouped[d_grouped$individual == "NA19101" &
                                d_grouped$depth %in% c(50000, 4000000), ]
```

## ERCC Pearson NA19239

```{r}
ercc_pearson_plot %+% d_grouped[d_grouped$individual == "NA19239", ]
```

---

## Endogenous Spearman NA19098

```{r}
ensg_spearman_plot <- ggplot(d_grouped[d_grouped$individual == "NA19098", ],
                 aes(x = subsampled_cells, y = mean_spearman_ensg,
                     color = as.factor(depth))) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_spearman_ensg - sem_spearman_ensg,
                    ymax = mean_spearman_ensg + sem_spearman_ensg),
                width = 1) +
  facet_grid(gene_subset~type)
ensg_spearman_plot
```

## Endogenous Spearman NA19101

```{r}
ensg_spearman_plot %+% d_grouped[d_grouped$individual == "NA19101", ]
ensg_spearman_plot %+% d_grouped[d_grouped$individual == "NA19101" &
                                 d_grouped$depth %in% c(50000, 4000000), ]
```

## Endogenous Spearman NA19239

```{r}
ensg_spearman_plot %+% d_grouped[d_grouped$individual == "NA19239", ]
```

## ERCC Spearman NA19098

```{r}
ercc_spearman_plot <- ggplot(d_grouped[d_grouped$individual == "NA19098", ],
                 aes(x = subsampled_cells, y = mean_spearman_ercc,
                     color = as.factor(depth))) +
  geom_line() +
  geom_errorbar(aes(ymin = mean_spearman_ercc - sem_spearman_ercc,
                    ymax = mean_spearman_ercc + sem_spearman_ercc),
                width = 1) +
  facet_grid(gene_subset~type)
ercc_spearman_plot
```

## ERCC Spearman NA19101

```{r}
ercc_spearman_plot %+% d_grouped[d_grouped$individual == "NA19101", ]
ercc_spearman_plot %+% d_grouped[d_grouped$individual == "NA19101" &
                                 d_grouped$depth %in% c(50000, 4000000), ]
```

## ERCC Spearman NA19239

```{r}
ercc_spearman_plot %+% d_grouped[d_grouped$individual == "NA19239", ]
```

## Available cells

Available cells - don't change up to 1.5 million reads, and then drop off at 4 million reads.

```{r}
ggplot(d_grouped, aes(x = potential_cells, y = available_cells,
                      color = as.factor(depth))) +
  geom_point() +
  facet_wrap(~individual)
```

## Session information

```{r info}
sessionInfo()
```

