---
title: "variance"
author: "PoYuan Tung"
date: 2015-06-15
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

## Input

```{r packages, message=FALSE}
library("dplyr")
library("ggplot2")
theme_set(theme_bw(base_size = 16))
library("edgeR")
```

Summary counts from featureCounts.
Created with [gather-summary-counts.py](https://github.com/jdblischak/singleCellSeq/blob/master/code/gather-summary-counts.py).

```{r input-summary-counts}
summary_counts <- read.table("../data/summary-counts.txt", header = TRUE,
                             stringsAsFactors = FALSE)
summary_per_sample <- summary_counts %>%
  filter(sickle == "quality-trimmed") %>%
  select(-sickle) %>%
  arrange(individual, batch, well, rmdup) %>%
  as.data.frame
```

Input annotation.

```{r input-annotation}
anno <- read.table("../data/annotation.txt", header = TRUE,
                   stringsAsFactors = FALSE)
head(anno)
```

Input read counts.

```{r input-read-counts}
reads <- read.table("../data/reads.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input molecule counts.

```{r input-molecule-counts}
molecules <- read.table("../data/molecules.txt", header = TRUE,
                    stringsAsFactors = FALSE)
```

Input single cell observational quality control data.

```{r input-qc}
qc <- read.table("../data/qc-ipsc.txt", header = TRUE,
                 stringsAsFactors = FALSE)
head(qc)
```

## Remove bad quality cells
### Remove cells with total reads < 2 millons
```{r remove}
#reads per sample
summary_per_sample_reads <- summary_per_sample %>% filter(rmdup == "reads")

#create unmapped ratios
summary_per_sample_reads$unmapped.ratios <- summary_per_sample_reads[,9]/apply(summary_per_sample_reads[,5:13],1,sum)

#create total mapped reads
summary_per_sample_reads$total.mapped <- apply(summary_per_sample_reads[,5:8],1,sum)

#creat ERCC ratios
summary_per_sample_reads$ERCC.ratios <- apply(reads[grep("ERCC", rownames(reads)), ],2,sum)/apply(summary_per_sample_reads[,5:8],1,sum)

#remove bulk keep single cell
summary_per_sample_reads_single <- summary_per_sample_reads[summary_per_sample_reads$well!="bulk",]

#add cell number per well by merging qc file
summary_per_sample_reads_single_qc <- merge(summary_per_sample_reads_single,qc,by=c("individual","batch","well"))

#qc filter
summary_per_sample_reads_single_qc$qc_filter <- summary_per_sample_reads_single_qc$cell_number == 1 & summary_per_sample_reads_single_qc$total.mapped > 2 * 10^6

sum(summary_per_sample_reads_single_qc$qc_filter)
```

## variance of gene expression between individuls

Calculate the variance betwwen individuls from the bulk samples
```{r variance-bulk}
# select bulk samples
reads_bulk<- reads %>% select(contains("bulk"))

# normalization
reads_bulk_cpm <- cpm(reads_bulk)

# create a new dataset
reads_var <- data.frame(reads_bulk_cpm)
sum(reads_var!=reads_bulk_cpm)

# add mean of each individauls
reads_var$mean19098 <- apply(reads_var[,grep("NA19098",names(reads_var))], 1, mean)
reads_var$mean19101 <- apply(reads_var[,grep("NA19101",names(reads_var))], 1, mean)
reads_var$mean19239 <- apply(reads_var[,grep("NA19239",names(reads_var))], 1, mean)

# add variance of bulk means
reads_var$bulk_variance <- apply(reads_var[,c("mean19098","mean19101","mean19239")],1,var)

```

Calculate the variance between individuals using the means from single cells

```{r variance-single-mean}
# select single cell samples
reads_single <- reads %>% select(-contains("bulk"))

# apply qc_filter
reads_single_qc <- reads_single[,summary_per_sample_reads_single_qc$qc_filter]
dim(reads_single_qc)

# normalization
reads_single_cpm <- data.frame(cpm(reads_single_qc))

# remove the ERCC of 19098.2 
  ## identify 19098.2
     sample_name <- names(reads_single_cpm)
     target.column <- sample_name[grep("19098.2",sample_name)]
  ## find out ERCC rows
		 g <- rownames(reads_single_cpm)
		 target.row <- g[grep("ERCC",g)]
	## replace the molecules numbers with NA 
		 reads_single_cpm[target.row,target.column] <- NA
		
# means of single cells within individuals
reads_var$mean.single19098 <- apply(reads_single_cpm[,grep("NA19098",names(reads_single_cpm))], 1, mean, na.rm = TRUE)
reads_var$mean.single19101 <- apply(reads_single_cpm[,grep("NA19101",names(reads_single_cpm))], 1, mean, na.rm = TRUE)
reads_var$mean.single19239 <- apply(reads_single_cpm[,grep("NA19239",names(reads_single_cpm))], 1, mean, na.rm = TRUE)


## variance of means from single cells
reads_var$mean_single_variance <- apply(reads_var[,c("mean.single19098","mean.single19101","mean.single19239")],1,var)

# sellect ERCC
reads_var$ERCC <- grepl("ERCC",rownames(reads_var))

# plot with color-blind-friendly palettes
cbPalette <- c("#999999", "#0000FF", "#990033", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#009E73")

ggplot(reads_var, aes(x = bulk_variance, y = mean_single_variance, col = ERCC)) + geom_point(size = 2, alpha = 0.5) + scale_colour_manual(values=cbPalette) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")
```

Calculate the variance between and within individuals using the single cell data
```{r variance-single}
## variance within individual 
variance.single19098 <- apply(reads_single_cpm[,grep("NA19098",names(reads_single_cpm))], 1, var, na.rm = TRUE)
variance.single19101 <- apply(reads_single_cpm[,grep("NA19101",names(reads_single_cpm))], 1, var, na.rm = TRUE)
variance.single19239 <- apply(reads_single_cpm[,grep("NA19239",names(reads_single_cpm))], 1, var, na.rm = TRUE)

# number of cell
number.of.cell.all   <- sum(grepl("19",sample_name))
number.of.cell.19098 <- sum(grepl("19098",sample_name))
number.of.cell.19101 <- sum(grepl("19101",sample_name))
number.of.cell.19239 <- sum(grepl("19239",sample_name))

# total within individual variance 
reads_var$var_within_individual<-  
 (variance.single19098   *(number.of.cell.19098)  +      
  variance.single19101   *(number.of.cell.19101)  + 
  variance.single19239   *(number.of.cell.19239)  ) / 
 (number.of.cell.all)

ggplot(reads_var, aes(x = bulk_variance, y = var_within_individual, col = ERCC)) + geom_point(size = 2, alpha = 0.5) + scale_colour_manual(values=cbPalette) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

## variance between all single cells
var_all_single <- apply(reads_single_cpm, 1, var, na.rm = TRUE)
reads_var$var_all_single <- var_all_single

## variance between individauls
reads_var$var_between_individual<-( 
  var_all_single         *(number.of.cell.all-1)-
  variance.single19098   *(number.of.cell.19098-1)  -      
  variance.single19101   *(number.of.cell.19101-1)  - 
  variance.single19239   *(number.of.cell.19239-1)  ) / 
 (number.of.cell.all-1)

## the variaance contributed by between individual
reads_var$ratio_var_between_individual <- reads_var[,"var_between_individual"]/var_all_single

ggplot(reads_var, aes(x = var_all_single, y = ratio_var_between_individual, col = ERCC)) + geom_point(size = 2, alpha = 0.5) + scale_colour_manual(values=cbPalette) + scale_x_log10()

ggplot(reads_var, aes(x = var_all_single, y = var_between_individual, col = ERCC)) + geom_point(size = 2, alpha = 0.5) + scale_colour_manual(values=cbPalette) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = bulk_variance, y = var_between_individual, col = ERCC)) + geom_point(size = 2, alpha = 0.5) + scale_colour_manual(values=cbPalette) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

```

## AVONA F-statistics

Pull the p-value
```{r p_of_f}
### create a function for f.test by anova
### compare the to fits:
### 1. lm from all single cells
### 2. lm from each individaul
f.test <- function(data.in){
  tt <- names(data.in)
  individual.id <- rep("19098",length(tt))
  individual.id[grep("19101",tt)] <- "19101"
  individual.id[grep("19239",tt)] <- "19239"
  
  dd <- data.frame(reads=unlist(data.in),individual.id=individual.id)
  fit1 <- lm(reads~1,data=dd)
  fit2 <- lm(reads~1 + individual.id,data=dd)
  anova(fit1,fit2)[2,"Pr(>F)"]
}

# creat the f test table
f.test.table <- do.call(rbind,lapply(rownames(reads_single_cpm),function(x){
  data.frame(gene_name=x,p_of_f=f.test(reads_single_cpm[x,]))  
}))

# sellect ERCC
f.test.table$ERCC <- grepl("ERCC",f.test.table[,1])

# sort 
f.test.table.sort <- f.test.table[order(f.test.table[,2]),]
head(f.test.table.sort)

plot(f.test.table.sort[,2], log = "y",col=as.numeric(f.test.table.sort$ERCC+1))

plot(f.test.table.sort[1:5000,2], log = "y",col=as.numeric(f.test.table.sort$ERCC[1:5000]+1))

plot(f.test.table.sort[1:5000,2],col=as.numeric(f.test.table.sort$ERCC[1:5000]+1),pch=20,cex=.3)

# plot the variance and show p 0f f
reads_var$p_of_f <- f.test.table$p_of_f

ggplot(reads_var, aes(x = bulk_variance, y = mean_single_variance, shape = as.factor(ERCC), col = p_of_f)) + geom_point(size = 2, alpha = 0.25) + scale_x_log10() + scale_y_log10() + scale_colour_gradient2(midpoint= 0.5, space="Lab") + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = bulk_variance, y = var_within_individual, shape = as.factor(ERCC), col = p_of_f)) + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_y_log10() + scale_colour_gradient2(midpoint= 0.5, space="Lab") + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = bulk_variance, y = var_between_individual, shape = as.factor(ERCC), col = p_of_f)) + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_y_log10() + scale_colour_gradient2(midpoint= 0.5, space="Lab") + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = var_all_single, y = var_between_individual, shape = as.factor(ERCC), col = p_of_f)) + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_y_log10() + scale_colour_gradient2(midpoint= 0.5, space="Lab") + stat_function(fun= function(x) {x}, col= "#56B4E9")
```

Pull the F value
```{r Fvalue}
# looking at the F value
f.test.F <- function(data.in){
  tt <- names(data.in)
  individual.id <- rep("19098",length(tt))
  individual.id[grep("19101",tt)] <- "19101"
  individual.id[grep("19239",tt)] <- "19239"
  
  dd <- data.frame(reads=unlist(data.in),individual.id=individual.id)
  fit1 <- lm(reads~1,data=dd)
  fit2 <- lm(reads~1 + individual.id,data=dd)
  anova(fit1,fit2)[2,"F"]
}
# creat the f test table of F value
f.test.F.table <- do.call(rbind,lapply(rownames(reads_single_cpm),function(x){
  data.frame(gene_name=x,F_value=f.test.F(reads_single_cpm[x,]))  
}))

# sellect ERCC
f.test.F.table$ERCC <- grepl("ERCC",f.test.F.table[,1])

# plot the variance and show F value
reads_var$F_value <- f.test.F.table$F_value

# calculate F value from p
# when p=0.01
qf(1-0.05,2,629)

# create color index
reads_var$F_color <- "1 < F < 3"
reads_var$F_color[reads_var$F_value >=3] <- "F >= 3"
reads_var$F_color[reads_var$F_value <=1] <- "F <= 1"

# use the F value from p to scale the gradient
ggplot(reads_var, aes(x = bulk_variance, y = mean_single_variance, shape = as.factor(ERCC), col = F_color)) + scale_colour_manual(values=cbPalette) + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = bulk_variance, y = var_within_individual, shape = as.factor(ERCC), col = F_color)) + scale_colour_manual(values=cbPalette) + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = bulk_variance, y = var_between_individual, shape = as.factor(ERCC), col = F_color)) + scale_colour_manual(values=cbPalette) + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = var_all_single, y = var_within_individual, shape = as.factor(ERCC), col = F_color)) + scale_colour_manual(values=cbPalette) + geom_point(size = 2, alpha = 0.5) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

ggplot(reads_var, aes(x = var_all_single, y = var_between_individual, shape = as.factor(ERCC), col = F_color)) + scale_colour_manual(values=cbPalette) + geom_point(size = 2, alpha = 0.2) + scale_x_log10() + scale_y_log10() + stat_function(fun= function(x) {x}, col= "#56B4E9")

```

## Session information

```{r info}
sessionInfo()
```
